### 2021-07-23 manuscript/Testing the cognitive foundations of Paleolithic social transmission.Rmd
add the relevant code and data

---
title: Testing the cognitive foundations of Paleolithic social transmission
subtitle: 
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.

authors: 
- name: Justin Pargeter 1, 2
  address: 
  - 1. Department of Anthropology, New York University, New York, NY, USA; 2. Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa
  email: justin.pargeter@nyu.edu

- name: Megan Beney Kilgore 3
  address: 3. Department of Anthropology, Emory University, Atlanta, GA, USA
  email: megan.elizabeth.beney@emory.edu

- name: Cheng Liu 3
  address: 3. Department of Anthropology, Emory University, Atlanta, GA, USA
  email: raylc1996@outlook.com

- name: Dietrich Stout 3
  address: 3. Department of Anthropology, Emory University, Atlanta, GA, USA
  email: dwstout@emory.edu

keywords:
- key
- dictionary
- word

#PACS: 
#- PAC1
#- superPAC

abstract: |
  Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret. Toolmaking skill-learning, in particular, has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability. Despite decades of research on stone toolmaking we still know little about the cognitive prerequisites of learning under different social transmission conditions that may have prevailed during the Paleolithic. This paper presents results from a pilot experimental study to trial new experimental methods for investigating the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition. We trained 32 participants for 2 hours to make simple stone tools under two different instructional conditions (observation-only vs. direct-active teaching) employing appropriate raw materials and in-person interaction. Participant performance was evaluated through analysis of the stone artifacts produced and was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study. Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with tendency to use social information in an observation-only group. These results have implications for debates surrounding the cumulative nature of human culture and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.

bibliography: bibliography.bib
bibstyle: spphys
# bibstyle options spbasic(default), spphys, spmpsci
output: rticles::springer_article
---

# Introduction {#intro}

Your text comes here. Separate text sections with \cite{Mislevy06Cog}.

# Section title {#sec:1}

Text with citations by \cite{Galyardt14mmm}.

## Subsection title {#sec:2}

as required. Don't forget to give each section
and subsection a unique label (see Sect. \ref{sec:1}).

#### Paragraph headings

Use paragraph headings as needed.

\begin{align}
a^2+b^2=c^2
\end{align}

### 2021-07-23 manuscript/Testing-the-cognitive-foundations-of-Paleolithic-social-transmission.tex
add the relevant code and data

% !TeX program = pdfLaTeX
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}

\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

%% load any required packages here

% Pandoc citation processing

\begin{document}

\title{Testing the cognitive foundations of Paleolithic social
transmission \thanks{Grants or other notes about the article that should
go on the front page should be placed here. General acknowledgments
should be placed at the end of the article.} }

\author{  Justin Pargeter 1, 2 \and  Megan Beney Kilgore 3 \and  Cheng
Liu 3 \and  Dietrich Stout 3 \and  }

\institute{
        Justin Pargeter 1, 2 \at
     1. Department of Anthropology, New York University, New York, NY,
USA; 2. Palaeo-Research Institute, University of Johannesburg, Auckland
Park, South Africa \\
     \email{\href{mailto:justin.pargeter@nyu.edu}{\nolinkurl{justin.pargeter@nyu.edu}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Megan Beney Kilgore 3 \at
     3. Department of Anthropology, Emory University, Atlanta, GA,
USA \\
     \email{\href{mailto:megan.elizabeth.beney@emory.edu}{\nolinkurl{megan.elizabeth.beney@emory.edu}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Cheng Liu 3 \at
     3. Department of Anthropology, Emory University, Atlanta, GA,
USA \\
     \email{\href{mailto:raylc1996@outlook.com}{\nolinkurl{raylc1996@outlook.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Dietrich Stout 3 \at
     3. Department of Anthropology, Emory University, Atlanta, GA,
USA \\
     \email{\href{mailto:dwstout@emory.edu}{\nolinkurl{dwstout@emory.edu}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
Stone tools provide key evidence of human cognitive evolution but remain
difficult to interpret. Toolmaking skill-learning, in particular, has
been understudied even though: 1) the most salient cognitive demands of
toolmaking should occur during learning, and 2) variation in learning
aptitude would have provided the raw material for any past selection
acting on tool making ability. Despite decades of research on stone
toolmaking we still know little about the cognitive prerequisites of
learning under different social transmission conditions that may have
prevailed during the Paleolithic. This paper presents results from a
pilot experimental study to trial new experimental methods for
investigating the effect of learning conditions and individual
differences on Oldowan flake-tool making skill acquisition. We trained
32 participants for 2 hours to make simple stone tools under two
different instructional conditions (observation-only vs.~direct-active
teaching) employing appropriate raw materials and in-person interaction.
Participant performance was evaluated through analysis of the stone
artifacts produced and was compared both across experimental groups and
with respect to individual participant differences in grip strength,
motor accuracy, and cognitive function measured for the study. Our
results show aptitude to be associated with fluid intelligence in a
verbally instructed group and with tendency to use social information in
an observation-only group. These results have implications for debates
surrounding the cumulative nature of human culture and the role of
evolved psychological mechanisms in ``high fidelity'' transmission of
information, particularly through imitation and teaching.
\\
\keywords{
        key \and
        dictionary \and
        word \and
    }

\end{abstract}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

Your text comes here. Separate text sections with \cite{Mislevy06Cog}.

\hypertarget{sec:1}{%
\section{Section title}\label{sec:1}}

Text with citations by \cite{Galyardt14mmm}.

\hypertarget{sec:2}{%
\subsection{Subsection title}\label{sec:2}}

as required. Don't forget to give each section and subsection a unique
label (see Sect. \ref{sec:1}).

\hypertarget{paragraph-headings}{%
\paragraph{Paragraph headings}\label{paragraph-headings}}

Use paragraph headings as needed.

\begin{align}
a^2+b^2=c^2
\end{align}

\bibliographystyle{spphys}
\bibliography{bibliography.bib}

\end{document}

### 2021-07-23 code/PaST_r_code.R

@@ -1,0 +1,973 @@
+##Libraries####
+detachAllPackages <- function() {
+      
+      basic.packages <- c("package:stats","package:graphics","package:grDevices","package:utils","package:datasets","package:methods","package:base")
+      
+      package.list <- search()[ifelse(unlist(gregexpr("package:",search()))==1,TRUE,FALSE)]
+      
+      package.list <- setdiff(package.list,basic.packages)
+      
+      if (length(package.list)>0)  for (package in package.list) detach(package, character.only=TRUE)
+      
+}
+
+detachAllPackages()
+library(car)
+library(ggplot2)
+library(reshape2)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(DescTools)
+library(stringr)
+library("factoextra")
+library("psych") 
+library(Hmisc)
+library(glmulti)
+library(report)
+library(ggeffects) 
+library(sjPlot)
+library(rcompanion)
+library(caret)
+library(permuco)
+
+#### Import datasets #### ####
+
+experiment_data<-read.csv("/Users/JustinPargeter/Google Drive/Projects/PaST/Pilot experiment/Lithics/Experiment database/past_stats/past_stats/past_cores_result.csv") %>%
+      rename(subject_id = Subject,
+             core_number = nodule_number) %>%
+      mutate(subject_id=as.factor(subject_id),
+             core_number=as.factor(core_number),
+             Condition=as.factor(Condition))
+
+experiment_core_usedata<-read.csv("/Users/JustinPargeter/Google Drive/Projects/PaST/Pilot experiment/Lithics/Experiment database/past_stats/past_stats/core_use_records.csv") %>%
+      mutate(subject=as.factor(subject),
+             core_number=as.factor(core_number))
+
+lithic_data<-read.csv("past_pilot_lithics.csv") %>%
+      rename(flake_mass = flake_mass,
+             flake_max_thickness = max_thickness,
+             core_final_mass=core_mass) %>%
+      mutate(core_number_nyu=core_number) %>%
+      rename(subject_id_nyu=subject_id) 
+
+extra_thickness_flakes<-read.csv("extra_flakes.csv") 
+  
+pre_exp_cores_data<-read.csv("/Users/JustinPargeter/Google Drive/Projects/PaST/Pilot experiment/Lithics/Experiment database/past_stats/past_stats/pre_exp_cores_data.csv") %>%
+      rename(core_number = nodule_number) %>%
+      mutate(core_number=as.factor(core_number))
+
[…]

### 2021-07-23 manuscript/Testing the cognitive foundations of Paleolithic social transmission.Rmd
Add the first paragraph and the data dictionary

@@
 ---
-title: Testing the cognitive foundations of Paleolithic social transmission
+title: Testing the motor and cognitive foundations of Paleolithic social transmission
 subtitle: 
@@
 keywords:
-- key
-- dictionary
-- word
+- Oldowan
+- Stone toolmaking
+- Social learning
+- Individual variation
+- Cognitive aptitudes
+- Motor skills
 
@@
   Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret.
-Toolmaking skill-learning, in particular, has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability.
-Despite decades of research on stone toolmaking we still know little about the cognitive prerequisites of learning under different social transmission conditions that may have prevailed during the Paleolithic.
-This paper presents results from a pilot experimental study to trial new experimental methods for investigating the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition.
-We trained 32 participants for 2 hours to make simple stone tools under two different instructional conditions (observation-only vs.
-direct-active teaching) employing appropriate raw materials and in-person interaction.
-Participant performance was evaluated through analysis of the stone artifacts produced and was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study.
-Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with tendency to use social information in an observation-only group.
-These results have implications for debates surrounding the cumulative nature of human culture and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.
+Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability.
+However, we actually know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic.
+This paper presents results from a pilot experimental study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition.
+We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs.
+direct active teaching) employing appropriate raw materials, practice time, and real human interaction.
+Participant performance was evaluated through analysis of the stone artifacts produced.
+Performance was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study.
+Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group.
+These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.
 
 bibliography: bibliography.bib
-bibstyle: spphys
+bibstyle: spbasic
 # bibstyle options spbasic(default), spphys, spmpsci
@@
 
-Your text comes here. Separate text sections with \cite{Mislevy06Cog}.
+Stone tools have long been seen as a key source of evidence for understanding human behavioral and cognitive evolution [@darwin1871; @oakley1949; @washburn1960].
+Pathbreaking attempts to infer specific cognitive capacities from this evidence largely focused on the basic requirements of tool production [@isaac1976; @wynn1979; @gowlett1984; @wynn2004].
+More recently, increasing attention has been directed to the processes and demands of stone tool making skill acquisition [@roux1995; @stout2002; @stout2005; @geribàs2010; @nonaka2010; @stout2011; @putt2014; @hecht2015; @duke2015; @morgan2015; @stout2015; @lombao2017; @putt2017; @cataldo2018; @putt2019; @pargeter2019; @pargeter2020].
+This is motivated by the expectation that the most salient cognitive demands of tool making should occur during learning rather than routine expert performance [@stout2015] and by interest in the relevance of different social learning mechanisms such as imitation [@rein2014; @stout2019], emulation [@tehrani2008; @wilkins2018], and language [@ohnuma1997; @putt2014; @morgan2015; @lombao2017; @putt2017; @cataldo2018] to the reproduction of Paleolithic technologies.
 
-# Section title {#sec:1}
+## **Individual Differences**
 
-Text with citations by \cite{Galyardt14mmm}.
+## **Teaching, Language, and Tool Making**
 
-## Subsection title {#sec:2}
+## **Raw materials and knapping skill**
 
-as required. Don't forget to give each section
-and subsection a unique label (see Sect. \ref{sec:1}).
+# **Materials and Methods**
 
-#### Paragraph headings
+## **Participants**
+## **Study Visit**
 
-Use paragraph headings as needed.
+## **Individual Difference Measures**
 
-\begin{align}
-a^2+b^2=c^2
-\end{align}
+##  **Stone Tool Making**
+
+### **Raw Materials**
+### **Experimental Conditions**
+##  **Lithic Analysis**
+##  **Statistical Analyses**
+
+# **Results** 
+## **Principal Component analyses**
+
+### **Flake size and shape**
+### **Lithic flaking performance measures**
+
+## **Do trained, untrained, and expert knappers perform differently?**
+
+## **Does training/practice time impact flaking performance?**
+
+## **Do individual differences in motor skill and psychometric measures predict flaking performance? **
+### **Model 1: Individual differences and quantity flaking**
+### **Model 2: Individual differences and quality flaking**
+
+# **Discussion**
+
+# **Conclusions**
+
+# **Acknowledgments**
+
+# **References**

### 2021-07-23 manuscript/Testing-the-cognitive-foundations-of-Paleolithic-social-transmission.tex
Add the first paragraph and the data dictionary

@@
 % Pandoc citation processing
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry sp
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set line spacing
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#3\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc} % for \widthof, \maxof
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\maxof{\widthof{#1}}{\csllabelwidth}}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth}{#1}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
-\title{Testing the cognitive foundations of Paleolithic social
+\title{Testing the motor and cognitive foundations of Paleolithic social
 transmission \thanks{Grants or other notes about the article that should
@@
 Stone tools provide key evidence of human cognitive evolution but remain
-difficult to interpret. Toolmaking skill-learning, in particular, has
-been understudied even though: 1) the most salient cognitive demands of
+difficult to interpret. Toolmaking skill-learning in particular has been
+understudied even though: 1) the most salient cognitive demands of
 toolmaking should occur during learning, and 2) variation in learning
 aptitude would have provided the raw material for any past selection
-acting on tool making ability. Despite decades of research on stone
-toolmaking we still know little about the cognitive prerequisites of
-learning under different social transmission conditions that may have
-prevailed during the Paleolithic. This paper presents results from a
-pilot experimental study to trial new experimental methods for
-investigating the effect of learning conditions and individual
-differences on Oldowan flake-tool making skill acquisition. We trained
-32 participants for 2 hours to make simple stone tools under two
-different instructional conditions (observation-only vs.~direct-active
-teaching) employing appropriate raw materials and in-person interaction.
-Participant performance was evaluated through analysis of the stone
-artifacts produced and was compared both across experimental groups and
-with respect to individual participant differences in grip strength,
-motor accuracy, and cognitive function measured for the study. Our
-results show aptitude to be associated with fluid intelligence in a
-verbally instructed group and with tendency to use social information in
-an observation-only group. These results have implications for debates
-surrounding the cumulative nature of human culture and the role of
-evolved psychological mechanisms in ``high fidelity'' transmission of
-information, particularly through imitation and teaching.
+acting on tool making ability. However, we actually know very little
+about the cognitive prerequisites of learning under different
+information transmission conditions that may have prevailed during the
+Paleolithic. This paper presents results from a pilot experimental study
+to trial new experimental methods for studying the effect of learning
+conditions and individual differences on Oldowan flake-tool making skill
+acquisition. We trained 23 participants for 2 hours to make stone flakes
+under two different instructional conditions (observation only
+vs.~direct active teaching) employing appropriate raw materials,
+practice time, and real human interaction. Participant performance was
+evaluated through analysis of the stone artifacts produced. Performance
+was compared both across experimental groups and with respect to
+individual participant differences in grip strength, motor accuracy, and
+cognitive function measured for the study. Our results show aptitude to
+be associated with fluid intelligence in a verbally instructed group and
+with a tendency to use social information in an observation-only group.
+These results have implications for debates surrounding the cumulative
+nature of human culture, the relative contributions of knowledge and
+know-how for stone tool making, and the role of evolved psychological
+mechanisms in ``high fidelity'' transmission of information,
+particularly through imitation and teaching.
 \\
 \keywords{
-        key \and
-        dictionary \and
-        word \and
+        Oldowan \and
+        Stone toolmaking \and
+        Social learning \and
+        Individual variation \and
+        Cognitive aptitudes \and
+        Motor skills \and
     }
@@
 
-Your text comes here. Separate text sections with \cite{Mislevy06Cog}.
+Stone tools have long been seen as a key source of evidence for
+understanding human behavioral and cognitive evolution (Darwin 1871;
+Oakley 1949; Washburn 1960). Pathbreaking attempts to infer specific
+cognitive capacities from this evidence largely focused on the basic
+requirements of tool production (Isaac 1976; Wynn 1979; Gowlett 1984;
+Wynn and Coolidge 2004). More recently, increasing attention has been
+directed to the processes and demands of stone tool making skill
+acquisition (Roux, Bril, and Dietrich 1995; Stout 2002; Stout et al.
+2005; Geribàs, Mosquera, and Vergès 2010; Nonaka, Bril, and Rein 2010;
+Stout et al. 2011; Putt, Woods, and Franciscus 2014; Hecht et al. 2015;
+Duke and Pargeter 2015; Morgan et al. 2015; Stout and Khreisheh 2015;
+Lombao, Guardiola, and Mosquera 2017; Putt et al. 2017; Cataldo,
+Migliano, and Vinicius 2018; Putt, Wijeakumar, and Spencer 2019;
+Pargeter and Shea 2019; Pargeter et al. 2020). This is motivated by the
+expectation that the most salient cognitive demands of tool making
+should occur during learning rather than routine expert performance
+(Stout and Khreisheh 2015) and by interest in the relevance of different
+social learning mechanisms such as imitation (Rein, Nonaka, and Bril
+2014; Stout et al. 2019), emulation (Tehrani and Riede 2008; Wilkins
+2018), and language (Ohnuma, Aoki, and Akazawa 1997; Putt, Woods, and
+Franciscus 2014; Morgan et al. 2015; Lombao, Guardiola, and Mosquera
+2017; Putt et al. 2017; Cataldo, Migliano, and Vinicius 2018) to the
+reproduction of Paleolithic technologies.
 
-\hypertarget{sec:1}{%
-\section{Section title}\label{sec:1}}
+\hypertarget{individual-differences}{%
+\subsection{\texorpdfstring{\textbf{Individual
+Differences}}{Individual Differences}}\label{individual-differences}}
 
-Text with citations by \cite{Galyardt14mmm}.
+\hypertarget{teaching-language-and-tool-making}{%
+\subsection{\texorpdfstring{\textbf{Teaching, Language, and Tool
+Making}}{Teaching, Language, and Tool Making}}\label{teaching-language-and-tool-making}}
 
-\hypertarget{sec:2}{%
-\subsection{Subsection title}\label{sec:2}}
+\hypertarget{raw-materials-and-knapping-skill}{%
+\subsection{\texorpdfstring{\textbf{Raw materials and knapping
+skill}}{Raw materials and knapping skill}}\label{raw-materials-and-knapping-skill}}
 
-as required. Don't forget to give each section and subsection a unique
-label (see Sect. \ref{sec:1}).
+\hypertarget{materials-and-methods}{%
+\section{\texorpdfstring{\textbf{Materials and
+Methods}}{Materials and Methods}}\label{materials-and-methods}}
 
-\hypertarget{paragraph-headings}{%
-\paragraph{Paragraph headings}\label{paragraph-headings}}
+\hypertarget{participants}{%
+\subsection{\texorpdfstring{\textbf{Participants}}{Participants}}\label{participants}}
 
-Use paragraph headings as needed.
+\hypertarget{study-visit}{%
+\subsection{\texorpdfstring{\textbf{Study
+Visit}}{Study Visit}}\label{study-visit}}
 
-\begin{align}
-a^2+b^2=c^2
-\end{align}
+\hypertarget{individual-difference-measures}{%
+\subsection{\texorpdfstring{\textbf{Individual Difference
+Measures}}{Individual Difference Measures}}\label{individual-difference-measures}}
 
-\bibliographystyle{spphys}
+\hypertarget{stone-tool-making}{%
+\subsection{\texorpdfstring{\textbf{Stone Tool
+Making}}{Stone Tool Making}}\label{stone-tool-making}}
+
+\hypertarget{raw-materials}{%
+\subsubsection{\texorpdfstring{\textbf{Raw
+Materials}}{Raw Materials}}\label{raw-materials}}
+
+\hypertarget{experimental-conditions}{%
+\subsubsection{\texorpdfstring{\textbf{Experimental
+Conditions}}{Experimental Conditions}}\label{experimental-conditions}}
+
+\hypertarget{lithic-analysis}{%
+\subsection{\texorpdfstring{\textbf{Lithic
+Analysis}}{Lithic Analysis}}\label{lithic-analysis}}
+
+\hypertarget{statistical-analyses}{%
+\subsection{\texorpdfstring{\textbf{Statistical
+Analyses}}{Statistical Analyses}}\label{statistical-analyses}}
+
+\hypertarget{results}{%
+\section{\texorpdfstring{\textbf{Results}}{Results}}\label{results}}
+
+\hypertarget{principal-component-analyses}{%
+\subsection{\texorpdfstring{\textbf{Principal Component
+analyses}}{Principal Component analyses}}\label{principal-component-analyses}}
+
+\hypertarget{flake-size-and-shape}{%
+\subsubsection{\texorpdfstring{\textbf{Flake size and
+shape}}{Flake size and shape}}\label{flake-size-and-shape}}
+
+\hypertarget{lithic-flaking-performance-measures}{%
+\subsubsection{\texorpdfstring{\textbf{Lithic flaking performance
+measures}}{Lithic flaking performance measures}}\label{lithic-flaking-performance-measures}}
+
+\hypertarget{do-trained-untrained-and-expert-knappers-perform-differently}{%
+\subsection{\texorpdfstring{\textbf{Do trained, untrained, and expert
+knappers perform
+differently?}}{Do trained, untrained, and expert knappers perform differently?}}\label{do-trained-untrained-and-expert-knappers-perform-differently}}
+
+\hypertarget{does-trainingpractice-time-impact-flaking-performance}{%
+\subsection{\texorpdfstring{\textbf{Does training/practice time impact
+flaking
+performance?}}{Does training/practice time impact flaking performance?}}\label{does-trainingpractice-time-impact-flaking-performance}}
+
+\hypertarget{do-individual-differences-in-motor-skill-and-psychometric-measures-predict-flaking-performance}{%
+\subsection{\texorpdfstring{\textbf{Do individual differences in motor
+skill and psychometric measures predict flaking performance?
+}}{Do individual differences in motor skill and psychometric measures predict flaking performance?
+}}\label{do-individual-differences-in-motor-skill-and-psychometric-measures-predict-flaking-performance}}
+
+\hypertarget{model-1-individual-differences-and-quantity-flaking}{%
+\subsubsection{\texorpdfstring{\textbf{Model 1: Individual differences
+and quantity
+flaking}}{Model 1: Individual differences and quantity flaking}}\label{model-1-individual-differences-and-quantity-flaking}}
+
+\hypertarget{model-2-individual-differences-and-quality-flaking}{%
+\subsubsection{\texorpdfstring{\textbf{Model 2: Individual differences
+and quality
+flaking}}{Model 2: Individual differences and quality flaking}}\label{model-2-individual-differences-and-quality-flaking}}
+
+\hypertarget{discussion}{%
+\section{\texorpdfstring{\textbf{Discussion}}{Discussion}}\label{discussion}}
+
+\hypertarget{conclusions}{%
+\section{\texorpdfstring{\textbf{Conclusions}}{Conclusions}}\label{conclusions}}
+
+\hypertarget{acknowledgments}{%
+\section{\texorpdfstring{\textbf{Acknowledgments}}{Acknowledgments}}\label{acknowledgments}}
+
+\hypertarget{references}{%
+\section*{\texorpdfstring{\textbf{References}}{References}}\label{references}}
+\addcontentsline{toc}{section}{\textbf{References}}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-cataldo2018}{}%
+Cataldo, Dana Michelle, Andrea Bamberg Migliano, and Lucio Vinicius.
+2018. {``Speech, Stone Tool-Making and the Evolution of Language.''}
+\emph{PLOS ONE} 13 (1): e0191071.
+\url{https://doi.org/10.1371/journal.pone.0191071}.
+
+\leavevmode\hypertarget{ref-darwin1871}{}%
+Darwin, Charles. 1871. \emph{The Descent of Man, and Selection in
+Relation to Sex}. 1st ed. London: John Murray.
+
+\leavevmode\hypertarget{ref-duke2015}{}%
+Duke, Hilary, and Justin Pargeter. 2015. {``Weaving Simple Solutions to
+Complex Problems: An Experimental Study of Skill in Bipolar
+Cobble-Splitting.''} \emph{Lithic Technology} 40 (4): 349--65.
+\url{https://doi.org/10.1179/2051618515Y.0000000016}.
+
+\leavevmode\hypertarget{ref-geribuxe0s2010}{}%
+Geribàs, Núria, Marina Mosquera, and Josep Maria Vergès. 2010. {``What
+Novice Knappers Have to Learn to Become Expert Stone Toolmakers.''}
+\emph{Journal of Archaeological Science} 37 (11): 2857--70.
+\url{https://doi.org/10.1016/j.jas.2010.06.026}.
+
+\leavevmode\hypertarget{ref-gowlett1984}{}%
+Gowlett, John A. J. 1984. {``Mental Abilities of Early Man: A Look at
+Some Hard Evidence.''} \emph{Higher Education Quarterly} 38 (3):
+199--220. \url{https://doi.org/10.1111/j.1468-2273.1984.tb01387.x}.
+
+\leavevmode\hypertarget{ref-hecht2015}{}%
+Hecht, E. E., D. A. Gutman, N. Khreisheh, S. V. Taylor, J. Kilner, A. A.
+Faisal, B. A. Bradley, T. Chaminade, and D. Stout. 2015. {``Acquisition
+of Paleolithic Toolmaking Abilities Involves Structural Remodeling to
+Inferior Frontoparietal Regions.''} \emph{Brain Structure and Function}
+220 (4): 2315--31. \url{https://doi.org/10.1007/s00429-014-0789-6}.
+
+\leavevmode\hypertarget{ref-isaac1976}{}%
+Isaac, Glynn L. 1976. {``Stages of Cultural Elaboration in the
+Pleistocene: Possible Archaeological Indicators of the Development of
+Language Capabilities.''} \emph{Annals of the New York Academy of
+Sciences} 280 (1): 275--88.
+\url{https://doi.org/10.1111/j.1749-6632.1976.tb25494.x}.
+
+\leavevmode\hypertarget{ref-lombao2017}{}%
+Lombao, D., M. Guardiola, and M. Mosquera. 2017. {``Teaching to Make
+Stone Tools: New Experimental Evidence Supporting a Technological
+Hypothesis for the Origins of Language.''} \emph{Scientific Reports} 7
+(1): 1--14. \url{https://doi.org/10.1038/s41598-017-14322-y}.
+
+\leavevmode\hypertarget{ref-morgan2015}{}%
+Morgan, T. J. H., N. T. Uomini, L. E. Rendell, L. Chouinard-Thuly, S. E.
+Street, H. M. Lewis, C. P. Cross, et al. 2015. {``Experimental Evidence
+for the Co-Evolution of Hominin Tool-Making Teaching and Language.''}
+\emph{Nature Communications} 6 (1): 6029.
+\url{https://doi.org/10.1038/ncomms7029}.
+
+\leavevmode\hypertarget{ref-nonaka2010}{}%
+Nonaka, Tetsushi, Blandine Bril, and Robert Rein. 2010. {``How Do Stone
+Knappers Predict and Control the Outcome of Flaking? Implications for
+Understanding Early Stone Tool Technology.''} \emph{Journal of Human
+Evolution} 59 (2): 155--67.
+\url{https://doi.org/10.1016/j.jhevol.2010.04.006}.
+
+\leavevmode\hypertarget{ref-oakley1949}{}%
+Oakley, Kenneth P. 1949. \emph{Man the Toolmaker}. London: Trustees of
+the British Museum.
+
+\leavevmode\hypertarget{ref-ohnuma1997}{}%
+Ohnuma, Katsuhiko, Kenichi Aoki, and And Takeru Akazawa. 1997.
+{``Transmission of Tool-Making Through Verbal and Non-Verbal
+Commu-Nication: Preliminary Experiments in Levallois Flake
+Production.''} \emph{Anthropological Science} 105 (3): 159--68.
+\url{https://doi.org/10.1537/ase.105.159}.
+
+\leavevmode\hypertarget{ref-pargeter2020}{}%
+Pargeter, Justin, Nada Khreisheh, John J. Shea, and Dietrich Stout.
+2020. {``Knowledge Vs. Know-How? Dissecting the Foundations of Stone
+Knapping Skill.''} \emph{Journal of Human Evolution} 145 (August):
+102807. \url{https://doi.org/10.1016/j.jhevol.2020.102807}.
+
+\leavevmode\hypertarget{ref-pargeter2019}{}%
+Pargeter, Justin, and John J. Shea. 2019. {``Going Big Versus Going
+Small: Lithic Miniaturization in Hominin Lithic Technology.''}
+\emph{Evolutionary Anthropology: Issues, News, and Reviews} 28 (2):
+72--85. \url{https://doi.org/10.1002/evan.21775}.
+
+\leavevmode\hypertarget{ref-putt2017}{}%
+Putt, Shelby S., Sobanawartiny Wijeakumar, Robert G. Franciscus, and
+John P. Spencer. 2017. {``The Functional Brain Networks That Underlie
+Early Stone Age Tool Manufacture.''} \emph{Nature Human Behaviour} 1
+(6): 1--8. \url{https://doi.org/10.1038/s41562-017-0102}.
+
+\leavevmode\hypertarget{ref-putt2019}{}%
+Putt, Shelby S., Sobanawartiny Wijeakumar, and John P. Spencer. 2019.
+{``Prefrontal Cortex Activation Supports the Emergence of Early Stone
+Age Toolmaking Skill.''} \emph{NeuroImage} 199 (October): 57--69.
+\url{https://doi.org/10.1016/j.neuroimage.2019.05.056}.
+
+\leavevmode\hypertarget{ref-putt2014}{}%
+Putt, Shelby S., Alexander D. Woods, and Robert G. Franciscus. 2014.
+{``The Role of Verbal Interaction During Experimental Bifacial Stone
+Tool Manufacture.''} \emph{Lithic Technology} 39 (2): 96--112.
+\url{https://doi.org/10.1179/0197726114Z.00000000036}.
+
+\leavevmode\hypertarget{ref-rein2014}{}%
+Rein, Robert, Tetsushi Nonaka, and Blandine Bril. 2014. {``Movement
+Pattern Variability in Stone Knapping: Implications for the Development
+of Percussive Traditions.''} \emph{PLOS ONE} 9 (11): e113567.
+\url{https://doi.org/10.1371/journal.pone.0113567}.
+
+\leavevmode\hypertarget{ref-roux1995}{}%
+Roux, Valentine, Blandine Bril, and Gilles Dietrich. 1995. {``Skills and
+Learning Difficulties Involved in Stone Knapping: The Case of
+Stone{-}Bead Knapping in Khambhat, India.''} \emph{World Archaeology} 27
+(1): 63--87. \url{https://doi.org/10.1080/00438243.1995.9980293}.
+
+\leavevmode\hypertarget{ref-stout2002}{}%
+Stout, Dietrich. 2002. {``Skill and Cognition in Stone Tool Production:
+An Ethnographic Case Study from Irian Jaya.''} \emph{Current
+Anthropology} 43 (5): 693--722. \url{https://doi.org/10.1086/342638}.
+
+\leavevmode\hypertarget{ref-stout2015}{}%
+Stout, Dietrich, and Nada Khreisheh. 2015. {``Skill Learning and Human
+Brain Evolution: An Experimental Approach.''} \emph{Cambridge
+Archaeological Journal} 25 (4): 867--75.
+\url{https://doi.org/10.1017/S0959774315000359}.
+
+\leavevmode\hypertarget{ref-stout2011}{}%
+Stout, Dietrich, Richard Passingham, Christopher Frith, Jan Apel, and
+Thierry Chaminade. 2011. {``Technology, expertise and social cognition
+in human evolution.''} \emph{The European Journal of Neuroscience} 33
+(7): 1328--38. \url{https://doi.org/10.1111/j.1460-9568.2011.07619.x}.
+
+\leavevmode\hypertarget{ref-stout2005}{}%
+Stout, Dietrich, Jay Quade, Sileshi Semaw, Michael J. Rogers, and Naomi
+E. Levin. 2005. {``Raw Material Selectivity of the Earliest Stone
+Toolmakers at Gona, Afar, Ethiopia.''} \emph{Journal of Human Evolution}
+48 (4): 365--80. \url{https://doi.org/10.1016/j.jhevol.2004.10.006}.
+
+\leavevmode\hypertarget{ref-stout2019}{}%
+Stout, Dietrich, Michael J. Rogers, Adrian V. Jaeggi, and Sileshi Semaw.
+2019. {``Archaeology and the Origins of Human Cumulative Culture: A Case
+Study from the Earliest Oldowan at Gona, Ethiopia.''} \emph{Current
+Anthropology} 60 (3): 309--40. \url{https://doi.org/10.1086/703173}.
+
+\leavevmode\hypertarget{ref-tehrani2008}{}%
+Tehrani, Jamshid J., and Felix Riede. 2008. {``Towards an Archaeology of
+Pedagogy: Learning, Teaching and the Generation of Material Culture
+Traditions.''} \emph{World Archaeology} 40 (3): 316--31.
+\url{https://doi.org/10.1080/00438240802261267}.
+
+\leavevmode\hypertarget{ref-washburn1960}{}%
+Washburn, Sherwood L. 1960. {``Tools and Human Evolution.''}
+\emph{Scientific American} 203 (3): 62--75.
+\url{https://doi.org/10.1038/scientificamerican0960-62}.
+
+\leavevmode\hypertarget{ref-wilkins2018}{}%
+Wilkins, Jayne. 2018. {``The Point Is the Point: Emulative Social
+Learning and Weapon Manufacture in the Middle Stone Age of South
+Africa.''} In, edited by Michael J. O'Brien, Briggs Buchanan, and Metin
+I. Eren, 153--74. Cambridge, MA: The MIT Press.
+
+\leavevmode\hypertarget{ref-wynn1979}{}%
+Wynn, Thomas. 1979. {``The Intelligence of Later Acheulean Hominids.''}
+\emph{Man} 14 (3): 371--91. \url{https://doi.org/10.2307/2801865}.
+
+\leavevmode\hypertarget{ref-wynn2004}{}%
+Wynn, Thomas, and Frederick L. Coolidge. 2004. {``The expert Neandertal
+mind.''} \emph{Journal of Human Evolution} 46 (4): 467--87.
+\url{https://doi.org/10.1016/j.jhevol.2004.01.005}.
+
+\end{CSLReferences}
+
+\bibliographystyle{spbasic}
 \bibliography{bibliography.bib}

### 2021-07-29 manuscript/Testing the cognitive foundations of Paleolithic social transmission.Rmd
individual variation part finished.

@@
 bibliography: bibliography.bib
-bibstyle: spbasic
+bibstyle: spmpsci
 # bibstyle options spbasic(default), spphys, spmpsci
@@
 Pathbreaking attempts to infer specific cognitive capacities from this evidence largely focused on the basic requirements of tool production [@isaac1976; @wynn1979; @gowlett1984; @wynn2004].
-More recently, increasing attention has been directed to the processes and demands of stone tool making skill acquisition [@roux1995; @stout2002; @stout2005; @geribàs2010; @nonaka2010; @stout2011; @putt2014; @hecht2015; @duke2015; @morgan2015; @stout2015; @lombao2017; @putt2017; @cataldo2018; @putt2019; @pargeter2019; @pargeter2020].
+More recently, increasing attention has been directed to the processes and demands of stone tool making skill acquisition [@roux1995; @stout2002; @stout2005; @geribàs2010; @nonaka2010; @stout2011; @putt2014; @hecht2015a; @duke2015; @morgan2015; @stout2015; @lombao2017; @putt2017; @cataldo2018; @putt2019; @pargeter2019; @pargeter2020].
 This is motivated by the expectation that the most salient cognitive demands of tool making should occur during learning rather than routine expert performance [@stout2015] and by interest in the relevance of different social learning mechanisms such as imitation [@rein2014; @stout2019], emulation [@tehrani2008; @wilkins2018], and language [@ohnuma1997; @putt2014; @morgan2015; @lombao2017; @putt2017; @cataldo2018] to the reproduction of Paleolithic technologies.
+
+Studies investigating these questions have used a range of different experimental designs (e.g., varying technological goals/instructions, training times, raw materials, live vs.
+recorded instruction, lithic/skill assessment metrics, pseudo-knapping tasks etc.) and reached disparate conclusions regarding the neurocognitive and social foundations of skill acquisition.
+It is plausible that these discordant results reflect actual diversity in how humans acquire and master stone tool making skills.
+However, this failure of results to generalize across artificial experimental manipulations [[cf](cf.%20Yarkoni%202020).
+@yarkoni2020] also raises doubts regarding the external validity [@eren2016] of conclusions with respect to real-world Paleolithic learning contexts.
+To address this, we conducted an exploratory study that draws on lessons from previous research in an attempt to balance the pragmatic and theoretical tradeoffs inherent in experimental studies of stone knapping skill acquisition [@pargeter2019; @stout2015].
+
+Learning real-world skills like stone knapping is highly demanding of time and materials and difficult to control experimentally without sacrificing generalizability to real world conditions.
+Prior efforts have attempted to navigate these challenges by using various combinations of 1) inauthentic raw materials that are less expensive, easier to standardize, and/or easier to knap, 2) video-recorded instruction that is uniform across participants and less demanding of experimenter time, 3) short learning periods, 4) small sample sizes, and 5) single learning conditions.
+The difficulty of interpreting results from this growing literature led Stout and Khreisheh [-@stout2015: 870, emphasis original] to call for "studies with sufficient sample sizes to manipulate learning conditions (e.g.
+instruction, motivation) and assess individual variation (e.g.
+performance, psychometrics, neuroanatomy) that *also* have realistic learning periods." The current study attempts to strike a viable balance between these demands by investigating early-stage learning of a relatively simple technology (least effort, "Oldowan," flake production [@reti2016; @shea2016] under two instructional conditions while collecting data on individual differences in strength, coordination, cognition, social learning, self-control, and task engagement.
+Unlike any previous study, this allows us to address the likelihood that group effects of training conditions might be impacted by interactions with individual participant differences in aptitude, motivation, or learning style.
+
+We focus on early stage learning because it has been found to be relatively rapid, variable across individuals, and predictive of later outcomes [@pargeter2019; @stout2015; @putt2019], and thus provides a reasonable expectation of generating meaningful data on skill and learning variation while minimizing training costs.
+Moreover, understanding the minimum training times necessary to detect changes in tool making skill will help archaeologists design more realistic and cost-effective experiments.
+To further manage costs, we limited our study to only two learning conditions (observation only vs.
+active teaching).
+This targets a key controversy in human evolution, namely the origins of teaching and language [@gärdenfors2017; @morgan2015], while avoiding highly artificial manipulations of dubious relevance to real-world Paleolithic learning.
+These choices allowed us to invest more in other aspects of research design that we identified as theoretically important, including measurement of individual differences in cognition and behavior, inclusion of an in-person, fully interactive teaching condition, and use of naturalistic raw materials.
+Sample size remained small in this internally funded exploratory study but could easily be scaled up at funding levels typical of pre- and post-doctoral research grants in archaeology.
 
@@
 
+"*The many slight differences...
+being observed in the individuals of the same species inhabiting the same confined locality, may be called individual differences...
+These individual differences are of the highest importance to us, for they are often inherited ...
+and they thus afford materials for natural selection to act on and accumulate...*" [@darwin1859, Chapter 2]
+
+Individuals vary in aptitude and learning style for particular skills [@jonassen1993] but this has largely been ignored in studies of knapping skill acquisition, which have instead focused on group effects of different experimental conditions.
+There are good pragmatic reasons for this, as individual difference studies typically require larger sample sizes and additional data collection.
+However, overlooking these distinctions is not ideal since individual differences can provide valuable insight into the mechanisms, development, and evolution of cognition and behavior [@boogert2018].
+In particular, patterns of association between cognitive traits and behavioral performance can be used to test hypotheses about the cognitive demands of learning particular skills and the likely targets of natural selection acting on aptitude.
+More prosaically, individual differences can introduce an unexamined and uncontrolled source of variation in group level results.
+This is especially true in the relatively small "samples of convenience" typical of experimental archaeology.
+
+While testing hypotheses in evolutionary cognitive archaeology remains a considerable challenge [@wynn2017], investigation of individual variation in modern research participants represents one promising direction.
+For any particular behavior of archaeological interest, it is expected that standing variation in modern populations should remain relevant to normal variation in learning aptitude.
+The presence of trait variation without impact on learning aptitude would provide strong evidence against the plausibility of the proposed evolutionary relationship.
+An absence of variation (i.e., past fixation and rigorous developmental canalization) is not expected given the known variability of human brains and cognition [@sherwood2017; @barrett2020].
+Any confirmatory findings of trait-aptitude correspondence would then have the testable implication that humans should be evolutionarily derived along the same dimension [e.g.
+@hecht2015b].
+
+To date, a small number of "neuroarchaeological" studies have reported associations between individual knapping performance and brain structure or physiological responses.
+Hecht et al.
+[-@hecht2015a] reported training-related changes in white matter integrity (fractional anisotropy [FA]) that correlated with individual differences in practice time and striking accuracy change.
+The regional patterning of FA changes also varied across individuals, with only those individuals who displayed early increases in FA under the right ventral precentral gyrus (premotor cortex involved in movement planning and guidance) showing striking accuracy improvement over the training period.
+Putt et al.
+[-@putt2019] similarly found that the proportion of flakes to shatter produced by individuals during handaxe making correlated with dorsal precentral gyrus (motor cortex) activation.
+Pargeter et al.
+[-@pargeter2020] used a flake prediction paradigm [modeled after @nonaka2010] to confirm that striking force and accuracy are important determinants of handaxe-making success.
+These findings all point to the central role of perceptual-motor systems [@stout2007] and coordination [@roux1995] in knapping skill acquisition.
+In addition, Putt et al.
+[-@putt2019] also found successful flake production to be associated with prefrontal (working memory/cognitive control) activation and Stout et al.
+(2015) found that prefrontal activation correlated with success at a strategic judgement (platform selection) task which in turn was predictive of success at out-of-scanner handaxe production.
+Such investigations are thus starting to chart out the more specific contributions of different neural systems to particular aspects of knapping skill acquisition.
+To date, however, the cognitive/functional interpretation of systems identified in this manner has largely relied on informal reverse inference (reasoning backward from observed activations to inferred mental processes) from published studies of other tasks that activated the same regions, an approach which is widely regarded as problematic [@poldrack2011].
+
+Here we take a more direct, psychometric approach to measuring individual differences in perceptual-motor coordination and cognition.
+Psychometric instruments (e.g., tasks, questionnaires) are designed to assess variation in cognitive traits and states, such as fluid intelligence, working memory, attention, motivation, and personality, that have been of theoretical interest to cognitive archaeologists [e.g., @wynn2016].
+It is thus surprising that they have been almost entirely neglected in experimental studies of knapping skill.
+In the only published example we are aware of, Pargeter et al.
+[-@pargeter2019] reported significant effects of variation in planning and problem solving (Tower of London test [@shallice1982]) and cognitive set shifting (Wisconsin Card Sort test [@grant1948]) on early stage handaxe learning.
+Of course, cognition is not the only thing that can affect knapping performance.
+Flake prediction experiments highlight the importance of regulating movement speed/accuracy trade-offs [@nonaka2010; @pargeter2020] and studies of muscle recruitment [@marzke1998] and manual pressure [@williams-hatala2018; @key2018] during knapping highlight basic strength requirements.
+Along these lines, Key and Lycett [-@key2019] found that individual differences in hand size, shape, and especially grip strength were better predictors of force loading during stone tool use than were attributes of the tools themselves.
+However, we are unaware of any such studies of biometric influences on variation in knapping success.
+Finally, the time and effort demands of knapping skill acquisition suggest that differences in personality (e.g., self-control and "grit" [@pargeter2019], motivation [@stout2002], and social vs.
+individual learning strategies [@miu2020] might also affect learning outcomes.
+We are again unaware of any previous studies that have assessed such effects.
+In this study, we assessed all participants with a battery of tests including grip strength, movement speed/accuracy, spatial working memory, fluid intelligence, self-control, tendency to use social information, and motivation/engagement with the tool making task.
+We were particularly interested in the possibility that these variables might not only impact learning generally, but might also have different effects under different learning conditions.
+
 ## **Teaching, Language, and Tool Making**
+
+*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.* [@montagu1976: 267]
+
+Possible links between tool making and language have been a subject of speculation for nearly 150 years [@engles2003[1873]], if not longer [@hewes1993], although compelling empirical tests have remained elusive.
+Over 25 years ago, Toth and Schick [-@toth1993] suggested that experiments teaching modern participants to make stone tools in verbal and non-verbal conditions could test the importance of language in the social reproduction of Paleolithic technologies.
+Ohnuma et al.
+[-@ohnuma1997] were the first to implement this suggestion in a study of Levallois flake production, followed by more recent studies of handaxe making [@putt2014; @putt2017] and simple flake production [@morgan2015; @cataldo2018; @lombao2017].
+This reflects recent interest in the hypothesis that language might be an adaptation for teaching [e.g., @laland2017; @stout2012].
+Teaching and learning demands of Paleolithic tool making would thus provide evidence of selective contexts favoring language evolution [@stout2010; @morgan2015; @montagu1976].
+
+\
 
@@
 ## **Participants**
+
 ## **Study Visit**
@@
 ### **Raw Materials**
+
 ### **Experimental Conditions**
+
 ## **Lithic Analysis**
+
 ## **Statistical Analyses**
@@
 # **Results**
+
 ## **Principal Component analyses**
@@
 ### **Flake size and shape**
+
 ### **Lithic flaking performance measures**
@@
 
-## **Do individual differences in motor skill and psychometric measures predict flaking performance? **
+## **Do individual differences in motor skill and psychometric measures predict flaking performance?**
+
 ### **Model 1: Individual differences and quantity flaking**
+
 ### **Model 2: Individual differences and quality flaking**

### 2021-07-29 manuscript/Testing-the-cognitive-foundations-of-Paleolithic-social-transmission.tex
individual variation part finished.

@@
 2005; Geribàs, Mosquera, and Vergès 2010; Nonaka, Bril, and Rein 2010;
-Stout et al. 2011; Putt, Woods, and Franciscus 2014; Hecht et al. 2015;
-Duke and Pargeter 2015; Morgan et al. 2015; Stout and Khreisheh 2015;
-Lombao, Guardiola, and Mosquera 2017; Putt et al. 2017; Cataldo,
-Migliano, and Vinicius 2018; Putt, Wijeakumar, and Spencer 2019;
-Pargeter and Shea 2019; Pargeter et al. 2020). This is motivated by the
-expectation that the most salient cognitive demands of tool making
-should occur during learning rather than routine expert performance
-(Stout and Khreisheh 2015) and by interest in the relevance of different
-social learning mechanisms such as imitation (Rein, Nonaka, and Bril
-2014; Stout et al. 2019), emulation (Tehrani and Riede 2008; Wilkins
-2018), and language (Ohnuma, Aoki, and Akazawa 1997; Putt, Woods, and
-Franciscus 2014; Morgan et al. 2015; Lombao, Guardiola, and Mosquera
-2017; Putt et al. 2017; Cataldo, Migliano, and Vinicius 2018) to the
-reproduction of Paleolithic technologies.
+Stout et al. 2011; Putt, Woods, and Franciscus 2014; Hecht, Gutman,
+Khreisheh, et al. 2015; Duke and Pargeter 2015; Morgan et al. 2015;
+Stout and Khreisheh 2015; Lombao, Guardiola, and Mosquera 2017; Putt et
+al. 2017; Cataldo, Migliano, and Vinicius 2018; Putt, Wijeakumar, and
+Spencer 2019; Pargeter and Shea 2019; Pargeter et al. 2020). This is
+motivated by the expectation that the most salient cognitive demands of
+tool making should occur during learning rather than routine expert
+performance (Stout and Khreisheh 2015) and by interest in the relevance
+of different social learning mechanisms such as imitation (Rein, Nonaka,
+and Bril 2014; Stout et al. 2019), emulation (Tehrani and Riede 2008;
+Wilkins 2018), and language (Ohnuma, Aoki, and Akazawa 1997; Putt,
+Woods, and Franciscus 2014; Morgan et al. 2015; Lombao, Guardiola, and
+Mosquera 2017; Putt et al. 2017; Cataldo, Migliano, and Vinicius 2018)
+to the reproduction of Paleolithic technologies.
+
+Studies investigating these questions have used a range of different
+experimental designs (e.g., varying technological goals/instructions,
+training times, raw materials, live vs.~recorded instruction,
+lithic/skill assessment metrics, pseudo-knapping tasks etc.) and reached
+disparate conclusions regarding the neurocognitive and social
+foundations of skill acquisition. It is plausible that these discordant
+results reflect actual diversity in how humans acquire and master stone
+tool making skills. However, this failure of results to generalize
+across artificial experimental manipulations
+(\href{cf.\%20Yarkoni\%202020}{cf}. Yarkoni 2020) also raises doubts
+regarding the external validity (Eren et al. 2016) of conclusions with
+respect to real-world Paleolithic learning contexts. To address this, we
+conducted an exploratory study that draws on lessons from previous
+research in an attempt to balance the pragmatic and theoretical
+tradeoffs inherent in experimental studies of stone knapping skill
+acquisition (Pargeter and Shea 2019; Stout and Khreisheh 2015).
+
+Learning real-world skills like stone knapping is highly demanding of
+time and materials and difficult to control experimentally without
+sacrificing generalizability to real world conditions. Prior efforts
+have attempted to navigate these challenges by using various
+combinations of 1) inauthentic raw materials that are less expensive,
+easier to standardize, and/or easier to knap, 2) video-recorded
+instruction that is uniform across participants and less demanding of
+experimenter time, 3) short learning periods, 4) small sample sizes, and
+5) single learning conditions. The difficulty of interpreting results
+from this growing literature led Stout and Khreisheh (2015: 870,
+emphasis original) to call for ``studies with sufficient sample sizes to
+manipulate learning conditions (e.g.~instruction, motivation) and assess
+individual variation (e.g.~performance, psychometrics, neuroanatomy)
+that \emph{also} have realistic learning periods.'' The current study
+attempts to strike a viable balance between these demands by
+investigating early-stage learning of a relatively simple technology
+(least effort, ``Oldowan,'' flake production (Reti 2016; Shea 2016)
+under two instructional conditions while collecting data on individual
+differences in strength, coordination, cognition, social learning,
+self-control, and task engagement. Unlike any previous study, this
+allows us to address the likelihood that group effects of training
+conditions might be impacted by interactions with individual participant
+differences in aptitude, motivation, or learning style.
+
+We focus on early stage learning because it has been found to be
+relatively rapid, variable across individuals, and predictive of later
+outcomes (Pargeter and Shea 2019; Stout and Khreisheh 2015; Putt,
+Wijeakumar, and Spencer 2019), and thus provides a reasonable
+expectation of generating meaningful data on skill and learning
+variation while minimizing training costs. Moreover, understanding the
+minimum training times necessary to detect changes in tool making skill
+will help archaeologists design more realistic and cost-effective
+experiments. To further manage costs, we limited our study to only two
+learning conditions (observation only vs.~active teaching). This targets
+a key controversy in human evolution, namely the origins of teaching and
+language (Gärdenfors and Högberg 2017; Morgan et al. 2015), while
+avoiding highly artificial manipulations of dubious relevance to
+real-world Paleolithic learning. These choices allowed us to invest more
+in other aspects of research design that we identified as theoretically
+important, including measurement of individual differences in cognition
+and behavior, inclusion of an in-person, fully interactive teaching
+condition, and use of naturalistic raw materials. Sample size remained
+small in this internally funded exploratory study but could easily be
+scaled up at funding levels typical of pre- and post-doctoral research
+grants in archaeology.
 
@@
 
+``\emph{The many slight differences\ldots{} being observed in the
+individuals of the same species inhabiting the same confined locality,
+may be called individual differences\ldots{} These individual
+differences are of the highest importance to us, for they are often
+inherited \ldots{} and they thus afford materials for natural selection
+to act on and accumulate\ldots{}}'' (Darwin 1859, Chapter 2)
+
+Individuals vary in aptitude and learning style for particular skills
+(Jonassen and Grabowski 1993) but this has largely been ignored in
+studies of knapping skill acquisition, which have instead focused on
+group effects of different experimental conditions. There are good
+pragmatic reasons for this, as individual difference studies typically
+require larger sample sizes and additional data collection. However,
+overlooking these distinctions is not ideal since individual differences
+can provide valuable insight into the mechanisms, development, and
+evolution of cognition and behavior (Boogert et al. 2018). In
+particular, patterns of association between cognitive traits and
+behavioral performance can be used to test hypotheses about the
+cognitive demands of learning particular skills and the likely targets
+of natural selection acting on aptitude. More prosaically, individual
+differences can introduce an unexamined and uncontrolled source of
+variation in group level results. This is especially true in the
+relatively small ``samples of convenience'' typical of experimental
+archaeology.
+
+While testing hypotheses in evolutionary cognitive archaeology remains a
+considerable challenge (Wynn 2017), investigation of individual
+variation in modern research participants represents one promising
+direction. For any particular behavior of archaeological interest, it is
+expected that standing variation in modern populations should remain
+relevant to normal variation in learning aptitude. The presence of trait
+variation without impact on learning aptitude would provide strong
+evidence against the plausibility of the proposed evolutionary
+relationship. An absence of variation (i.e., past fixation and rigorous
+developmental canalization) is not expected given the known variability
+of human brains and cognition (Sherwood and Gómez-Robles 2017; Barrett
+2020). Any confirmatory findings of trait-aptitude correspondence would
+then have the testable implication that humans should be evolutionarily
+derived along the same dimension (e.g. Hecht, Gutman, Bradley, et al.
+2015).
+
+To date, a small number of ``neuroarchaeological'' studies have reported
+associations between individual knapping performance and brain structure
+or physiological responses. Hecht et al. (2015) reported
+training-related changes in white matter integrity (fractional
+anisotropy {[}FA{]}) that correlated with individual differences in
+practice time and striking accuracy change. The regional patterning of
+FA changes also varied across individuals, with only those individuals
+who displayed early increases in FA under the right ventral precentral
+gyrus (premotor cortex involved in movement planning and guidance)
+showing striking accuracy improvement over the training period. Putt et
+al. (2019) similarly found that the proportion of flakes to shatter
+produced by individuals during handaxe making correlated with dorsal
+precentral gyrus (motor cortex) activation. Pargeter et al. (2020) used
+a flake prediction paradigm (modeled after Nonaka, Bril, and Rein 2010)
+to confirm that striking force and accuracy are important determinants
+of handaxe-making success. These findings all point to the central role
+of perceptual-motor systems (Stout and Chaminade 2007) and coordination
+(Roux, Bril, and Dietrich 1995) in knapping skill acquisition. In
+addition, Putt et al. (2019) also found successful flake production to
+be associated with prefrontal (working memory/cognitive control)
+activation and Stout et al.~(2015) found that prefrontal activation
+correlated with success at a strategic judgement (platform selection)
+task which in turn was predictive of success at out-of-scanner handaxe
+production. Such investigations are thus starting to chart out the more
+specific contributions of different neural systems to particular aspects
+of knapping skill acquisition. To date, however, the
+cognitive/functional interpretation of systems identified in this manner
+has largely relied on informal reverse inference (reasoning backward
+from observed activations to inferred mental processes) from published
+studies of other tasks that activated the same regions, an approach
+which is widely regarded as problematic (Poldrack 2011).
+
+Here we take a more direct, psychometric approach to measuring
+individual differences in perceptual-motor coordination and cognition.
+Psychometric instruments (e.g., tasks, questionnaires) are designed to
+assess variation in cognitive traits and states, such as fluid
+intelligence, working memory, attention, motivation, and personality,
+that have been of theoretical interest to cognitive archaeologists
+(e.g., Wynn and Coolidge 2016). It is thus surprising that they have
+been almost entirely neglected in experimental studies of knapping
+skill. In the only published example we are aware of, Pargeter et al.
+(2019) reported significant effects of variation in planning and problem
+solving (Tower of London test (Shallice, Broadbent, and Weiskrantz
+1982)) and cognitive set shifting (Wisconsin Card Sort test (Grant and
+Berg 1948)) on early stage handaxe learning. Of course, cognition is not
+the only thing that can affect knapping performance. Flake prediction
+experiments highlight the importance of regulating movement
+speed/accuracy trade-offs (Nonaka, Bril, and Rein 2010; Pargeter et al.
+2020) and studies of muscle recruitment (Marzke et al. 1998) and manual
+pressure (Williams-Hatala et al. 2018; Alastair J. M. Key and Dunmore
+2018) during knapping highlight basic strength requirements. Along these
+lines, Key and Lycett (2019) found that individual differences in hand
+size, shape, and especially grip strength were better predictors of
+force loading during stone tool use than were attributes of the tools
+themselves. However, we are unaware of any such studies of biometric
+influences on variation in knapping success. Finally, the time and
+effort demands of knapping skill acquisition suggest that differences in
+personality (e.g., self-control and ``grit'' (Pargeter and Shea 2019),
+motivation (Stout 2002), and social vs.~individual learning strategies
+(Miu et al. 2020) might also affect learning outcomes. We are again
+unaware of any previous studies that have assessed such effects. In this
+study, we assessed all participants with a battery of tests including
+grip strength, movement speed/accuracy, spatial working memory, fluid
+intelligence, self-control, tendency to use social information, and
+motivation/engagement with the tool making task. We were particularly
+interested in the possibility that these variables might not only impact
+learning generally, but might also have different effects under
+different learning conditions.~
+
 \hypertarget{teaching-language-and-tool-making}{%
@@
 Making}}{Teaching, Language, and Tool Making}}\label{teaching-language-and-tool-making}}
+
+\emph{A creature that learns to make tools to a complex pre-existing
+pattern\ldots must have the kind of abstracting mind that would be of
+high selective value in facilitating the development of the ability to
+communicate such skills by the necessary verbal acts.} (Montagu 1976:
+267)
+
+Possible links between tool making and language have been a subject of
+speculation for nearly 150 years (Engles 2003, {[}1873{]}), if not
+longer (Hewes 1993), although compelling empirical tests have remained
+elusive. Over 25 years ago, Toth and Schick (1993) suggested that
+experiments teaching modern participants to make stone tools in verbal
+and non-verbal conditions could test the importance of language in the
+social reproduction of Paleolithic technologies. Ohnuma et al. (1997)
+were the first to implement this suggestion in a study of Levallois
+flake production, followed by more recent studies of handaxe making
+(Putt, Woods, and Franciscus 2014; Putt et al. 2017) and simple flake
+production (Morgan et al. 2015; Cataldo, Migliano, and Vinicius 2018;
+Lombao, Guardiola, and Mosquera 2017). This reflects recent interest in
+the hypothesis that language might be an adaptation for teaching (e.g.,
+Laland 2017; Stout and Chaminade 2012). Teaching and learning demands of
+Paleolithic tool making would thus provide evidence of selective
+contexts favoring language evolution (Stout 2010; Morgan et al. 2015;
+Montagu 1976).
+
+\hfill\break
 
@@
 \subsection{\texorpdfstring{\textbf{Do individual differences in motor
-skill and psychometric measures predict flaking performance?
-}}{Do individual differences in motor skill and psychometric measures predict flaking performance?
-}}\label{do-individual-differences-in-motor-skill-and-psychometric-measures-predict-flaking-performance}}
+skill and psychometric measures predict flaking
+performance?}}{Do individual differences in motor skill and psychometric measures predict flaking performance?}}\label{do-individual-differences-in-motor-skill-and-psychometric-measures-predict-flaking-performance}}
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-barrett2020}{}%
+Barrett, H. Clark. 2020. {``Towards a Cognitive Science of the Human:
+Cross-Cultural Approaches and Their Urgency.''} \emph{Trends in
+Cognitive Sciences} 24 (8): 620--38.
+\url{https://doi.org/10.1016/j.tics.2020.05.007}.
+
+\leavevmode\hypertarget{ref-boogert2018}{}%
+Boogert, Neeltje J., Joah R. Madden, Julie Morand-Ferron, and Alex
+Thornton. 2018. {``Measuring and Understanding Individual Differences in
+Cognition.''} \emph{Philosophical Transactions of the Royal Society B:
+Biological Sciences} 373 (1756): 20170280.
+\url{https://doi.org/10.1098/rstb.2017.0280}.
+
 \leavevmode\hypertarget{ref-cataldo2018}{}%
@@
 
+\leavevmode\hypertarget{ref-darwin1859}{}%
+Darwin, Charles. 1859. \emph{On the Origin of Species by Means of
+Natural Selection, or, The Preservation of Favoured Races in the
+Struggle for Life}. 1st ed. London: John Murray.
+
 \leavevmode\hypertarget{ref-darwin1871}{}%
-Darwin, Charles. 1871. \emph{The Descent of Man, and Selection in
-Relation to Sex}. 1st ed. London: John Murray.
+---------. 1871. \emph{The Descent of Man, and Selection in Relation to
+Sex}. 1st ed. London: John Murray.
 
@@
 \url{https://doi.org/10.1179/2051618515Y.0000000016}.
+
+\leavevmode\hypertarget{ref-engles2003}{}%
+Engles, Friedrich. 2003. {``The Part Played by Labour in the Transition
+from Ape to Man.''} In, edited by Robert C. Scharff and Val Dusek,
+71--77. London: Blackwell.
+
+\leavevmode\hypertarget{ref-eren2016}{}%
+Eren, Metin I., Stephen J. Lycett, Robert J. Patten, Briggs Buchanan,
+Justin Pargeter, and Michael J. O'Brien. 2016. {``Test, Model, and
+Method Validation: The Role of Experimental Stone Artifact Replication
+in Hypothesis-Driven Archaeology.''} \emph{Ethnoarchaeology: Journal of
+Archaeological, Ethnographic and Experimental Studies} 8 (2): 103--36.
+\url{https://doi.org/10.1080/19442890.2016.1213972}.
+
+\leavevmode\hypertarget{ref-guxe4rdenfors2017}{}%
+Gärdenfors, Peter, and Anders Högberg. 2017. {``The Archaeology of
+Teaching and the Evolution of Homo Docens.''} \emph{Current
+Anthropology} 58 (2): 188--208. \url{https://doi.org/10.1086/691178}.
 
@@
 
-\leavevmode\hypertarget{ref-hecht2015}{}%
-Hecht, E. E., D. A. Gutman, N. Khreisheh, S. V. Taylor, J. Kilner, A. A.
-Faisal, B. A. Bradley, T. Chaminade, and D. Stout. 2015. {``Acquisition
-of Paleolithic Toolmaking Abilities Involves Structural Remodeling to
-Inferior Frontoparietal Regions.''} \emph{Brain Structure and Function}
-220 (4): 2315--31. \url{https://doi.org/10.1007/s00429-014-0789-6}.
+\leavevmode\hypertarget{ref-grant1948}{}%
+Grant, David A., and Esta Berg. 1948. {``A Behavioral Analysis of Degree
+of Reinforcement and Ease of Shifting to New Responses in a Weigl-Type
+Card-Sorting Problem.''} \emph{Journal of Experimental Psychology} 38
+(4): 404--11. \url{https://doi.org/10.1037/h0059831}.
+
+\leavevmode\hypertarget{ref-hecht2015b}{}%
+Hecht, Erin E., David A. Gutman, Bruce A. Bradley, Todd M. Preuss, and
+Dietrich Stout. 2015. {``Virtual dissection and comparative connectivity
+of the superior longitudinal fasciculus in chimpanzees and humans.''}
+\emph{NeuroImage} 108 (March): 124--37.
+\url{https://doi.org/10.1016/j.neuroimage.2014.12.039}.
+
+\leavevmode\hypertarget{ref-hecht2015a}{}%
+Hecht, Erin E., David. A. Gutman, Nada Khreisheh, S. V. Taylor, J.
+Kilner, A. A. Faisal, Bruce A. Bradley, T. Chaminade, and D. Stout.
+2015. {``Acquisition of Paleolithic toolmaking abilities involves
+structural remodeling to inferior frontoparietal regions.''} \emph{Brain
+Structure \& Function} 220 (4): 2315--31.
+\url{https://doi.org/10.1007/s00429-014-0789-6}.
+
+\leavevmode\hypertarget{ref-hewes1993}{}%
+Hewes, Gordon W. 1993. {``A History of Speculation on the Relation
+Between Tools and Language.''} In, edited by Kathleen R. Gibson and Tim
+Ingold, 20--31. Cambridge: Cambridge University Press.
 
@@
 
+\leavevmode\hypertarget{ref-jonassen1993}{}%
+Jonassen, David H., and Barbara L. Grabowski. 1993. \emph{Handbook of
+Individual Differences, Learning, and Instruction}. Hillsdale, NJ:
+Lawrence Erlbaum,.
+
+\leavevmode\hypertarget{ref-key2019}{}%
+Key, A. J. M., and S. J. Lycett. 2019. {``Biometric Variables Predict
+Stone Tool Functional Performance More Effectively Than Tool-Form
+Attributes: A Case Study in Handaxe Loading Capabilities.''}
+\emph{Archaeometry} 61 (3): 539--55.
+\url{https://doi.org/10.1111/arcm.12439}.
+
+\leavevmode\hypertarget{ref-key2018}{}%
+Key, Alastair J. M., and Christopher J. Dunmore. 2018. {``Manual
+Restrictions on Palaeolithic Technological Behaviours.''} \emph{PeerJ} 6
+(August): e5399. \url{https://doi.org/10.7717/peerj.5399}.
+
+\leavevmode\hypertarget{ref-laland2017}{}%
+Laland, Kevin N. 2017. {``The Origins of Language in Teaching.''}
+\emph{Psychonomic Bulletin \& Review} 24 (1): 225--31.
+\url{https://doi.org/10.3758/s13423-016-1077-7}.
+
 \leavevmode\hypertarget{ref-lombao2017}{}%
@@
 (1): 1--14. \url{https://doi.org/10.1038/s41598-017-14322-y}.
+
+\leavevmode\hypertarget{ref-marzke1998}{}%
+Marzke, Mary W., N. Toth, K. Schick, S. Reece, B. Steinberg, K. Hunt, R.
+L. Linscheid, and K.-N. An. 1998. {``EMG Study of Hand Muscle
+Recruitment During Hard Hammer Percussion Manufacture of Oldowan
+Tools.''} \emph{American Journal of Physical Anthropology} 105 (3):
+315--32.
+\url{https://doi.org/10.1002/(SICI)1096-8644(199803)105:3\%3C315::AID-AJPA3\%3E3.0.CO;2-Q}.
+
+\leavevmode\hypertarget{ref-miu2020}{}%
+Miu, Elena, Ned Gulley, Kevin N. Laland, and Luke Rendell. 2020.
+{``Flexible Learning, Rather Than Inveterate Innovation or Copying,
+Drives Cumulative Knowledge Gain.''} \emph{Science Advances} 6 (23):
+eaaz0286. \url{https://doi.org/10.1126/sciadv.aaz0286}.
+
+\leavevmode\hypertarget{ref-montagu1976}{}%
+Montagu, Ashley. 1976. {``Toolmaking, Hunting, and the Origin of
+Language.''} \emph{Annals of the New York Academy of Sciences} 280 (1):
+266--74. \url{https://doi.org/10.1111/j.1749-6632.1976.tb25493.x}.
 
@@
 
+\leavevmode\hypertarget{ref-poldrack2011}{}%
+Poldrack, Russell A. 2011. {``Inferring Mental States from Neuroimaging
+Data: From Reverse Inference to Large-Scale Decoding.''} \emph{Neuron}
+72 (5): 692--97. \url{https://doi.org/10.1016/j.neuron.2011.11.001}.
+
 \leavevmode\hypertarget{ref-putt2017}{}%
@@
 
+\leavevmode\hypertarget{ref-reti2016}{}%
+Reti, Jay S. 2016. {``Quantifying Oldowan Stone Tool Production at
+Olduvai Gorge, Tanzania.''} \emph{PLOS ONE} 11 (1): e0147352.
+\url{https://doi.org/10.1371/journal.pone.0147352}.
+
 \leavevmode\hypertarget{ref-roux1995}{}%
@@
 
+\leavevmode\hypertarget{ref-shallice1982}{}%
+Shallice, Timothy, Donald Eric Broadbent, and Lawrence Weiskrantz. 1982.
+{``Specific Impairments of Planning.''} \emph{Philosophical Transactions
+of the Royal Society of London. B, Biological Sciences} 298 (1089):
+199--209. \url{https://doi.org/10.1098/rstb.1982.0082}.
+
+\leavevmode\hypertarget{ref-shea2016}{}%
+Shea, John J. 2016. \emph{Stone Tools in Human Evolution: Behavioral
+Differences Among Technological Primates}. Cambridge: Cambridge
+University Press. \url{https://doi.org/10.1017/9781316389355}.
+
+\leavevmode\hypertarget{ref-sherwood2017}{}%
+Sherwood, Chet C., and Aida Gómez-Robles. 2017. {``Brain Plasticity and
+Human Evolution.''} \emph{Annual Review of Anthropology} 46 (1):
+399--419. \url{https://doi.org/10.1146/annurev-anthro-102215-100009}.
+
 \leavevmode\hypertarget{ref-stout2002}{}%
@@
 Anthropology} 43 (5): 693--722. \url{https://doi.org/10.1086/342638}.
+
+\leavevmode\hypertarget{ref-stout2010}{}%
+---------. 2010. {``Possible Relations Between Language and Technology
+in Human Evolution.''} In, edited by April Nowell and Iain Davidson,
+159184. Boulder, CO: University Press of Colorado.
+
+\leavevmode\hypertarget{ref-stout2007}{}%
+Stout, Dietrich, and Thierry Chaminade. 2007. {``The Evolutionary
+Neuroscience of Tool Making.''} \emph{Neuropsychologia} 45 (5):
+1091--1100.
+\url{https://doi.org/10.1016/j.neuropsychologia.2006.09.014}.
+
+\leavevmode\hypertarget{ref-stout2012}{}%
+---------. 2012. {``Stone Tools, Language and the Brain in Human
+Evolution.''} \emph{Philosophical Transactions of the Royal Society B:
+Biological Sciences} 367 (1585): 75--87.
+\url{https://doi.org/10.1098/rstb.2011.0099}.
 
@@
 
+\leavevmode\hypertarget{ref-toth1993}{}%
+Toth, Nicholas, and Kathy Schick. 1993. {``Early Stone Industries and
+Inferences Regarding Language and Cognition.''} In, edited by Kathleen
+R. Gibson and Tim Ingold, 346362. Cambridge: Cambridge University Press.
+
 \leavevmode\hypertarget{ref-washburn1960}{}%
@@
 
+\leavevmode\hypertarget{ref-williams-hatala2018}{}%
+Williams-Hatala, Erin Marie, Kevin G. Hatala, McKenzie Gordon, Alastair
+Key, Margaret Kasper, and Tracy L. Kivell. 2018. {``The Manual Pressures
+of Stone Tool Behaviors and Their Implications for the Evolution of the
+Human Hand.''} \emph{Journal of Human Evolution} 119 (June): 14--26.
+\url{https://doi.org/10.1016/j.jhevol.2018.02.008}.
+
 \leavevmode\hypertarget{ref-wynn1979}{}%
@@
 \emph{Man} 14 (3): 371--91. \url{https://doi.org/10.2307/2801865}.
+
+\leavevmode\hypertarget{ref-wynn2017}{}%
+---------. 2017. {``Evolutionary Cognitive Archaeology.''} In, edited by
+Thomas Wynn and Frederick Coolidge, 120. Oxford: Oxford University
+Press.
 
@@
 
+\leavevmode\hypertarget{ref-wynn2016}{}%
+---------. 2016. {``Archeological Insights into Hominin Cognitive
+Evolution.''} \emph{Evolutionary Anthropology: Issues, News, and
+Reviews} 25 (4): 200--213. \url{https://doi.org/10.1002/evan.21496}.
+
+\leavevmode\hypertarget{ref-yarkoni2020}{}%
+Yarkoni, Tal. 2020. {``The Generalizability Crisis.''} \emph{Behavioral
+and Brain Sciences}, 1--37.
+\url{https://doi.org/10.1017/S0140525X20001685}.
+
 \end{CSLReferences}
 
-\bibliographystyle{spbasic}
+\bibliographystyle{spmpsci}
 \bibliography{bibliography.bib}

### 2021-07-31 manuscript/Testing the cognitive foundations of Paleolithic social transmission.Rmd
update the text to the latest google doc version

@@
 [-@putt2019] also found successful flake production to be associated with prefrontal (working memory/cognitive control) activation and Stout et al.
-(2015) found that prefrontal activation correlated with success at a strategic judgement (platform selection) task which in turn was predictive of success at out-of-scanner handaxe production.
+[-@stoutCognitiveDemandsLower2015] found that prefrontal activation correlated with success at a strategic judgement (platform selection) task which in turn was predictive of success at out-of-scanner handaxe production.
 Such investigations are thus starting to chart out the more specific contributions of different neural systems to particular aspects of knapping skill acquisition.
@@
 
-\
+Toth and Schick [-@toth1993] were, however, careful to point out that extinct hominid learning strategies and capacities might differ from modern experimental participants.
+Even leaving aside potential species differences in social learning [cf.
+@morgan2015; @stout2019], reliance on explicit verbal instruction varies widely across modern human societies [e.g., @boyette2017].
+The WEIRD (Western, educated, industrialized, rich, democratic [@henrich2010]) teachers and learners typical of knapping experiments arguably represent an extreme bias toward such instruction.
+Simply instructing such participants not to speak during an experiment (or to demonstrate but not gesture, etc.
+[@morgan2015]) is likely to underestimate the efficacy of non-verbal teaching and learning in cultural contexts where it is more common, let alone in a hypothetical pre-linguistic hominid species.
+
+Such concerns are exacerbated in experiments using pre-recorded instructional videos or extremely short training periods.
+Video does not allow the interactive teaching that is favored even in formal academic knapping classes [e.g., @shea2015] and is almost certainly typical of traditional learning contexts [e.g., @stout2002].
+It is not known how video presentation affects the efficacy of teaching generally, or the relative effectiveness of different forms of instruction.
+Going further, some experiments have manipulated the presence/absence of verbal instruction by presenting the same video with and without sound [@putt2017] or the sound track without the video [@cataldo2018].
+While this provides experimental control, it does not allow the instructor to adjust their multi-modal [@levinson2014] communication strategies as they would naturally do, for example through pointing and pantomime.
+To simply remove a communication channel without allowing any such adaptation is highly artificial and risks generating results that cannot be generalized beyond the specific context of the experiment [@yarkoni2020].
+Similarly, unnaturally short training periods (e.g., 5-15 minutes [@morgan2015; @lombao2017]) might misrepresent the relative efficacy of different teaching strategies under more realistic conditions [@whiten2015; @stout2015].
+Even the longest training times to date [@stout2015; @pargeter2019] have not produced knapping skills comparable to relevant archaeological examples, and were achieved by limiting sample size and using only one teaching condition.
+
+For these reasons, we sought to explore a middle path between experimental expedience and realism by limiting our experiment to two relatively naturalistic learning conditions and a moderate learning period of two hours.
+As in previous experiments [@stout2011; @hecht2015a; @pargeter2019] the first condition was unrestricted, interactive instruction in small groups, essentially reproducing the "natural" teaching/learning context familiar [cf.
+@shea2015] to our WEIRD instructor and student participants.
+The second condition allowed observation only, with the experimenter visible making flakes but not interacting in any way with learners.
+This absence of teaching is again a familiar social context for our participants and did not require any novel behaviors from the instructor.
+It matches the "imitation/emulation" condition of Morgan et al.
+[-@morgan2015] although we make no assumptions regarding learning mechanisms.
+We did not include a "reverse engineering" or "end-state emulation" condition in which only finished products were visible.
+This has been advocated as an important baseline or control condition [@whiten2015] to distinguish observational from individual learning, but is not likely to model any typical Paleolithic learning context nor to stand as an adequate proxy for the cognition of hominid species with different social learning capacities.
+There is no reason to assume neurocognitive and behavioral processes of reverse-engineering problem solving in modern humans [e.g., @allen2020] approximate the social learning processes of hominids with more ape-like action observation/imitation capacities [@hecht2013a; @hecht2013b; @stout2019].
+
+We selected a two-hour learning period for both pragmatic and theoretical reasons.
+Pargeter et al.
+[@pargeter2019] found that even \~90 hours of fully interactive instruction and practice was insufficient to achieve handaxe-making skills comparable to the later Acheulean site of Boxgrove [@garcía-medrano2019; @stout2014], and estimated actual time to mastery as ranging from 121 to 441 hours for different participants.
+However, they observed the greatest, fastest, and most individually variable skill increases during the first 20 hours of practice.
+In addition, initial performance was moderately correlated with later achievement.
+This suggests that studying early-stage learning may be a pragmatic alternative, especially for research investigating individual differences in aptitude.
+Studies of simple flake production similarly document large initial variation [@stout2015] and rapid early progress [@stout2015; @putt2019; @stoutKnappingSkillEarliest2006].
+We designed the current study to test the utility of studying learning and variation during the first two hours of simple flaking instruction/practice, in hopes of finding a viable compromise between experimental realism and cost
 
@@
 
+Lithic raw materials vary in size, shape, and fracture mechanical properties that affect the difficulty of achieving different knapping goals [@eren2014].
+Unfortunately, it can be difficult and/or expensive to procure authentic raw materials.
+Experimental studies of knapping skill have often used proxy materials such as flint [@morgan2015; @nonaka2010; @cataldo2018], limestone [@stoutKnappingSkillEarliest2006], porcelain [@khreisheh2013], or heat-treated chert [@putt2014; @putt2017; @putt2019]to model Oldowan and early Acheulean technologies executed in other materials.
+As well as being more readily available, these proxies are generally easier to knap.
+This has the benefit of reducing required practice time, but it is unclear how it might affect learning demands more generally or the efficacy of different learning conditions/strategies specifically.
+
+To address this, some studies have attempted to more closely match experimental and archaeological raw material types [@stout2011; @duke2015; @pargeter2019].
+However, raw materials vary across individual clasts within as well as between types.
+This has led to interest in standardizing experimental core morphology [@nonaka2010] and composition, even if this means using artificial materials such as porcelain [@khreisheh2013], brick [@geribàs2010; @lombao2017], or foam blocks [@schillinger2014].
+Such manipulations enhance experimental control and internal validity [@eren2016] at the expense of external generalizability to actual archaeological conditions.
+Specifically, they allow more robust results from smaller samples but eliminate a core element of real-world knapping skill: the ability to produce consistent results from variable materials [@pelegrin1990; @stout2013].
+For example, Pargeter et al.
+[-@pargeter2020] found that predicting specific flaking outcomes on actual handaxe preforms was both more difficult and less technologically important than expected from previous work with standardized, frustum-shaped cores [@nonaka2010].
+The alternative to control is to incorporate raw material size, shape, and composition as experimental variables [e.g., @stout2019].
+This allows consideration of raw material selection and response to variation as aspects of skill but correspondingly increases the sample sizes required to identify patterning.
+In considering these issues, we again chose to explore a middle path between pragmatism and realism by employing commercially purchased basalt similar to that known from East African Oldowan sites, allowing clast size and shape to vary within set limits, and selecting the particular clasts provided to each participant to approximate the same distribution.
+
+\
+
 # **Materials and Methods**
+
+This research was approved by the Emory Institutional Review Board (IRB00113024).
+All participants provided written informed consent and completed a video release form (<https://databrary.org/support/irb/release-template.html>).
 
@@
 
+Twenty-four adult participants with no prior stone knapping experience were recruited from the Emory community using paper fliers and e-mail listserv advertisements.
+We were unable to replace one participant who failed to attend their scheduled session, resulting in a total sample of 23.
+Eleven participants (6 female, 5 male) completed the Untaught condition and 12 (8 female, 4 male) completed the Taught condition.
+
 ## **Study Visit**
+
+Participants were asked to visit the Paleolithic Technology Lab at Emory University to complete one three-hour session.
+Participants were scheduled to attend in six groups of four, however one of these groups had only three participants due to a no-show on the day of the experiment.
+Each visit began with the collection of individual differences measures, which took approximately one hour.
+After that, participants undertook 105 minutes (two hours minus a 15-minute break after 1 hour) of stone tool making practice.
+This session was video-recorded, and all lithic products were collected.
+After the tool making task, participants completed an "exit questionnaire" comprising the Intrinsic Motivation Inventory (see below).
+
+Participants were compensated for their time with a \$30 gift card.
+They also had the opportunity to earn a performance bonus of \$5, \$10, \$15 or \$20 on the gift card.
+They were told that this bonus would depend on "how well they did" on the last core of their practice session.
+The actual performance measure was not specified, but in order to allow on the spot payment a simple measure of the percentage of starting weight removed from the final core was used such that: \> 30% earned \$5, \> 40% earned \$10, \> 50% earned \$15, \> 75% earned \$20.
+
+\
 
@@
 
+We used five individual difference measures for this study:
+
+1)  Grip strength was measured in kilograms using an electronic hand dynamometer (Camry EH101).
+Strength was measured twice and the higher value recorded.
+Grip strength is a simple measure that is well correlated with overall muscular strength [@wind2010] and a range of other health and fitness measures [@sasaki2007].
+It is hypothesized to be relevant to generating kinetic energy for fracture initiation [@nonaka2010] as well as control and support of the hammerstone [@williams-hatala2018] and core [@faisal2010; @key2015].
+
+2)  Motor accuracy was assessed using a "Fitts Law" reciprocal tapping task.
+Fitts Law describes the trade-off between speed and accuracy in human movement, classically measured by tapping back and forth between two targets of varying size and spacing [@fitts1954].
+Archaeologists have proposed [@stout2002; @pargeter2020] that management of this trade-off is critical to the accurate application of appropriate force seen in skilled knapping [@nonaka2010; @roux1995].
+We implemented this test on a Surface Pro tablet running free software (FittsStudy Version 4.2.8, default settings) developed by the Accessible Computing Experiences lab (Jacob O.
+Wobbrock, director) at the University of Washington ([depts.washington.edu/acelab/proj/fittsstudy/index.html](depts.washington.edu/acelab/proj/fittsstudy/index.html)).
+Participants use a touchscreen pen to tap between ribbons on the screen, with average movement time as the performance metric.
+
+3)  Visuospatial working memory is the capacity to "hold in mind", which researchers have hypothesized to be important in stone toolmaking performance [@coolidge2005].
+It also might support a learning process known as 'chunking,' in which multiple items or operations are combined into summary chunks stored in long term memory, that is thought to be important in the acquisition of knapping and other skills [@pargeter2019].
+We measured visuospatial working memory using a free n-back task ([wmp.education.uci.edu/software/](https://wmp.education.uci.edu/software/)) developed by the Working Memory and Plasticity Laboratory at the University of California, Irvine (Susanne Jaeggi, PI) and implemented in E-Prime software on a desktop computer.
+In this task, participants are asked to remember the position of blue squares presented sequentially on the screen and touch a key when the current position matches that 1, 2, 3...n iterations back.
+Progression to blocks with increasing values of n is contingent on exceeding a threshold success rate.
+Performance was measured as the highest n achieved.
+
+4)  Fluid intelligence [@cattell1963] refers to the capacity to engage in abstract reasoning and problem solving in a way that is minimally dependent on prior experience.
+It complements "crystallized intelligence" (the ability to apply learned procedures and knowledge) as one of the two factors (gf, gc) comprising so-called "general intelligence" (g).
+Fluid intelligence is closely related to the executive control of attention and manipulation of information held in working memory [@engle2018](Engle 2018).
+It is hypothesized to support technological innovation [@coolidge2005] and/or the intentional learning of new skills [@unsworth2005; @stout2015].
+We measured fluid intelligence using the short version [@bilker2012] of the classic Raven Progressive Matrices task, which requires participants to complete increasingly difficult pattern matching questions.
+
+5)  The use of social information for learning and decision making varies across individuals and societies [@molleman2019].
+Such variation is a key topic for understanding social learning and cultural evolutionary processes [@kendal2018; @heyes2018; @miu2020] and represents a potential confound for assessing experimental effects of different social learning conditions.
+We measured participants' tendency to rely on social information vs.
+their own insights using the Berlin Estimate AdjuStment Task (BEAST) developed by Molleman et al.
+[-@molleman2019].
+In this task, participants are present with large arrays of items on a screen and asked to estimate the number present.
+They are then provided with another person's estimate and allowed to provide a second estimate.
+The participants' average adjustment between first and second estimates provides a measure of their propensity to rely on social information.
+
+\
+
 ## **Stone Tool Making**
+
+After individual difference testing, participants engaged in a 2-hour stone tool making session, with a 15-minute break after 1 hour.
+Participants were instructed not to seek out additional training or information on stone tool making (i.e., via the internet) during these breaks.
+Each group of participants was randomly assigned to one of two experimental conditions: no teaching or teaching.
+In both conditions, participants were first given an opportunity to inspect and handle examples (**Figure**) of the kind of stone tools (flakes) they are being asked to produce.
+They were told that their objective was to produce as many flakes as possible from the materials provided.
+This meant that even the untaught condition included some minimal instruction (being told the objective) , however this was considered to be unavoidable without creating a much more elaborate and naturalistic context in which participants would develop their own technological goals.
+Such a design would also be expected to increase behavioral variability, demanding correspondingly larger samples of participants to identify patterns and making direct comparisons with the taught condition.
 
@@
 
+Each participant was provided with 9 cores for use over the 2-hour experiment.
+These cores were produced from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer.
+This produced irregular, angular chunks for use in the experiment, weighing between 459g - 1876g (mean = 975g).
+All cores were weighed, measured (Length, Width, Thickness), and painted white so that new fracture surfaces could be discriminated from those created during production.
+Cores were sorted by shape and weight and then distributed evenly to each participant.
+As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length x Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p = 0.9) of cores provided.
+This was also true comparing the two experimental conditions (Taught vs.
+Untaught mean weight = 1001g vs.
+956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4; mean shape = 221.43 vs.
+221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p = 0.05).
+Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores actually used across participants and conditions could still emerge as a result of selection bias.
+
+Sixty pounds of 3-to-5 inch basalt "Mexican Beach Pebbles" were purchased from a landscaping supply company for use as hammerstones in the experiment.
+Of these, 90 were selected as suitable for use.
+These weighed between 213g-1360g (mean = 425) and varied in elongation (L/W = 1.01 to 2.65) and relative thickness (LxW/T = 90.48 to 283.67).
+Forty-five stones were placed in the middle of the knapping area (Figure 2) for participants to freely choose from during the experiment.
+Broken hammerstones were replaced from the reserve to maintain a consistent number and range of choices.
+Each hammerstone was numbered and participants' choices were recorded along with the number of the core(s) being worked on with a particular hammerstone.\
+
 ### **Experimental Conditions**
 
-## **Lithic Analysis**
+In both conditions, three researchers were present to record activities and collect materials.
+Participants were seated in a circle [@shea2015] and experiments were video recorded using two cameras (Figure 1).
+Participants were free to select hammerstones from the common pile and to work on any or all of their nine assigned cores in any order they preferred.
+However, each core and all associated debitage were collected before participants were allowed to start working on a new core, so it was not possible to partially work and then return to a particular core later.
+The order of cores used and associated hammerstones were recorded for each participant during the experiment.
+
+In the untaught condition, a researcher (DS) sat with the participants and made stone tools but remained silent and made no effort to facilitate learning (e.g., through gesture, modified performance, facial expression, attention direction, or verbal instruction).
+Over the 2-hour period, the researcher completely reduced four cores (one every \~30 minutes).
+Participants were not restricted from talking to each other, as this would create an unnatural and potentially stressful social context that might affect learning.
+Participants were asked to avoid any form of communication about the tool making task specifically, and they complied with this request.
+Participants in this condition thus had the opportunity to observe tool making by an expert and/or by other learners, should they choose to do so, but received no intentional instruction.
+
+In the Taught condition, there were no restrictions on participant interaction and the researcher engaged in direct active teaching [@kline2015] of tool-making techniques through verbal instruction, demonstration, gesture, and shaping of behavior.
+The instructor has a moderate level of experience teaching basic knapping skills to students in undergraduate archaeology classes and to participants in previous knapping research [e.g., @stout2011].
+The pedagogical strategy employed was based on the instructor's own learning experiences and theoretical interpretations [e.g., @pargeter2020], and focused on coaching participants in effective body postures, movement patterns, and grips as well as the assessment of viable core morphology.
+
+**Lithic Analysis**
+
+All finished cores were weighed and measured (L, W, T).
+Delta weight was calculated as (Start weight-End weight)/Start weight.
+All detached pieces (DPs) were collected and weighed.
+We did not sort DPs into types (e.g., whole flakes, fragments) as this would have greatly increased processing time and it is not clear that such distinctions add relevant information regarding utility/desirability beyond that supplied by metrics [@stout2019].
+All DPs larger than 40mm in maximum dimension were photographed and measured.
+It is conventional in Early Stone Age lithic analysis to employ a 20 mm cut-off.
+We selected a higher threshold for both pragmatic (analysis time) and theoretical reasons.
+Flake use experiments have shown that flakes weighing less than 5--10 g or with a surface area below 7--10 cm2 [@prasciunas2007] or with a maximum dimension \<50-60 mm [@key2014] become markedly inefficient for basic cutting tasks.
+Similarly, data from Oldowan replication experiments [@stout2019] show that the utility index (flake cutting edge/flake mass1/3) \* (1 - exp[-0.31 \* (flake maximum dimension -- 1.81)]) developed by Morgan et al.
+(2015) falls off rapidly below 40mm maximum dimension ( Mean Utility \< 40mm = 0.508; \>=40mm = 0.946; t= 11.99, df = 707, p \< 0.000).
+By including weight in our cut-off criteria we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.
+
+For measurement, DP length was defined as the longest axis and width as the maximum dimension orthogonal to length.
+Thickness was defined as the maximum dimension orthogonal to the plane formed by L and W and was measured using calipers.
+L, W, and plan-view area measurements were taken from photographs captured using a Canon Rebel T3i fitted with a 60 mm macro lens and attached to a photographic stand with adjustable upper and lower light fittings.
+The camera was positioned directly above the flakes and kept at a constant height.
+DPs were positioned irrespective of any technological features so that the longest axis was vertical, and the wider end was placed toward the bottom of the photograph.
+
+Photographs were post-processed using Equalight software to adjust for lens and lighting falloff that result from bending light through a lens and its aperture which can affect measurements taken from photographs.
+Each image was shot with a scale that was then used to rectify the photograph's pixel scale to a real-world measurement scale in Adobe Photoshop.
+Images were converted to binary black and white format and silhouettes of the tools were extracted in Adobe Photoshop.
+We then used a custom ImageJ [@rueden2017] script [@pargeter2019] to measure DP length and take nine width measurements at 10% increments of length starting at the base of each DP.
+We used the built-in ImageJ tool to measure DP area.
+A "Proportion Larger DPs" was calculated per core as the combined weight of all DPs \>40mm in maximum dimension and 5g in weight divided by the weight of all DPs.
+Higher values show cores with proportionally more large DPs.
 
@@
 
+To evaluate the association between psychometric, motor-skill, and training measures and technological outcomes, we adopted an information-theoretic approach [@burnham2002].
+Information-theoretic approaches provide methods for model selection using all possible combinations of variables while avoiding problems associated with significance-threshold stepwise selection.
+We used the corrected Akaike information criterion (AICc) to rate each possible combination of predictors on the balance between goodness of fit (likelihood of the data given the model) and parsimony (number of parameters).
+The AICc consists of the log likelihood (i.e., how well does the model fit the data?) and a penalty term for the number of parameters that must be estimated in the model (i.e., how parsimonious is the model?), with a correction for small sample sizes (AICc converges to the standard AIC at large samples).
+A lower AICc indicates a more generalizable model and we used it to compare and rank various possible models.
+Each analysis begins with a full model that includes all predictors of interest.
+All possible combinations of predictors are then fit, and the resulting models are ranked and weighted based on their AICc.
+The "best" model is chosen because it has the lowest AICc score.
+
+Continuous predictors were centered such that zero represents the sample average, and units are standard deviations.
+The full model was fitted with the lm function in R 3.2.3, and the glmulti package [@Bartoń 2015] was used for multi-modal selection and model comparison.
+
+\
+
 # **Results**
+
+Following a recent protocol to enhance the reproducibility and data transparency of archaeological research [@marwick2017], detailed results of all analyses and assessments of the data structure are available in our paper's supplementary materials and through Github (<https://github.com/Raylc/PaST-pilot>).
+Here we limit discussion to the major ﬁndings regarding flaking performance and individual differences.
+The purpose of this section is to determine whether training impacted subject flaking performance and if any of the subject's individual psychometric and motor-skill aptitudes predict flaking performance.
 
@@
 
+The following two sections outline factor analyses designed to summarize our main study metrics tracking individual variation in flake sizes and shapes and lithic performance measures.
+
 ### **Flake size and shape**
+
+To better understand the relationship between flake shape and training/individual variation, we entered our nine flake linear plan measurements along with maximum flake length and thickness into a principal component analysis (PCA) from which summary coordinates were extracted.
+Bartlett's Test of Sphericity was significant ($\chi ^2$ (10) =4480, p \< .01) indicating that the set of variables are adequately related for factor analysis.
+
+The analysis yielded three factors explaining a total of 90% of the variance for the entire 11 measurement variable set **(Table**).
+Factor 1 tracks flake size with higher scores indicating larger flakes since all 11 measures load positively on this factor.
+Factor 2's loadings track the increasing relationship between thickness, length, and flake width.
+As factor 2 scores increase, flakes get thicker, longer, and narrow, resembling irregular splinters.
+Factor 3 tracks the relationship between flake proximal and distal width relative to thickness.
+As factor 3 scores go up, flakes get thinner and narrower at the distal ends and wider at the base.
+Factor 3 therefore tracks flakes with a typical shape having a thin cross-section, wider base, and narrower tip.
+We used these three flake shape coordinates to approximate flake size and shape in the project's flake performance factor analysis.
 
@@
 
+To better understand the relationship between our various lithic performance measurements and to reduce these data dimensionality, we conducted a second principal component analysis examining the study's seven lithic performance measures (mass of flakes relative to flaked core mass, count of large flakes [\>40mm and 5g], core delta mass, three flake shape factors, and the total number of cores used).
+All of these measures except the total cores flaked were summarized for each core.
+The Bartlett's Test of Sphericity was significant ($\chi ^2$ (6) =3950, p \< .01) indicating that the set of variables are at least adequately related for factor analysis.
+
+The analysis yielded two factors explaining a total of 52.3% of the variance for the entire set of variables.
+Factor 1 tracks flaking quantity due to high positive loadings on mass flakes/flaked mass, core delta mass, total cores used, and flake shape factor 2.
+Negative loadings on flake factor 1 (flake size) and 3 (basal/tip width shape) suggest that flaking quantity comes at the expense of producing lots of small and thick stone splinters/chunks.
+This first factor explained 28% of the variance.
+Higher factor 1 values reflect higher flaking quantities.
+The second factor covers 24% of the sample variance.
+Factor 2 measures one's ability to carefully flake cores due to high positive loadings on large flake count and lower or neutral loadings on mass flakes/flaked mass, core delta mass, and total cores used.
+The resulting flakes are larger, relatively thinner, and more typically shaped due to high positive loadings on flake factor 1 (size) and flake factor 2 (relative thickness) contrasted with a high negative loading on flake shape factor 3 (basal/tip width).
+Higher factor two values show increased quality flaking performance.
+
+**Figure** shows the covariance between our two flake performance principal components.
+While there is no significant relationship between the two factors, and the two study groups do not show significantly different slopes (p = 0.3), there are never-the-less differences between the two groups.
+Trained individuals are able to increase flake quality while increasing flake quality, whereas untrained individuals increase their flaking quantity at the expense of flake quality.
+
 ## **Do trained, untrained, and expert knappers perform differently?**
+
+Here we compare our flaking outcomes (flake size/shape and flaking performance factors) between the trained and untrained groups to see if training in any way impacted their flaking performance.
+We added a performance comparison (individual flake and core knapping performance metrics) between these two novice groups and our expert knapper to test for differences at different points in the stone flaking skill spectrum.
+
+**Table** summarizes the group level performance tests.
+The results show no significant differences in flaking performance between the trained and untrained groups measured by our two flaking performance factors.
+Trained and untrained individuals overall performed equally as well in terms of flaking quality and quantity.
+Three-way flake size and shape comparisons between our expert knapper and the two novice groups show significant differences in flake shape factor 2 (relative thickness), but with a very low (\<0.01) effect size.
+This difference is driven by the expert's lower overall relative flake thickness.
+Regardless of training, our experimental subjects produced flakes that were on average the same size and shape as those of the expert trainer.
+We did, however, find several significant differences in the three-way comparisons of our individual flaking performance measures.
+The expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and on average had significantly smaller cores (effect size = 0.27) (**Figure**).
+All three of these results show either medium or large effect sizes.
+In all three comparisons, the trained group's data distributions tended towards the expert sample (although they were not significantly different from the untrained group) (**Figures-Core examples too**).
+These results show that core reduction intensity and large flake production track this experiment's greatest differences between expert and novice performance.
 
@@
 
+Here we use the relative order in which each subject flaked their cores to test for changes in flaking performance across the 2hr experiment.
+For these analyses, we calculated the relative percentage for each core relative to the total cores knapped by each subject.
+These relative core use-order percentages were then binned into 20 percent brackets for core-order and group-level comparisons.
+Flaking outcomes were tracked using the two flake performance factors (quality and quantity flaking).
+We added the nodule starting mass to track whether training/practice times impacted raw material selection.
+
+**Table** shows no significant training effects across the two flaking performance measures either as grouped data or between individuals (**Figures)**.
+This result demonstrated that flaking outcomes did not change dramatically across the study interval.
+The one significant main training effect related to core starting mass (with a strong main training effect size = 0.25).
+On average, core starting masses increase as subjects flake more cores, showing that subjects started with smaller nodules first that were smaller and easier to hold/flake.
+As the experiment wore on, they were left with larger and more challenging nodules.
+The small main effect of training condition is driven by higher starting nodule masses in the untrained group at the beginning, half way through, and at the end of the experiment.
+
 ## **Do individual differences in motor skill and psychometric measures predict flaking performance?**
+
+One of the experiment's primary goals was to test if measures of individual variation in motor skill and intelligence/motivation predict success in stone flaking.
+To address this goal, we built two multivariate models examining the relations between our various psychometric measures, subject's motor skill scores, and our two lithic performance factors (quantity flaking and quality flaking).
+These models enabled us to determine which of the psychometric and motor skill factors are better predictors of a participant's flaking performance in the study.
+
+We have already demonstrated that subjects selected progressively larger nodules throughout the experiment.
+It is important now to understand whether nodule variability had any impact on our flaking results.
+Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p \< 0.01, R2 = 0.54) we included nodule mass as a covariate to control for any variance in flaking performance that may be driven by nodule differences.
+Our two motor skill and strength measures (grip strength and Fitt's performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, R2 = 0.41).
+However, these two measures track complementary components of athleticism (strength vs.
+speed/accuracy tradeoffs) and so we decided to include both in the model selection process.
+
+We considered all possible interactions between five individual difference measures, core size, training condition, and the two lithic flaking performance factors (quantity and quality flaking) (wherein each subject provides one data point).
+Each model's continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitt's score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations.
 
@@
 
+The first full model examined variance in the quantity of flaking tracked by our first performance factor explaining increases in the number of cores used, the degree of reduction on each core, and large/relatively thick flake production.
+The full model was fitted with the lm function in R 3.2.3, and we used the Glmulti package's automated model selection algorithm to select the best performing model (lowest AICc score) (see methods for further details on the multimodal selection process).
+The complete model statement is as follows:
+
+*Quantity flaking $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength*
+
+From a candidate pool of 55893 possible multivariate models, the best performing model returned an AICc value of 36 (Average AIC = 52).
+This model comprised the following statement with three main and four interaction effects:
+
+*Quantity flaking $\sim$ Training condition + Highest n-back level  + BEAST score + Grip strength + Fitt’s score$\cdot$Raven’s Progressive Matrix score + Training condition$\cdot$Highest n-back level + Training condition$\cdot$BEAST score + Nodule mass (as control)*
+
+This model explains a statistically significant and substantial proportion of variance in quantity flaking (R2 = 0.84, F (8, 12) = 8.2, p \< 0.01, adj.
+R2 = 0.74).
+A model evaluation comparing training vs.
+test data subsets shows no evidence for overfitting (Difference in R2 between training and test models = 0.02).
+A model residuals normality test shows no significant differences with the normal distribution (p = 0.35) indicating that this relationship (as required) is linear.
+A Breusch-Pagan test showed no evidence for heteroskedasticity (whether variance for all observations in our data set are the same) (BP = 8.2, df = 7, p = 0.3).
+
+\
+**Table** presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average.
+The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable.
+Significant increases in quantity flaking were found for subjects with training (Est.
+= 0.63 [0.15, 1.11]), higher n-back levels (Est.
+= 0.71 [0.29, 1.13]), grip strength (Est.
+= 0.66 [0.36, 0.95]), and BEAST scores (Est.
+= 0.41 [0.04, 0.78]) regardless of starting nodule sizes (**Figure**).
+This suggests that several independent factors tracking the effects of training, social information use, visuo-spatial working memory, and strength improve an individual's ability to flake in greater quantities.
+
+The model produced two significant interactions between training condition and n-back level and BEAST scores.
+Subjects in the trained group with higher n-back levels show lower quantity flaking scores (Est.
+= -1.35 [-1.85, -0.84]) while subjects in the trained group with higher BEAST scores also show lower quantity flaking scores (Est.
+= -0.83 [-1.37, -0.28]) when all the other variables are held constant.
+At first glance, these results appear to contrast with expectations regarding the effects of visuo-spatial working memory and social information use on success in technological tasks.
+However, looking at the matter more holistically it is clear that untrained subjects produce flaking quantities in a different way that relies on visuo-spatial working memory rather than taught information.
+When teaching is provided, being aware of one's surroundings (i.e.
+copying from other novices) can have a negative effect on flaking quantity.
+The interaction between BEAST scores and training shows that social information use acts to level the effects of an individual's propensity to "muscle" through the flaking task (both groups converge on an average quantity flaking score with higher BEAST scores).
+However, the direction of this change is different depending on training with untrained and socially inclined individuals increasing flaking quantities in contrast to trained and socially inclined individuals decreasing flaking quantities.
+
 ### **Model 2: Individual differences and quality flaking**
 
+Our second model examining variance in quality flaking follows the same complete model statement we used for the quantity flaking with six covariates.
+From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of 32 (Average AIC = 41).
+This model comprised the following statement with three main and four interaction effects:
+
+*Quality flaking $\sim$ Highest n-back level  + Fitt’s score + Grip strength + Fitt’s score$\cdot$BEAST score + Grip strength$\cdot$BEAST score + Grip strength$\cdot$Fitt’s score + Training condition$\cdot$Grip strength + Nodule mass (as control)*
+
+This model explains a statistically significant and substantial proportion of variance in careful flaking (R2 = 0.78, F (8, 12) = 5.5, p \< 0.01, adj.
+R2 = 0.64) in the absence of any main training effects.
+A model evaluation comparing training vs.
+test data subsets shows some evidence for overfitting (Difference in R2 between training and test models = 0.17), which is unfortunately unavoidable with our small sample sizes.
+A model residuals normality test shows no significant differences with the normal distribution (p = 0.13) indicating that this relationship is linear.
+A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 3.4, df = 8, p = 0.9).
+
+\
+**Table** presents this model's coefficients and summary outputs following the same data format as **Table**.
+The results show three significant main effects between our measures of motor-skill, strength, visuo-spatial working memory, and quality flaking.
+A one unit increase in Fitt's score (Est.
+= -0.7 [-1.08, -0.31]), grip strength (Est.
+= -0.9 [-1.32, -0.53]), visuo-spatial working memory (n-back level) (Est.
+= -0.3 [-0.58, -0.07]), and nodule starting mass (Est.
+= -0.3 [-0.58, -0.04]) results in decreased flaking quality regardless of training.
+Collectively, these results show that quality flaking comes at a cost in terms of a subject's motor abilities and that quality flaking decreases as subjects are faced with flaking larger and more challenging nodules.
+
+The quality flaking model shows two significant interaction effects involving grip strength, BEAST scores, and training.
+The interaction between grip strength and BEAST scores has a significant positive effect on careful flaking (Est.
+= 0.5 [0.22, 0.86]).
+**Figure** illustrates this interaction whereby a subject's propensity to use social information mitigates against individual differences in grip strength.
+Strong subjects with a low propensity to use social information tend to perform worst in terms of quality flaking.
+This result makes sense if one considers that our quantity flaking model showed how social information equalizes performance levels in trained and untrained subjects. \
+\
+The second interaction involves grip strength and training.
+A one unit increase in this interaction results in a significant positive effect on careful flaking (Est.
+= 1 [0.51, 1.47]), but only in the untrained group.
+**Figure** illustrates this interaction showing how training helps even out flaking performance differences between stronger and weaker individuals.
+Stronger individuals perform worse in terms of quality flaking when they are untrained.
+Trained subjects, regardless of grip strength, perform equally well in terms of quality flaking.
+
 # **Discussion**
+
+we have little basis other than personal experience and/or tradition [@callahan1979; @whittaker1994; @shea2015] and theoretical speculation [@stout2013; @whiten2015] from which to assess what pedagogical techniques are most effective even in WEIRD contexts.
+For example, no study to date has considered how variation in teacher skill [@shea2015] or social relationship to participants might impact learning under different conditions.
+To properly address these questions would require a major research program, including both cross-cultural comparative studies [@barrett2020] (Barrett 2020) and more naturalistic study designs.
+While costly, such research would produce results of broad relevance to anthropologists, biologists, psychologists, and sociologists interested in teaching and learning, well beyond any particular implications for language evolution.
 

### 2021-07-31 manuscript/Testing-the-cognitive-foundations-of-Paleolithic-social-transmission.tex
update the text to the latest google doc version

@@
 al. 2017; Cataldo, Migliano, and Vinicius 2018; Putt, Wijeakumar, and
-Spencer 2019; Pargeter and Shea 2019; Pargeter et al. 2020). This is
-motivated by the expectation that the most salient cognitive demands of
-tool making should occur during learning rather than routine expert
-performance (Stout and Khreisheh 2015) and by interest in the relevance
-of different social learning mechanisms such as imitation (Rein, Nonaka,
-and Bril 2014; Stout et al. 2019), emulation (Tehrani and Riede 2008;
-Wilkins 2018), and language (Ohnuma, Aoki, and Akazawa 1997; Putt,
-Woods, and Franciscus 2014; Morgan et al. 2015; Lombao, Guardiola, and
-Mosquera 2017; Putt et al. 2017; Cataldo, Migliano, and Vinicius 2018)
-to the reproduction of Paleolithic technologies.
+Spencer 2019; Pargeter, Khreisheh, and Stout 2019; Pargeter et al.
+2020). This is motivated by the expectation that the most salient
+cognitive demands of tool making should occur during learning rather
+than routine expert performance (Stout and Khreisheh 2015) and by
+interest in the relevance of different social learning mechanisms such
+as imitation (Rein, Nonaka, and Bril 2014; Stout et al. 2019), emulation
+(Tehrani and Riede 2008; Wilkins 2018), and language (Ohnuma, Aoki, and
+Akazawa 1997; Putt, Woods, and Franciscus 2014; Morgan et al. 2015;
+Lombao, Guardiola, and Mosquera 2017; Putt et al. 2017; Cataldo,
+Migliano, and Vinicius 2018) to the reproduction of Paleolithic
+technologies.
 
@@
 tradeoffs inherent in experimental studies of stone knapping skill
-acquisition (Pargeter and Shea 2019; Stout and Khreisheh 2015).
+acquisition (Pargeter, Khreisheh, and Stout 2019; Stout and Khreisheh
+2015).
 
@@
 relatively rapid, variable across individuals, and predictive of later
-outcomes (Pargeter and Shea 2019; Stout and Khreisheh 2015; Putt,
-Wijeakumar, and Spencer 2019), and thus provides a reasonable
+outcomes (Pargeter, Khreisheh, and Stout 2019; Stout and Khreisheh 2015;
+Putt, Wijeakumar, and Spencer 2019), and thus provides a reasonable
 expectation of generating meaningful data on skill and learning
@@
 be associated with prefrontal (working memory/cognitive control)
-activation and Stout et al.~(2015) found that prefrontal activation
+activation and Stout et al. (2015) found that prefrontal activation
 correlated with success at a strategic judgement (platform selection)
@@
 effort demands of knapping skill acquisition suggest that differences in
-personality (e.g., self-control and ``grit'' (Pargeter and Shea 2019),
-motivation (Stout 2002), and social vs.~individual learning strategies
-(Miu et al. 2020) might also affect learning outcomes. We are again
-unaware of any previous studies that have assessed such effects. In this
-study, we assessed all participants with a battery of tests including
-grip strength, movement speed/accuracy, spatial working memory, fluid
-intelligence, self-control, tendency to use social information, and
-motivation/engagement with the tool making task. We were particularly
-interested in the possibility that these variables might not only impact
-learning generally, but might also have different effects under
-different learning conditions.~
+personality (e.g., self-control and ``grit'' (Pargeter, Khreisheh, and
+Stout 2019), motivation (Stout 2002), and social vs.~individual learning
+strategies (Miu et al. 2020) might also affect learning outcomes. We are
+again unaware of any previous studies that have assessed such effects.
+In this study, we assessed all participants with a battery of tests
+including grip strength, movement speed/accuracy, spatial working
+memory, fluid intelligence, self-control, tendency to use social
+information, and motivation/engagement with the tool making task. We
+were particularly interested in the possibility that these variables
+might not only impact learning generally, but might also have different
+effects under different learning conditions.~
 
@@
 
-\hfill\break
+Toth and Schick (1993) were, however, careful to point out that extinct
+hominid learning strategies and capacities might differ from modern
+experimental participants. Even leaving aside potential species
+differences in social learning (cf. Morgan et al. 2015; Stout et al.
+2019), reliance on explicit verbal instruction varies widely across
+modern human societies (e.g., Boyette and Hewlett 2017). The WEIRD
+(Western, educated, industrialized, rich, democratic (Henrich, Heine,
+and Norenzayan 2010)) teachers and learners typical of knapping
+experiments arguably represent an extreme bias toward such instruction.
+Simply instructing such participants not to speak during an experiment
+(or to demonstrate but not gesture, etc. (Morgan et al. 2015)) is likely
+to underestimate the efficacy of non-verbal teaching and learning in
+cultural contexts where it is more common, let alone in a hypothetical
+pre-linguistic hominid species.
+
+Such concerns are exacerbated in experiments using pre-recorded
+instructional videos or extremely short training periods. Video does not
+allow the interactive teaching that is favored even in formal academic
+knapping classes (e.g., Shea 2015) and is almost certainly typical of
+traditional learning contexts (e.g., Stout 2002). It is not known how
+video presentation affects the efficacy of teaching generally, or the
+relative effectiveness of different forms of instruction. Going further,
+some experiments have manipulated the presence/absence of verbal
+instruction by presenting the same video with and without sound (Putt et
+al. 2017) or the sound track without the video (Cataldo, Migliano, and
+Vinicius 2018). While this provides experimental control, it does not
+allow the instructor to adjust their multi-modal (Levinson and Holler
+2014) communication strategies as they would naturally do, for example
+through pointing and pantomime. To simply remove a communication channel
+without allowing any such adaptation is highly artificial and risks
+generating results that cannot be generalized beyond the specific
+context of the experiment (Yarkoni 2020). Similarly, unnaturally short
+training periods (e.g., 5-15 minutes (Morgan et al. 2015; Lombao,
+Guardiola, and Mosquera 2017)) might misrepresent the relative efficacy
+of different teaching strategies under more realistic conditions (Whiten
+2015; Stout and Khreisheh 2015). Even the longest training times to date
+(Stout and Khreisheh 2015; Pargeter, Khreisheh, and Stout 2019) have not
+produced knapping skills comparable to relevant archaeological examples,
+and were achieved by limiting sample size and using only one teaching
+condition.
+
+For these reasons, we sought to explore a middle path between
+experimental expedience and realism by limiting our experiment to two
+relatively naturalistic learning conditions and a moderate learning
+period of two hours. As in previous experiments (Stout et al. 2011;
+Hecht, Gutman, Khreisheh, et al. 2015; Pargeter, Khreisheh, and Stout
+2019) the first condition was unrestricted, interactive instruction in
+small groups, essentially reproducing the ``natural'' teaching/learning
+context familiar (cf. Shea 2015) to our WEIRD instructor and student
+participants. The second condition allowed observation only, with the
+experimenter visible making flakes but not interacting in any way with
+learners. This absence of teaching is again a familiar social context
+for our participants and did not require any novel behaviors from the
+instructor. It matches the ``imitation/emulation'' condition of Morgan
+et al. (2015) although we make no assumptions regarding learning
+mechanisms. We did not include a ``reverse engineering'' or ``end-state
+emulation'' condition in which only finished products were visible. This
+has been advocated as an important baseline or control condition (Whiten
+2015) to distinguish observational from individual learning, but is not
+likely to model any typical Paleolithic learning context nor to stand as
+an adequate proxy for the cognition of hominid species with different
+social learning capacities. There is no reason to assume neurocognitive
+and behavioral processes of reverse-engineering problem solving in
+modern humans (e.g., Allen, Smith, and Tenenbaum 2020) approximate the
+social learning processes of hominids with more ape-like action
+observation/imitation capacities (Hecht, Gutman, et al. 2013; Hecht,
+Murphy, et al. 2013; Stout et al. 2019).
+
+We selected a two-hour learning period for both pragmatic and
+theoretical reasons. Pargeter et al. (Pargeter, Khreisheh, and Stout
+2019) found that even \textasciitilde90 hours of fully interactive
+instruction and practice was insufficient to achieve handaxe-making
+skills comparable to the later Acheulean site of Boxgrove
+(García-Medrano et al. 2019; Stout et al. 2014), and estimated actual
+time to mastery as ranging from 121 to 441 hours for different
+participants. However, they observed the greatest, fastest, and most
+individually variable skill increases during the first 20 hours of
+practice. In addition, initial performance was moderately correlated
+with later achievement. This suggests that studying early-stage learning
+may be a pragmatic alternative, especially for research investigating
+individual differences in aptitude. Studies of simple flake production
+similarly document large initial variation (Stout and Khreisheh 2015)
+and rapid early progress (Stout and Khreisheh 2015; Putt, Wijeakumar,
+and Spencer 2019; Stout and Semaw 2006). We designed the current study
+to test the utility of studying learning and variation during the first
+two hours of simple flaking instruction/practice, in hopes of finding a
+viable compromise between experimental realism and cost
 
@@
 
+Lithic raw materials vary in size, shape, and fracture mechanical
+properties that affect the difficulty of achieving different knapping
+goals (Eren et al. 2014). Unfortunately, it can be difficult and/or
+expensive to procure authentic raw materials. Experimental studies of
+knapping skill have often used proxy materials such as flint (Morgan et
+al. 2015; Nonaka, Bril, and Rein 2010; Cataldo, Migliano, and Vinicius
+2018), limestone (Stout and Semaw 2006), porcelain (Khreisheh, Davies,
+and Bradley 2013), or heat-treated chert (Putt, Woods, and Franciscus
+2014; Putt et al. 2017; Putt, Wijeakumar, and Spencer 2019)to model
+Oldowan and early Acheulean technologies executed in other materials. As
+well as being more readily available, these proxies are generally easier
+to knap. This has the benefit of reducing required practice time, but it
+is unclear how it might affect learning demands more generally or the
+efficacy of different learning conditions/strategies specifically.
+
+To address this, some studies have attempted to more closely match
+experimental and archaeological raw material types (Stout et al. 2011;
+Duke and Pargeter 2015; Pargeter, Khreisheh, and Stout 2019). However,
+raw materials vary across individual clasts within as well as between
+types. This has led to interest in standardizing experimental core
+morphology (Nonaka, Bril, and Rein 2010) and composition, even if this
+means using artificial materials such as porcelain (Khreisheh, Davies,
+and Bradley 2013), brick (Geribàs, Mosquera, and Vergès 2010; Lombao,
+Guardiola, and Mosquera 2017), or foam blocks (Schillinger, Mesoudi, and
+Lycett 2014). Such manipulations enhance experimental control and
+internal validity (Eren et al. 2016) at the expense of external
+generalizability to actual archaeological conditions. Specifically, they
+allow more robust results from smaller samples but eliminate a core
+element of real-world knapping skill: the ability to produce consistent
+results from variable materials (Pelegrin 1990; Stout 2013). For
+example, Pargeter et al. (2020) found that predicting specific flaking
+outcomes on actual handaxe preforms was both more difficult and less
+technologically important than expected from previous work with
+standardized, frustum-shaped cores (Nonaka, Bril, and Rein 2010). The
+alternative to control is to incorporate raw material size, shape, and
+composition as experimental variables (e.g., Stout et al. 2019). This
+allows consideration of raw material selection and response to variation
+as aspects of skill but correspondingly increases the sample sizes
+required to identify patterning. In considering these issues, we again
+chose to explore a middle path between pragmatism and realism by
+employing commercially purchased basalt similar to that known from East
+African Oldowan sites, allowing clast size and shape to vary within set
+limits, and selecting the particular clasts provided to each participant
+to approximate the same distribution.
+
+\hfill\break
+
 \hypertarget{materials-and-methods}{%
@@
 
+This research was approved by the Emory Institutional Review Board
+(IRB00113024). All participants provided written informed consent and
+completed a video release form
+(\url{https://databrary.org/support/irb/release-template.html}).
+
 \hypertarget{participants}{%
 \subsection{\texorpdfstring{\textbf{Participants}}{Participants}}\label{participants}}
+
+Twenty-four adult participants with no prior stone knapping experience
+were recruited from the Emory community using paper fliers and e-mail
+listserv advertisements. We were unable to replace one participant who
+failed to attend their scheduled session, resulting in a total sample of
+23. Eleven participants (6 female, 5 male) completed the Untaught
+condition and 12 (8 female, 4 male) completed the Taught condition.
 
@@
 
+Participants were asked to visit the Paleolithic Technology Lab at Emory
+University to complete one three-hour session. Participants were
+scheduled to attend in six groups of four, however one of these groups
+had only three participants due to a no-show on the day of the
+experiment. Each visit began with the collection of individual
+differences measures, which took approximately one hour. After that,
+participants undertook 105 minutes (two hours minus a 15-minute break
+after 1 hour) of stone tool making practice. This session was
+video-recorded, and all lithic products were collected. After the tool
+making task, participants completed an ``exit questionnaire'' comprising
+the Intrinsic Motivation Inventory (see below).
+
+Participants were compensated for their time with a \$30 gift card. They
+also had the opportunity to earn a performance bonus of \$5, \$10, \$15
+or \$20 on the gift card. They were told that this bonus would depend on
+``how well they did'' on the last core of their practice session. The
+actual performance measure was not specified, but in order to allow on
+the spot payment a simple measure of the percentage of starting weight
+removed from the final core was used such that: \textgreater{} 30\%
+earned \$5, \textgreater{} 40\% earned \$10, \textgreater{} 50\% earned
+\$15, \textgreater{} 75\% earned \$20.
+
+\hfill\break
+
 \hypertarget{individual-difference-measures}{%
@@
 
+We used five individual difference measures for this study:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi})}
+\item
+  Grip strength was measured in kilograms using an electronic hand
+  dynamometer (Camry EH101). Strength was measured twice and the higher
+  value recorded. Grip strength is a simple measure that is well
+  correlated with overall muscular strength (Wind et al. 2010) and a
+  range of other health and fitness measures (Sasaki et al. 2007). It is
+  hypothesized to be relevant to generating kinetic energy for fracture
+  initiation (Nonaka, Bril, and Rein 2010) as well as control and
+  support of the hammerstone (Williams-Hatala et al. 2018) and core
+  (Faisal et al. 2010; Alastair J. M. Key and Dunmore 2015).
+\item
+  Motor accuracy was assessed using a ``Fitts Law'' reciprocal tapping
+  task. Fitts Law describes the trade-off between speed and accuracy in
+  human movement, classically measured by tapping back and forth between
+  two targets of varying size and spacing (Fitts 1954). Archaeologists
+  have proposed (Stout 2002; Pargeter et al. 2020) that management of
+  this trade-off is critical to the accurate application of appropriate
+  force seen in skilled knapping (Nonaka, Bril, and Rein 2010; Roux,
+  Bril, and Dietrich 1995). We implemented this test on a Surface Pro
+  tablet running free software (FittsStudy Version 4.2.8, default
+  settings) developed by the Accessible Computing Experiences lab (Jacob
+  O. Wobbrock, director) at the University of Washington
+  (\url{depts.washington.edu/acelab/proj/fittsstudy/index.html}).
+  Participants use a touchscreen pen to tap between ribbons on the
+  screen, with average movement time as the performance metric.
+\item
+  Visuospatial working memory is the capacity to ``hold in mind,'' which
+  researchers have hypothesized to be important in stone toolmaking
+  performance (Coolidge and Wynn 2005). It also might support a learning
+  process known as `chunking,' in which multiple items or operations are
+  combined into summary chunks stored in long term memory, that is
+  thought to be important in the acquisition of knapping and other
+  skills (Pargeter, Khreisheh, and Stout 2019). We measured visuospatial
+  working memory using a free n-back task
+  (\href{https://wmp.education.uci.edu/software/}{wmp.education.uci.edu/software/})
+  developed by the Working Memory and Plasticity Laboratory at the
+  University of California, Irvine (Susanne Jaeggi, PI) and implemented
+  in E-Prime software on a desktop computer. In this task, participants
+  are asked to remember the position of blue squares presented
+  sequentially on the screen and touch a key when the current position
+  matches that 1, 2, 3\ldots n iterations back. Progression to blocks
+  with increasing values of n is contingent on exceeding a threshold
+  success rate. Performance was measured as the highest n achieved.
+\item
+  Fluid intelligence (Cattell 1963) refers to the capacity to engage in
+  abstract reasoning and problem solving in a way that is minimally
+  dependent on prior experience. It complements ``crystallized
+  intelligence'' (the ability to apply learned procedures and knowledge)
+  as one of the two factors (gf, gc) comprising so-called ``general
+  intelligence'' (g). Fluid intelligence is closely related to the
+  executive control of attention and manipulation of information held in
+  working memory (Engle 2018)(Engle 2018). It is hypothesized to support
+  technological innovation (Coolidge and Wynn 2005) and/or the
+  intentional learning of new skills (Unsworth and Engle 2005; Stout and
+  Khreisheh 2015). We measured fluid intelligence using the short
+  version (Bilker et al. 2012) of the classic Raven Progressive Matrices
+  task, which requires participants to complete increasingly difficult
+  pattern matching questions.~~
+\item
+  The use of social information for learning and decision making varies
+  across individuals and societies (Molleman, Kurvers, and van den Bos
+  2019). Such variation is a key topic for understanding social learning
+  and cultural evolutionary processes (Kendal et al. 2018; Heyes 2018;
+  Miu et al. 2020) and represents a potential confound for assessing
+  experimental effects of different social learning conditions. We
+  measured participants' tendency to rely on social information
+  vs.~their own insights using the Berlin Estimate AdjuStment Task
+  (BEAST) developed by Molleman et al. (2019). In this task,
+  participants are present with large arrays of items on a screen and
+  asked to estimate the number present. They are then provided with
+  another person's estimate and allowed to provide a second estimate.
+  The participants' average adjustment between first and second
+  estimates provides a measure of their propensity to rely on social
+  information.
+\end{enumerate}
+
+\hfill\break
+
 \hypertarget{stone-tool-making}{%
@@
 
+After individual difference testing, participants engaged in a 2-hour
+stone tool making session, with a 15-minute break after 1 hour.
+Participants were instructed not to seek out additional training or
+information on stone tool making (i.e., via the internet) during these
+breaks. Each group of participants was randomly assigned to one of two
+experimental conditions: no teaching or teaching. In both conditions,
+participants were first given an opportunity to inspect and handle
+examples (\textbf{Figure}) of the kind of stone tools (flakes) they are
+being asked to produce. They were told that their objective was to
+produce as many flakes as possible from the materials provided. This
+meant that even the untaught condition included some minimal instruction
+(being told the objective) , however this was considered to be
+unavoidable without creating a much more elaborate and naturalistic
+context in which participants would develop their own technological
+goals. Such a design would also be expected to increase behavioral
+variability, demanding correspondingly larger samples of participants to
+identify patterns and making direct comparisons with the taught
+condition.
+
 \hypertarget{raw-materials}{%
@@
 
+Each participant was provided with 9 cores for use over the 2-hour
+experiment. These cores were produced from larger chunks of a
+fine-grained basalt purchased from neolithics.com by fracturing them
+with a sledgehammer. This produced irregular, angular chunks for use in
+the experiment, weighing between 459g - 1876g (mean = 975g). All cores
+were weighed, measured (Length, Width, Thickness), and painted white so
+that new fracture surfaces could be discriminated from those created
+during production. Cores were sorted by shape and weight and then
+distributed evenly to each participant. As a result, there were no
+significant difference across participants in the mean weight (ANOVA, df
+= 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04,
+df1=22, df2 = 184, p = 0.4) or shape (Length x Width/Thickness: ANOVA,
+df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p =
+0.9) of cores provided. This was also true comparing the two
+experimental conditions (Taught vs.~Untaught mean weight = 1001g
+vs.~956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4;
+mean shape = 221.43 vs.~221.45, t = -0.003, df = 205, p = 0.9, Levene's
+Test F = 3.8, p = 0.05). Participants were, however, allowed to choose
+which cores to work on so that differences in the weight and shape of
+cores actually used across participants and conditions could still
+emerge as a result of selection bias.
+
+Sixty pounds of 3-to-5 inch basalt ``Mexican Beach Pebbles'' were
+purchased from a landscaping supply company for use as hammerstones in
+the experiment. Of these, 90 were selected as suitable for use. These
+weighed between 213g-1360g (mean = 425) and varied in elongation (L/W =
+1.01 to 2.65) and relative thickness (LxW/T = 90.48 to 283.67).
+Forty-five stones were placed in the middle of the knapping area (Figure
+2) for participants to freely choose from during the experiment. Broken
+hammerstones were replaced from the reserve to maintain a consistent
+number and range of choices. Each hammerstone was numbered and
+participants' choices were recorded along with the number of the core(s)
+being worked on with a particular hammerstone.\\
+
 \hypertarget{experimental-conditions}{%
@@
 
-\hypertarget{lithic-analysis}{%
-\subsection{\texorpdfstring{\textbf{Lithic
-Analysis}}{Lithic Analysis}}\label{lithic-analysis}}
+In both conditions, three researchers were present to record activities
+and collect materials. Participants were seated in a circle (Shea 2015)
+and experiments were video recorded using two cameras (Figure 1).
+Participants were free to select hammerstones from the common pile and
+to work on any or all of their nine assigned cores in any order they
+preferred. However, each core and all associated debitage were collected
+before participants were allowed to start working on a new core, so it
+was not possible to partially work and then return to a particular core
+later. The order of cores used and associated hammerstones were recorded
+for each participant during the experiment.
+
+In the untaught condition, a researcher (DS) sat with the participants
+and made stone tools but remained silent and made no effort to
+facilitate learning (e.g., through gesture, modified performance, facial
+expression, attention direction, or verbal instruction). Over the 2-hour
+period, the researcher completely reduced four cores (one every
+\textasciitilde30 minutes). Participants were not restricted from
+talking to each other, as this would create an unnatural and potentially
+stressful social context that might affect learning. Participants were
+asked to avoid any form of communication about the tool making task
+specifically, and they complied with this request. Participants in this
+condition thus had the opportunity to observe tool making by an expert
+and/or by other learners, should they choose to do so, but received no
+intentional instruction.
+
+In the Taught condition, there were no restrictions on participant
+interaction and the researcher engaged in direct active teaching (Kline
+2015) of tool-making techniques through verbal instruction,
+demonstration, gesture, and shaping of behavior. The instructor has a
+moderate level of experience teaching basic knapping skills to students
+in undergraduate archaeology classes and to participants in previous
+knapping research (e.g., Stout et al. 2011). The pedagogical strategy
+employed was based on the instructor's own learning experiences and
+theoretical interpretations (e.g., Pargeter et al. 2020), and focused on
+coaching participants in effective body postures, movement patterns, and
+grips as well as the assessment of viable core morphology.~ ~
+
+\textbf{Lithic Analysis}
+
+All finished cores were weighed and measured (L, W, T). Delta weight was
+calculated as (Start weight-End weight)/Start weight. All detached
+pieces (DPs) were collected and weighed. We did not sort DPs into types
+(e.g., whole flakes, fragments) as this would have greatly increased
+processing time and it is not clear that such distinctions add relevant
+information regarding utility/desirability beyond that supplied by
+metrics (Stout et al. 2019). All DPs larger than 40mm in maximum
+dimension were photographed and measured. It is conventional in Early
+Stone Age lithic analysis to employ a 20 mm cut-off. We selected a
+higher threshold for both pragmatic (analysis time) and theoretical
+reasons. Flake use experiments have shown that flakes weighing less than
+5--10 g or with a surface area below 7--10 cm2 (Prasciunas 2007) or with
+a maximum dimension \textless50-60 mm (Alastair J. M. Key and Lycett
+2014) become markedly inefficient for basic cutting tasks. Similarly,
+data from Oldowan replication experiments (Stout et al. 2019) show that
+the utility index (flake cutting edge/flake mass1/3) * (1 - exp{[}-0.31
+* (flake maximum dimension -- 1.81){]}) developed by Morgan et
+al.~(2015) falls off rapidly below 40mm maximum dimension ( Mean Utility
+\textless{} 40mm = 0.508; \textgreater=40mm = 0.946; t= 11.99, df = 707,
+p \textless{} 0.000). By including weight in our cut-off criteria we
+also avoid skewing the flake shape distribution by selectively retaining
+long, thin pieces (i.e., MD \textgreater{} 40, weight \textless{} 5g)
+while discarding rounder pieces of similar (or greater) weight and area.
+
+For measurement, DP length was defined as the longest axis and width as
+the maximum dimension orthogonal to length. Thickness was defined as the
+maximum dimension orthogonal to the plane formed by L and W and was
+measured using calipers. L, W, and plan-view area measurements were
+taken from photographs captured using a Canon Rebel T3i fitted with a 60
+mm macro lens and attached to a photographic stand with adjustable upper
+and lower light fittings. The camera was positioned directly above the
+flakes and kept at a constant height. DPs were positioned irrespective
+of any technological features so that the longest axis was vertical, and
+the wider end was placed toward the bottom of the photograph.
+
+Photographs were post-processed using Equalight software to adjust for
+lens and lighting falloff that result from bending light through a lens
+and its aperture which can affect measurements taken from photographs.
+Each image was shot with a scale that was then used to rectify the
+photograph's pixel scale to a real-world measurement scale in Adobe
+Photoshop. Images were converted to binary black and white format and
+silhouettes of the tools were extracted in Adobe Photoshop. We then used
+a custom ImageJ (Rueden et al. 2017) script (Pargeter, Khreisheh, and
+Stout 2019) to measure DP length and take nine width measurements at
+10\% increments of length starting at the base of each DP. We used the
+built-in ImageJ tool to measure DP area. A ``Proportion Larger DPs'' was
+calculated per core as the combined weight of all DPs \textgreater40mm
+in maximum dimension and 5g in weight divided by the weight of all DPs.
+Higher values show cores with proportionally more large DPs.
 
@@
 
+To evaluate the association between psychometric, motor-skill, and
+training measures and technological outcomes, we adopted an
+information-theoretic approach (Burnham and Anderson 2002).
+Information-theoretic approaches provide methods for model selection
+using all possible combinations of variables while avoiding problems
+associated with significance-threshold stepwise selection. We used the
+corrected Akaike information criterion (AICc) to rate each possible
+combination of predictors on the balance between goodness of fit
+(likelihood of the data given the model) and parsimony (number of
+parameters). The AICc consists of the log likelihood (i.e., how well
+does the model fit the data?) and a penalty term for the number of
+parameters that must be estimated in the model (i.e., how parsimonious
+is the model?), with a correction for small sample sizes (AICc converges
+to the standard AIC at large samples). A lower AICc indicates a more
+generalizable model and we used it to compare and rank various possible
+models. Each analysis begins with a full model that includes all
+predictors of interest. All possible combinations of predictors are then
+fit, and the resulting models are ranked and weighted based on their
+AICc. The ``best'' model is chosen because it has the lowest AICc score.
+
+Continuous predictors were centered such that zero represents the sample
+average, and units are standard deviations. The full model was fitted
+with the lm function in R 3.2.3, and the glmulti package
+(\textbf{Bartoń?}) was used for multi-modal selection and model
+comparison.
+
+\hfill\break
+
 \hypertarget{results}{%
 \section{\texorpdfstring{\textbf{Results}}{Results}}\label{results}}
+
+Following a recent protocol to enhance the reproducibility and data
+transparency of archaeological research (Marwick 2017), detailed results
+of all analyses and assessments of the data structure are available in
+our paper's supplementary materials and through Github
+(\url{https://github.com/Raylc/PaST-pilot}). Here we limit discussion to
+the major ﬁndings regarding flaking performance and individual
+differences. The purpose of this section is to determine whether
+training impacted subject flaking performance and if any of the
+subject's individual psychometric and motor-skill aptitudes predict
+flaking performance.
 
@@
 
+The following two sections outline factor analyses designed to summarize
+our main study metrics tracking individual variation in flake sizes and
+shapes and lithic performance measures.
+
 \hypertarget{flake-size-and-shape}{%
@@
 
+To better understand the relationship between flake shape and
+training/individual variation, we entered our nine flake linear plan
+measurements along with maximum flake length and thickness into a
+principal component analysis (PCA) from which summary coordinates were
+extracted. Bartlett's Test of Sphericity was significant (\(\chi ^2\)
+(10) =4480, p \textless{} .01) indicating that the set of variables are
+adequately related for factor analysis.
+
+The analysis yielded three factors explaining a total of 90\% of the
+variance for the entire 11 measurement variable set \textbf{(Table}).
+Factor 1 tracks flake size with higher scores indicating larger flakes
+since all 11 measures load positively on this factor. Factor 2's
+loadings track the increasing relationship between thickness, length,
+and flake width. As factor 2 scores increase, flakes get thicker,
+longer, and narrow, resembling irregular splinters. Factor 3 tracks the
+relationship between flake proximal and distal width relative to
+thickness. As factor 3 scores go up, flakes get thinner and narrower at
+the distal ends and wider at the base. Factor 3 therefore tracks flakes
+with a typical shape having a thin cross-section, wider base, and
+narrower tip. We used these three flake shape coordinates to approximate
+flake size and shape in the project's flake performance factor analysis.
+
 \hypertarget{lithic-flaking-performance-measures}{%
@@
 measures}}{Lithic flaking performance measures}}\label{lithic-flaking-performance-measures}}
+
+To better understand the relationship between our various lithic
+performance measurements and to reduce these data dimensionality, we
+conducted a second principal component analysis examining the study's
+seven lithic performance measures (mass of flakes relative to flaked
+core mass, count of large flakes {[}\textgreater40mm and 5g{]}, core
+delta mass, three flake shape factors, and the total number of cores
+used). All of these measures except the total cores flaked were
+summarized for each core. The Bartlett's Test of Sphericity was
+significant (\(\chi ^2\) (6) =3950, p \textless{} .01) indicating that
+the set of variables are at least adequately related for factor
+analysis.
+
+The analysis yielded two factors explaining a total of 52.3\% of the
+variance for the entire set of variables. Factor 1 tracks flaking
+quantity due to high positive loadings on mass flakes/flaked mass, core
+delta mass, total cores used, and flake shape factor 2. Negative
+loadings on flake factor 1 (flake size) and 3 (basal/tip width shape)
+suggest that flaking quantity comes at the expense of producing lots of
+small and thick stone splinters/chunks. This first factor explained 28\%
+of the variance. Higher factor 1 values reflect higher flaking
+quantities. The second factor covers 24\% of the sample variance. Factor
+2 measures one's ability to carefully flake cores due to high positive
+loadings on large flake count and lower or neutral loadings on mass
+flakes/flaked mass, core delta mass, and total cores used. The resulting
+flakes are larger, relatively thinner, and more typically shaped due to
+high positive loadings on flake factor 1 (size) and flake factor 2
+(relative thickness) contrasted with a high negative loading on flake
+shape factor 3 (basal/tip width). Higher factor two values show
+increased quality flaking performance.
+
+\textbf{Figure} shows the covariance between our two flake performance
+principal components. While there is no significant relationship between
+the two factors, and the two study groups do not show significantly
+different slopes (p = 0.3), there are never-the-less differences between
+the two groups. Trained individuals are able to increase flake quality
+while increasing flake quality, whereas untrained individuals increase
+their flaking quantity at the expense of flake quality.
 
@@
 
+Here we compare our flaking outcomes (flake size/shape and flaking
+performance factors) between the trained and untrained groups to see if
+training in any way impacted their flaking performance. We added a
+performance comparison (individual flake and core knapping performance
+metrics) between these two novice groups and our expert knapper to test
+for differences at different points in the stone flaking skill
+spectrum.~
+
+\textbf{Table} summarizes the group level performance tests. The results
+show no significant differences in flaking performance between the
+trained and untrained groups measured by our two flaking performance
+factors. Trained and untrained individuals overall performed equally as
+well in terms of flaking quality and quantity. Three-way flake size and
+shape comparisons between our expert knapper and the two novice groups
+show significant differences in flake shape factor 2 (relative
+thickness), but with a very low (\textless0.01) effect size. This
+difference is driven by the expert's lower overall relative flake
+thickness. Regardless of training, our experimental subjects produced
+flakes that were on average the same size and shape as those of the
+expert trainer. We did, however, find several significant differences in
+the three-way comparisons of our individual flaking performance
+measures. The expert knapper made significantly more large flakes
+(effect size = 0.14), had a significantly higher core delta mass signal
+than either of the novice groups (effect size = 0.26), and on average
+had significantly smaller cores (effect size = 0.27) (\textbf{Figure}).
+All three of these results show either medium or large effect sizes. In
+all three comparisons, the trained group's data distributions tended
+towards the expert sample (although they were not significantly
+different from the untrained group) (\textbf{Figures-Core examples
+too}). These results show that core reduction intensity and large flake
+production track this experiment's greatest differences between expert
+and novice performance.
+
 \hypertarget{does-trainingpractice-time-impact-flaking-performance}{%
@@
 
+Here we use the relative order in which each subject flaked their cores
+to test for changes in flaking performance across the 2hr experiment.
+For these analyses, we calculated the relative percentage for each core
+relative to the total cores knapped by each subject. These relative core
+use-order percentages were then binned into 20 percent brackets for
+core-order and group-level comparisons. Flaking outcomes were tracked
+using the two flake performance factors (quality and quantity flaking).
+We added the nodule starting mass to track whether training/practice
+times impacted raw material selection.
+
+\textbf{Table} shows no significant training effects across the two
+flaking performance measures either as grouped data or between
+individuals (\textbf{Figures)}. This result demonstrated that flaking
+outcomes did not change dramatically across the study interval. The one
+significant main training effect related to core starting mass (with a
+strong main training effect size = 0.25). On average, core starting
+masses increase as subjects flake more cores, showing that subjects
+started with smaller nodules first that were smaller and easier to
+hold/flake. As the experiment wore on, they were left with larger and
+more challenging nodules. The small main effect of training condition is
+driven by higher starting nodule masses in the untrained group at the
+beginning, half way through, and at the end of the experiment.~
+
 \hypertarget{do-individual-differences-in-motor-skill-and-psychometric-measures-predict-flaking-performance}{%
@@
 
+One of the experiment's primary goals was to test if measures of
+individual variation in motor skill and intelligence/motivation predict
+success in stone flaking. To address this goal, we built two
+multivariate models examining the relations between our various
+psychometric measures, subject's motor skill scores, and our two lithic
+performance factors (quantity flaking and quality flaking). These models
+enabled us to determine which of the psychometric and motor skill
+factors are better predictors of a participant's flaking performance in
+the study.
+
+We have already demonstrated that subjects selected progressively larger
+nodules throughout the experiment. It is important now to understand
+whether nodule variability had any impact on our flaking results.
+Because starting nodule size (mass) and shape were strongly correlated
+(F {[}1,157{]} = 186, p \textless{} 0.01, R2 = 0.54) we included nodule
+mass as a covariate to control for any variance in flaking performance
+that may be driven by nodule differences. Our two motor skill and
+strength measures (grip strength and Fitt's performance scores) are also
+strongly correlated (F {[}1,19{]} = 15, p \textless{} 0.01, R2 = 0.41).
+However, these two measures track complementary components of
+athleticism (strength vs.~speed/accuracy tradeoffs) and so we decided to
+include both in the model selection process.
+
+~
+
+We considered all possible interactions between five individual
+difference measures, core size, training condition, and the two lithic
+flaking performance factors (quantity and quality flaking) (wherein each
+subject provides one data point). Each model's continuous predictors
+(highest n-back level, Raven's Progressive Matrix score, BEAST score,
+starting nodule mass, Fitt's score, and grip strength) were centered
+such that zero represents the sample average, and units are standard
+deviations.~
+
 \hypertarget{model-1-individual-differences-and-quantity-flaking}{%
@@
 
+The first full model examined variance in the quantity of flaking
+tracked by our first performance factor explaining increases in the
+number of cores used, the degree of reduction on each core, and
+large/relatively thick flake production. The full model was fitted with
+the lm function in R 3.2.3, and we used the Glmulti package's automated
+model selection algorithm to select the best performing model (lowest
+AICc score) (see methods for further details on the multimodal selection
+process). The complete model statement is as follows:
+
+\emph{Quantity flaking \(\sim\) Training condition + Highest n-back
+level + Raven's Progressive Matrix score + BEAST score + Fitt's score +
+Grip strength}
+
+From a candidate pool of 55893 possible multivariate models, the best
+performing model returned an AICc value of 36 (Average AIC = 52). This
+model comprised the following statement with three main and four
+interaction effects:
+
+\emph{Quantity flaking \(\sim\) Training condition + Highest n-back
+level + BEAST score + Grip strength + Fitt's score\(\cdot\)Raven's
+Progressive Matrix score + Training condition\(\cdot\)Highest n-back
+level + Training condition\(\cdot\)BEAST score + Nodule mass (as
+control)}
+
+This model explains a statistically significant and substantial
+proportion of variance in quantity flaking (R2 = 0.84, F (8, 12) = 8.2,
+p \textless{} 0.01, adj. R2 = 0.74). A model evaluation comparing
+training vs.~test data subsets shows no evidence for overfitting
+(Difference in R2 between training and test models = 0.02). A model
+residuals normality test shows no significant differences with the
+normal distribution (p = 0.35) indicating that this relationship (as
+required) is linear. A Breusch-Pagan test showed no evidence for
+heteroskedasticity (whether variance for all observations in our data
+set are the same) (BP = 8.2, df = 7, p = 0.3).
+
+\hfill\break
+\textbf{Table} presents this model's coefficients and summary outputs,
+wherein baseline refers to the untrained condition with all continuous
+predictors at the sample average. The parameter estimates for the
+continuous predictors reflect the expected change in utility for 1
+standard deviation change in the predictor variable. Significant
+increases in quantity flaking were found for subjects with training
+(Est. = 0.63 {[}0.15, 1.11{]}), higher n-back levels (Est. = 0.71
+{[}0.29, 1.13{]}), grip strength (Est. = 0.66 {[}0.36, 0.95{]}), and
+BEAST scores (Est. = 0.41 {[}0.04, 0.78{]}) regardless of starting
+nodule sizes (\textbf{Figure}). This suggests that several independent
+factors tracking the effects of training, social information use,
+visuo-spatial working memory, and strength improve an individual's
+ability to flake in greater quantities.
+
+The model produced two significant interactions between training
+condition and n-back level and BEAST scores. Subjects in the trained
+group with higher n-back levels show lower quantity flaking scores (Est.
+= -1.35 {[}-1.85, -0.84{]}) while subjects in the trained group with
+higher BEAST scores also show lower quantity flaking scores (Est. =
+-0.83 {[}-1.37, -0.28{]}) when all the other variables are held
+constant. At first glance, these results appear to contrast with
+expectations regarding the effects of visuo-spatial working memory and
+social information use on success in technological tasks. However,
+looking at the matter more holistically it is clear that untrained
+subjects produce flaking quantities in a different way that relies on
+visuo-spatial working memory rather than taught information. When
+teaching is provided, being aware of one's surroundings (i.e.~copying
+from other novices) can have a negative effect on flaking quantity. The
+interaction between BEAST scores and training shows that social
+information use acts to level the effects of an individual's propensity
+to ``muscle'' through the flaking task (both groups converge on an
+average quantity flaking score with higher BEAST scores). However, the
+direction of this change is different depending on training with
+untrained and socially inclined individuals increasing flaking
+quantities in contrast to trained and socially inclined individuals
+decreasing flaking quantities.
+
 \hypertarget{model-2-individual-differences-and-quality-flaking}{%
@@
 
+Our second model examining variance in quality flaking follows the same
+complete model statement we used for the quantity flaking with six
+covariates. From the same candidate pool size of 55893 possible
+multivariate models, the best performing model returned an AICc value of
+32 (Average AIC = 41). This model comprised the following statement with
+three main and four interaction effects:
+
+\emph{Quality flaking \(\sim\) Highest n-back level + Fitt's score +
+Grip strength + Fitt's score\(\cdot\)BEAST score + Grip
+strength\(\cdot\)BEAST score + Grip strength\(\cdot\)Fitt's score +
+Training condition\(\cdot\)Grip strength + Nodule mass (as control)}
+
+This model explains a statistically significant and substantial
+proportion of variance in careful flaking (R2 = 0.78, F (8, 12) = 5.5, p
+\textless{} 0.01, adj. R2 = 0.64) in the absence of any main training
+effects. A model evaluation comparing training vs.~test data subsets
+shows some evidence for overfitting (Difference in R2 between training
+and test models = 0.17), which is unfortunately unavoidable with our
+small sample sizes. A model residuals normality test shows no
+significant differences with the normal distribution (p = 0.13)
+indicating that this relationship is linear. A Breusch-Pagan test showed
+no evidence for heteroskedasticity (BP = 3.4, df = 8, p = 0.9).
+
+\hfill\break
+\textbf{Table} presents this model's coefficients and summary outputs
+following the same data format as \textbf{Table}. The results show three
+significant main effects between our measures of motor-skill, strength,
+visuo-spatial working memory, and quality flaking. A one unit increase
+in Fitt's score (Est. = -0.7 {[}-1.08, -0.31{]}), grip strength (Est. =
+-0.9 {[}-1.32, -0.53{]}), visuo-spatial working memory (n-back level)
+(Est. = -0.3 {[}-0.58, -0.07{]}), and nodule starting mass (Est. = -0.3
+{[}-0.58, -0.04{]}) results in decreased flaking quality regardless of
+training. Collectively, these results show that quality flaking comes at
+a cost in terms of a subject's motor abilities and that quality flaking
+decreases as subjects are faced with flaking larger and more challenging
+nodules.
+
+The quality flaking model shows two significant interaction effects
+involving grip strength, BEAST scores, and training. The interaction
+between grip strength and BEAST scores has a significant positive effect
+on careful flaking (Est. = 0.5 {[}0.22, 0.86{]}). \textbf{Figure}
+illustrates this interaction whereby a subject's propensity to use
+social information mitigates against individual differences in grip
+strength. Strong subjects with a low propensity to use social
+information tend to perform worst in terms of quality flaking. This
+result makes sense if one considers that our quantity flaking model
+showed how social information equalizes performance levels in trained
+and untrained subjects.~\\
+~\\
+The second interaction involves grip strength and training. A one unit
+increase in this interaction results in a significant positive effect on
+careful flaking (Est. = 1 {[}0.51, 1.47{]}), but only in the untrained
+group. \textbf{Figure} illustrates this interaction showing how training
+helps even out flaking performance differences between stronger and
+weaker individuals. Stronger individuals perform worse in terms of
+quality flaking when they are untrained. Trained subjects, regardless of
+grip strength, perform equally well in terms of quality flaking.
+
 \hypertarget{discussion}{%
 \section{\texorpdfstring{\textbf{Discussion}}{Discussion}}\label{discussion}}
+
+we have little basis other than personal experience and/or tradition
+(Callahan 1979; Whittaker 1994; Shea 2015) and theoretical speculation
+(Stout 2013; Whiten 2015) from which to assess what pedagogical
+techniques are most effective even in WEIRD contexts. For example, no
+study to date has considered how variation in teacher skill (Shea 2015)
+or social relationship to participants might impact learning under
+different conditions. To properly address these questions would require
+a major research program, including both cross-cultural comparative
+studies (Barrett 2020) (Barrett 2020) and more naturalistic study
+designs. While costly, such research would produce results of broad
+relevance to anthropologists, biologists, psychologists, and
+sociologists interested in teaching and learning, well beyond any
+particular implications for language evolution.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-allen2020}{}%
+Allen, Kelsey R., Kevin A. Smith, and Joshua B. Tenenbaum. 2020.
+{``Rapid Trial-and-Error Learning with Simulation Supports Flexible Tool
+Use and Physical Reasoning.''} \emph{Proceedings of the National Academy
+of Sciences} 117 (47): 29302--10.
+\url{https://doi.org/10.1073/pnas.1912341117}.
+
 \leavevmode\hypertarget{ref-barrett2020}{}%
@@
 \url{https://doi.org/10.1016/j.tics.2020.05.007}.
+
+\leavevmode\hypertarget{ref-bilker2012}{}%
+Bilker, Warren B., John A. Hansen, Colleen M. Brensinger, Jan Richard,
+Raquel E. Gur, and Ruben C. Gur. 2012. {``Development of Abbreviated
+Nine-Item Forms of the Raven{'}s Standard Progressive Matrices Test.''}
+\emph{Assessment} 19 (3): 354--69.
+\url{https://doi.org/10.1177/1073191112446655}.
 
@@
 
+\leavevmode\hypertarget{ref-boyette2017}{}%
+Boyette, Adam H., and Barry S. Hewlett. 2017. {``Autonomy, Equality, and
+Teaching Among Aka Foragers and Ngandu Farmers of the Congo Basin.''}
+\emph{Human Nature} 28 (3): 289--322.
+\url{https://doi.org/10.1007/s12110-017-9294-y}.
+
+\leavevmode\hypertarget{ref-burnham2002}{}%
+Burnham, Kenneth P., and David R. Anderson. 2002. \emph{Model Selection
+and Multimodel Inference: A Practical Information-Theoretic Approach}.
+2nd ed. New York: Springer-Verlag. \url{https://doi.org/10.1007/b97636}.
+
+\leavevmode\hypertarget{ref-callahan1979}{}%
+Callahan, Errett. 1979. {``THE BASICS OF BIFACE KNAPPING IN THE EASTERN
+FLUTED POINT TRADITION: A MANUAL FOR FLINTKNAPPERS AND LITHIC
+ANALYSTS.''} \emph{Archaeology of Eastern North America} 7 (1): 1--180.
+\url{https://www.jstor.org/stable/40914177}.
+
 \leavevmode\hypertarget{ref-cataldo2018}{}%
@@
 
+\leavevmode\hypertarget{ref-cattell1963}{}%
+Cattell, Raymond B. 1963. {``Theory of Fluid and Crystallized
+Intelligence: A Critical Experiment.''} \emph{Journal of Educational
+Psychology} 54 (1): 1--22. \url{https://doi.org/10.1037/h0046743}.
+
+\leavevmode\hypertarget{ref-coolidge2005}{}%
+Coolidge, Frederick L., and Thomas Wynn. 2005. {``Working Memory, Its
+Executive Functions, and the Emergence of Modern Thinking.''}
+\emph{Cambridge Archaeological Journal} 15 (1): 5--26.
+\url{https://doi.org/10.1017/S0959774305000016}.
+
 \leavevmode\hypertarget{ref-darwin1859}{}%
@@
 \url{https://doi.org/10.1179/2051618515Y.0000000016}.
+
+\leavevmode\hypertarget{ref-engle2018}{}%
+Engle, Randall W. 2018. {``Working Memory and Executive Attention: A
+Revisit.''} \emph{Perspectives on Psychological Science} 13 (2):
+190--93. \url{https://doi.org/10.1177/1745691617720478}.
 
@@
 
+\leavevmode\hypertarget{ref-eren2014}{}%
+Eren, Metin I., Christopher I. Roos, Brett A. Story, Noreen von
+Cramon-Taubadel, and Stephen J. Lycett. 2014. {``The Role of Raw
+Material Differences in Stone Tool Shape Variation: An Experimental
+Assessment.''} \emph{Journal of Archaeological Science} 49: 472--87.
+\url{https://doi.org/10.1016/j.jas.2014.05.034}.
+
+\leavevmode\hypertarget{ref-faisal2010}{}%
+Faisal, Aldo, Dietrich Stout, Jan Apel, and Bruce Bradley. 2010. {``The
+Manipulative Complexity of Lower Paleolithic Stone Toolmaking.''}
+\emph{PLOS ONE} 5 (11): e13718.
+\url{https://doi.org/10.1371/journal.pone.0013718}.
+
+\leavevmode\hypertarget{ref-fitts1954}{}%
+Fitts, Paul M. 1954. {``The Information Capacity of the Human Motor
+System in Controlling the Amplitude of Movement.''} \emph{Journal of
+Experimental Psychology} 47 (6): 381--91.
+\url{https://doi.org/10.1037/h0055392}.
+
+\leavevmode\hypertarget{ref-garcuxeda-medrano2019}{}%
+García-Medrano, Paula, Andreu Ollé, Nick Ashton, and Mark B. Roberts.
+2019. {``The Mental Template in Handaxe Manufacture: New Insights into
+Acheulean Lithic Technological Behavior at Boxgrove, Sussex, UK.''}
+\emph{Journal of Archaeological Method and Theory} 26 (1): 396--422.
+\url{https://doi.org/10.1007/s10816-018-9376-0}.
+
 \leavevmode\hypertarget{ref-guxe4rdenfors2017}{}%
@@
 
+\leavevmode\hypertarget{ref-hecht2013a}{}%
+Hecht, Erin E., David A. Gutman, Todd M. Preuss, Mar M. Sanchez, Lisa A.
+Parr, and James K. Rilling. 2013. {``Process Versus Product in Social
+Learning: Comparative Diffusion Tensor Imaging of Neural Systems for
+Action Execution{{}}observation Matching in Macaques, Chimpanzees, and
+Humans.''} \emph{Cerebral Cortex} 23 (5): 1014--24.
+\url{https://doi.org/10.1093/cercor/bhs097}.
+
+\leavevmode\hypertarget{ref-hecht2013b}{}%
+Hecht, Erin E., Lauren E. Murphy, David A. Gutman, John R. Votaw, David
+M. Schuster, Todd M. Preuss, Guy A. Orban, Dietrich Stout, and Lisa A.
+Parr. 2013. {``Differences in Neural Activation for Object-Directed
+Grasping in Chimpanzees and Humans.''} \emph{The Journal of
+Neuroscience} 33 (35): 14117--34.
+\url{https://doi.org/10.1523/JNEUROSCI.2172-13.2013}.
+
+\leavevmode\hypertarget{ref-henrich2010}{}%
+Henrich, Joseph, Steven J. Heine, and Ara Norenzayan. 2010. {``Most
+People Are Not WEIRD.''} \emph{Nature} 466 (7302): 29--29.
+\url{https://doi.org/10.1038/466029a}.
+
 \leavevmode\hypertarget{ref-hewes1993}{}%
@@
 Ingold, 20--31. Cambridge: Cambridge University Press.
+
+\leavevmode\hypertarget{ref-heyes2018}{}%
+Heyes, Cecilia. 2018. {``Enquire Within: Cultural Evolution and
+Cognitive Science.''} \emph{Philosophical Transactions of the Royal
+Society B: Biological Sciences} 373 (1743): 20170051.
+\url{https://doi.org/10.1098/rstb.2017.0051}.
 
@@
 
+\leavevmode\hypertarget{ref-kendal2018}{}%
+Kendal, Rachel L., Neeltje J. Boogert, Luke Rendell, Kevin N. Laland,
+Mike Webster, and Patricia L. Jones. 2018. {``Social Learning
+Strategies: Bridge-Building Between Fields.''} \emph{Trends in Cognitive
+Sciences} 22 (7): 651--65.
+\url{https://doi.org/10.1016/j.tics.2018.04.003}.
+
 \leavevmode\hypertarget{ref-key2019}{}%
@@
 
+\leavevmode\hypertarget{ref-key2015}{}%
+Key, Alastair J. M., and Christopher J. Dunmore. 2015. {``The Evolution
+of the Hominin Thumb and the Influence Exerted by the Non-Dominant Hand
+During Stone Tool Production.''} \emph{Journal of Human Evolution} 78
+(January): 60--69. \url{https://doi.org/10.1016/j.jhevol.2014.08.006}.
+
 \leavevmode\hypertarget{ref-key2018}{}%
-Key, Alastair J. M., and Christopher J. Dunmore. 2018. {``Manual
-Restrictions on Palaeolithic Technological Behaviours.''} \emph{PeerJ} 6
-(August): e5399. \url{https://doi.org/10.7717/peerj.5399}.
+---------. 2018. {``Manual Restrictions on Palaeolithic Technological
+Behaviours.''} \emph{PeerJ} 6 (August): e5399.
+\url{https://doi.org/10.7717/peerj.5399}.
+
+\leavevmode\hypertarget{ref-key2014}{}%
+Key, Alastair J. M., and Stephen J. Lycett. 2014. {``Are Bigger Flakes
+Always Better? An Experimental Assessment of Flake Size Variation on
+Cutting Efficiency and Loading.''} \emph{Journal of Archaeological
+Science} 41 (January): 140--46.
+\url{https://doi.org/10.1016/j.jas.2013.07.033}.
+
+\leavevmode\hypertarget{ref-khreisheh2013}{}%
+Khreisheh, Nada N., Danielle Davies, and Bruce A. Bradley. 2013.
+{``Extending Experimental Control: The Use of Porcelain in Flaked Stone
+Experimentation.''} \emph{Advances in Archaeological Practice} 1 (1):
+38--46. \url{https://doi.org/10.7183/2326-3768.1.1.37}.
+
+\leavevmode\hypertarget{ref-kline2015}{}%
+Kline, Michelle Ann. 2015. {``How to learn about teaching: An
+evolutionary framework for the study of teaching behavior in humans and
+other animals.''} \emph{The Behavioral and Brain Sciences} 38: e31.
+\url{https://doi.org/10.1017/S0140525X14000090}.
 
@@
 \url{https://doi.org/10.3758/s13423-016-1077-7}.
+
+\leavevmode\hypertarget{ref-levinson2014}{}%
+Levinson, Stephen C., and Judith Holler. 2014. {``The Origin of Human
+Multi-Modal Communication.''} \emph{Philosophical Transactions of the
+Royal Society B: Biological Sciences} 369 (1651): 20130302.
+\url{https://doi.org/10.1098/rstb.2013.0302}.
 
@@
 (1): 1--14. \url{https://doi.org/10.1038/s41598-017-14322-y}.
+
+\leavevmode\hypertarget{ref-marwick2017}{}%
+Marwick, Ben. 2017. {``Computational Reproducibility in Archaeological
+Research: Basic Principles and a Case Study of Their Implementation.''}
+\emph{Journal of Archaeological Method and Theory} 24 (2): 424--50.
+\url{https://doi.org/10.1007/s10816-015-9272-9}.
 
@@
 
+\leavevmode\hypertarget{ref-molleman2019}{}%
+Molleman, Lucas, Ralf H. J. M. Kurvers, and Wouter van den Bos. 2019.
+{``Unleashing the BEAST: A Brief Measure of Human Social Information
+Use.''} \emph{Evolution and Human Behavior} 40 (5): 492--99.
+\url{https://doi.org/10.1016/j.evolhumbehav.2019.06.005}.
+
 \leavevmode\hypertarget{ref-montagu1976}{}%
@@
 \leavevmode\hypertarget{ref-pargeter2019}{}%
-Pargeter, Justin, and John J. Shea. 2019. {``Going Big Versus Going
-Small: Lithic Miniaturization in Hominin Lithic Technology.''}
-\emph{Evolutionary Anthropology: Issues, News, and Reviews} 28 (2):
-72--85. \url{https://doi.org/10.1002/evan.21775}.
+Pargeter, Justin, Nada Khreisheh, and Dietrich Stout. 2019.
+{``Understanding Stone Tool-Making Skill Acquisition: {Experimental}
+Methods and Evolutionary Implications.''} \emph{Journal of Human
+Evolution} 133 (August): 146--66.
+\url{https://doi.org/10.1016/j.jhevol.2019.05.010}.
+
+\leavevmode\hypertarget{ref-pelegrin1990}{}%
+Pelegrin, Jacques. 1990. {``Prehistoric Lithic Technology : Some Aspects
+of Research.''} \emph{Archaeological Review from Cambridge} 9 (1):
+116--25.
+\href{https:///paper/Prehistoric-Lithic-Technology-}{/paper/Prehistoric-Lithic-Technology-}.
 
@@
 72 (5): 692--97. \url{https://doi.org/10.1016/j.neuron.2011.11.001}.
+
+\leavevmode\hypertarget{ref-prasciunas2007}{}%
+Prasciunas, Mary M. 2007. {``Bifacial Cores and Flake Production
+Efficiency: An Experimental Test of Technological Assumptions.''}
+\emph{American Antiquity} 72 (2): 334--48.
+\url{https://doi.org/10.2307/40035817}.
 
@@
 
+\leavevmode\hypertarget{ref-rueden2017}{}%
+Rueden, Curtis T., Johannes Schindelin, Mark C. Hiner, Barry E. DeZonia,
+Alison E. Walter, Ellen T. Arena, and Kevin W. Eliceiri. 2017.
+{``ImageJ2: ImageJ for the Next Generation of Scientific Image Data.''}
+\emph{BMC Bioinformatics} 18 (1): 529.
+\url{https://doi.org/10.1186/s12859-017-1934-z}.
+
+\leavevmode\hypertarget{ref-sasaki2007}{}%
+Sasaki, Hideo, Fumiyoshi Kasagi, Michiko Yamada, and Shoichiro Fujita.
+2007. {``Grip Strength Predicts Cause-Specific Mortality in Middle-Aged
+and Elderly Persons.''} \emph{The American Journal of Medicine} 120 (4):
+337--42. \url{https://doi.org/10.1016/j.amjmed.2006.04.018}.
+
+\leavevmode\hypertarget{ref-schillinger2014}{}%
+Schillinger, Kerstin, Alex Mesoudi, and Stephen J. Lycett. 2014.
+{``Copying Error and the Cultural Evolution of {``}Additive{''} Vs.
+{``}Reductive{''} Material Traditions: An Experimental Assessment.''}
+\emph{American Antiquity} 79 (1): 128--43.
+\url{https://doi.org/10.7183/0002-7316.79.1.128}.
+
 \leavevmode\hypertarget{ref-shallice1982}{}%
@@
 
+\leavevmode\hypertarget{ref-shea2015}{}%
+Shea, John J. 2015. {``Making and Using Stone Tools: Advice for Learners
+and Teachers and Insights for Archaeologists.''} \emph{Lithic
+Technology} 40 (3): 231--48.
+\url{https://doi.org/10.1179/2051618515Y.0000000011}.
+
 \leavevmode\hypertarget{ref-shea2016}{}%
-Shea, John J. 2016. \emph{Stone Tools in Human Evolution: Behavioral
+---------. 2016. \emph{Stone Tools in Human Evolution: Behavioral
 Differences Among Technological Primates}. Cambridge: Cambridge
@@
 159184. Boulder, CO: University Press of Colorado.
+
+\leavevmode\hypertarget{ref-stout2013}{}%
+---------. 2013. {``Neuroscience of Technology.''} In, edited by Peter
+J. Richerson and Morten H. Christiansen, 157--73. Cambridge, MA: The MIT
+Press.
+
+\leavevmode\hypertarget{ref-stout2014}{}%
+Stout, Dietrich, Jan Apel, Julia Commander, and Mark Roberts. 2014.
+{``Late Acheulean Technology and Cognition at Boxgrove, UK.''}
+\emph{Journal of Archaeological Science} 41 (January): 576--90.
+\url{https://doi.org/10.1016/j.jas.2013.10.001}.
 
@@
 
+\leavevmode\hypertarget{ref-stoutCognitiveDemandsLower2015}{}%
+Stout, Dietrich, Erin Hecht, Nada Khreisheh, Bruce Bradley, and Thierry
+Chaminade. 2015. {``Cognitive {Demands} of {Lower} {Paleolithic}
+{Toolmaking}.''} \emph{PLOS ONE} 10 (4): e0121804.
+\url{https://doi.org/10.1371/journal.pone.0121804}.
+
 \leavevmode\hypertarget{ref-stout2015}{}%
@@
 
+\leavevmode\hypertarget{ref-stoutKnappingSkillEarliest2006}{}%
+Stout, Dietrich, and Sileshi Semaw. 2006. {``Knapping Skill of the
+Earliest Stone Toolmakers: Insights from the Study of Modern Human
+Novices.''} In \emph{The {Oldowan}: {Case} Studies into the Earliest
+{Stone} {Age}}, edited by Nicholas Toth and Kathy Schick, 307--20.
+Gosport, IN: Stone Age Institute Press.
+
 \leavevmode\hypertarget{ref-tehrani2008}{}%
@@
 
+\leavevmode\hypertarget{ref-unsworth2005}{}%
+Unsworth, Nash, and Randall W. Engle. 2005. {``Individual Differences in
+Working Memory Capacity and Learning: Evidence from the Serial Reaction
+Time Task.''} \emph{Memory \& Cognition} 33 (2): 213--20.
+\url{https://doi.org/10.3758/BF03195310}.
+
 \leavevmode\hypertarget{ref-washburn1960}{}%
@@
 \url{https://doi.org/10.1038/scientificamerican0960-62}.
+
+\leavevmode\hypertarget{ref-whiten2015}{}%
+Whiten, Andrew. 2015. {``Experimental Studies Illuminate the Cultural
+Transmission of Percussive Technologies in Homo and Pan.''}
+\emph{Philosophical Transactions of the Royal Society B: Biological
+Sciences} 370 (1682): 20140359.
+\url{https://doi.org/10.1098/rstb.2014.0359}.
+
+\leavevmode\hypertarget{ref-whittaker1994}{}%
+Whittaker, John C. 1994. \emph{Flintknapping Making and Understanding
+Stone Tools By John C. Whittaker}. Austin, TX: University of Texas
+Press.
 
@@
 
+\leavevmode\hypertarget{ref-wind2010}{}%
+Wind, Anne E., Tim Takken, Paul J. M. Helders, and Raoul H. H.
+Engelbert. 2010. {``Is Grip Strength a Predictor for Total Muscle
+Strength in Healthy Children, Adolescents, and Young Adults?''}
+\emph{European Journal of Pediatrics} 169 (3): 281--87.
+\url{https://doi.org/10.1007/s00431-009-1010-4}.
+
 \leavevmode\hypertarget{ref-wynn1979}{}%

### 2021-08-14 manuscript/manuscript.Rmd
change the layout and add the discussion section

--- 2021-07-31 manuscript/Testing the cognitive foundations of Paleolithic social transmission.Rmd
@@
 title: Testing the motor and cognitive foundations of Paleolithic social transmission
-subtitle: 
-thanks: | 
-    Grants or other notes about the article that should go on the front 
-    page should be placed here. General acknowledgments should be placed at the
-    end of the article.
-
-authors: 
-- name: Justin Pargeter 1, 2
-  address: 
-  - 1. Department of Anthropology, New York University, New York, NY, USA; 2. Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa
-  email: justin.pargeter@nyu.edu
-
-- name: Megan Beney Kilgore 3
-  address: 3. Department of Anthropology, Emory University, Atlanta, GA, USA
-  email: megan.elizabeth.beney@emory.edu
-
-- name: Cheng Liu 3
-  address: 3. Department of Anthropology, Emory University, Atlanta, GA, USA
-  email: raylc1996@outlook.com
-
-- name: Dietrich Stout 3
-  address: 3. Department of Anthropology, Emory University, Atlanta, GA, USA
-  email: dwstout@emory.edu
-
-keywords:
-- Oldowan
-- Stone toolmaking
-- Social learning
-- Individual variation
-- Cognitive aptitudes
-- Motor skills
-
-#PACS: 
-#- PAC1
-#- superPAC
-
-abstract: |
-  Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret.
+author: 
+- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
+- Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
+- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
+- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
+abstract: "Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret.
 Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability.
@@
 Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group.
-These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.
-
+These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.\\par \\textbf{Keywords:} Oldowan; Stone toolmaking; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
+fontsize: 11pt
+output: bookdown::pdf_document2
+header-includes:
+   - \usepackage{textcomp}
+   - \usepackage{utopia}
+   - \usepackage{lineno}
+   - \linenumbers
+linestretch: 1.5
+link-citations: yes
+linkcolor: blue
 bibliography: bibliography.bib
-bibstyle: spmpsci
-# bibstyle options spbasic(default), spphys, spmpsci
-output: rticles::springer_article
+csl: apa.csl
 ---
 
-# Introduction {#intro}
+# Introduction
 
@@
 
-## **Individual Differences**
+## Individual Differences
 
@@
 
-## **Teaching, Language, and Tool Making**
+## Teaching, Language, and Tool Making
 
-*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.* [@montagu1976: 267]
+“*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.*”[@montagu1976: 267]
 
@@
 
-## **Raw materials and knapping skill**
+## Raw materials and knapping skill
 
@@
 
-# **Materials and Methods**
+# Materials and Methods
 
@@
 
-## **Participants**
+## Participants
 
@@
 
-## **Study Visit**
+## Study Visit
 
@@
 
-\
-
-## **Individual Difference Measures**
+## Individual Difference Measures
 
@@
 
-\
-
-## **Stone Tool Making**
+## Stone Tool Making
 
@@
 
-### **Raw Materials**
+### Raw Materials
 
@@
 
-### **Experimental Conditions**
+### Experimental Conditions
 
@@
 
-**Lithic Analysis**
+## Lithic Analysis
 
@@
 
-## **Statistical Analyses**
+## Statistical Analyses
 
@@
 
-# **Results**
+# Results
 
@@
 
-## **Principal Component analyses**
+## Principal Component analyses
 
@@
 
-### **Flake size and shape**
+### Detached Piece size and shape
 
@@
 
-### **Lithic flaking performance measures**
+### Lithic flaking performance measures
 
@@
 
-## **Do trained, untrained, and expert knappers perform differently?**
+## Relationships between Performance Measures
+
+This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (Figures).
+As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (Fig.
+Xa), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj.
+R2 = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj.
+R2 = 0.4) group.
+Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj.
+R2 = 0.7), but no relation in the Trained group (Fig.
+Xb).
+A qualitatively similar trend with respect to Quality vs.
+Quantity (Fig.
+Xc) did not achieve significance (F[1, 17] = 0.6, p 0.2).
+
+Thus, it appears that Trained participants achieved higher Total Productivity by increasing average flaking Quantity across cores and without sacrificing Quality, whereas Untrained participants found other ways to vary Total Productivity (e.g., number of cores knapped rather than Quantity per core, see variance Table and Figure) and generally increased productivity at the expense of Quality.
+Experimental artifacts illustrating these trade-offs are presented in Figure.
+
+## Do trained, untrained, and expert knappers perform differently?
 
@@
 
-## **Does training/practice time impact flaking performance?**
+## Does performance change over time?
 
@@
 
-## **Do individual differences in motor skill and psychometric measures predict flaking performance?**
+## Do individual differences in motor skill and psychometric measures predict flaking performance?
 
@@
 
-### **Model 1: Individual differences and quantity flaking**
+### Model 1: Individual differences and quantity flaking
 
@@
 
-*Quantity flaking $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength*
+*Quantity flaking* $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength
 
@@
 
-*Quantity flaking $\sim$ Training condition + Highest n-back level  + BEAST score + Grip strength + Fitt’s score$\cdot$Raven’s Progressive Matrix score + Training condition$\cdot$Highest n-back level + Training condition$\cdot$BEAST score + Nodule mass (as control)*
+*Quantity flaking* $\sim$ Training condition + Highest n-back level + BEAST score + Grip strength + Fitt's score$\cdot$Raven's Progressive Matrix score + Training condition$\cdot$Highest n-back level + Training condition$\cdot$BEAST score + Nodule mass (as control)
 
@@
 
-### **Model 2: Individual differences and quality flaking**
+### Model 2: Individual differences and quality flaking
 
@@
 
-*Quality flaking $\sim$ Highest n-back level  + Fitt’s score + Grip strength + Fitt’s score$\cdot$BEAST score + Grip strength$\cdot$BEAST score + Grip strength$\cdot$Fitt’s score + Training condition$\cdot$Grip strength + Nodule mass (as control)*
+*Quality flaking* $\sim$ Highest n-back level + Fitt's score + Grip strength + Fitt's score$\cdot$BEAST score + Grip strength$\cdot$BEAST score + Grip strength$\cdot$Fitt's score + Training condition$\cdot$Grip strength + Nodule mass (as control)
 
@@
 Strong subjects with a low propensity to use social information tend to perform worst in terms of quality flaking.
-This result makes sense if one considers that our quantity flaking model showed how social information equalizes performance levels in trained and untrained subjects. \
-\
+This result makes sense if one considers that our quantity flaking model showed how social information equalizes performance levels in trained and untrained subjects.
+
 The second interaction involves grip strength and training.
@@
 
-# **Discussion**
+### Model 3: Individual differences and quality flaking
 
-we have little basis other than personal experience and/or tradition [@callahan1979; @whittaker1994; @shea2015] and theoretical speculation [@stout2013; @whiten2015] from which to assess what pedagogical techniques are most effective even in WEIRD contexts.
-For example, no study to date has considered how variation in teacher skill [@shea2015] or social relationship to participants might impact learning under different conditions.
-To properly address these questions would require a major research program, including both cross-cultural comparative studies [@barrett2020] (Barrett 2020) and more naturalistic study designs.
-While costly, such research would produce results of broad relevance to anthropologists, biologists, psychologists, and sociologists interested in teaching and learning, well beyond any particular implications for language evolution.
+## Behavioral observations
 
-# **Conclusions**
+We designed this exploratory study primarily to trial experimental design elements such as training time, conditions, and raw materials and to collect preliminary data on the effect of individual differences and training on knapping outcomes.
+We thus focused on collecting quantitative psychometric and lithic data.
+However, we also considered that quantifying participant knapping behaviors as well as products could be important for future studies.
+To support methods development in this regard, we made ad hoc notes on observed behaviors during the experiments and video-recorded all experiments to enable later, more systematic analyses yet to be completed.
+However, even casual behavior observation was sufficient to reveal an unexpected effect.
+Whereas all trained participants copied the general posture and technique of the expert (free hand knapping seated in a chair) fully half (6) of the uninstructed participants experimented with or even knapped all of their cores using the floor as a support (Figure).
+Three of these participants were in the same session, which is also the only group composed of just three individuals.
+In this group, knapping on the ground appears to have been transmitted from one participant to the other two based on appearance order and the point of gaze of participants.
 
-# **Acknowledgments**
+# Discussion
 
-# **References**
+The most salient finding of our exploratory study is that the presence/absence of teaching clearly impacted knapping performance but did so in nuanced and individually variable ways that have not been explored in previous studies.
+In fact, we did not observe any significant differences in mean performance between our experimental conditions.
+Some non-significant tendencies toward enhanced Instructed group performance suggest that a larger participant sample might detect significant effects, but also that the size of any such effects would likely remain small.
+This could reasonably lead to the conclusion that teaching does not substantially facilitate early stage knapping skill acquisition [cf.
+@putt2014; @ohnuma1997].
+Looking closer, however, we found a number of important teaching effects.
+
+## Variance Reduction
+
+In our experiment, the strongest effects of teaching were to reduce variance (Figure, Table) rather than shift mean values.
+In particular, teaching acted as a "safety net" that homogenized performance by reducing the frequency of extremely poor outcomes (i.e., learning failures).
+This finding provides additional support for the hypothesis that teaching would have increased the reliability of Oldowan skill reproduction [@morgan2015] while simultaneously corroborating the view that basic flaking competence can be achieved in its absence [@tennie2017].
+Our results thus do not imply that teaching was required or even present during Oldowan times (but see @gärdenfors2017), but rather serve to reinforce the plausibility of co-evolutionary scenarios positing the cost/reliability of technological skill acquisition as a selection pressure favoring the evolution of teaching and language [@stout2010; @morgan2015; @stout2017].
+
+## Knapping Behaviors
+
+The current study complements the transmission chain design of Morgan et al.
+[-@morgan2015] by finding similar effects in more naturalistic learning contexts.
+Our design further allowed us to examine individual variation to better understand how teaching produces its effects.
+Whereas transmission chains are optimized to investigate iterative learning effects [but see @caldwell2020], they necessarily involve a different instructor/model for each participant.
+We sacrificed this iterative component in order to consider how the presence/absence of teaching affected the behavior of individuals under otherwise standardized learning conditions.
+
+We found that a key impact of teaching was to alter basic flake production strategies, as reflected in the relationship between Total Productivity and detached piece Quality (Fig.Xb).
+Whereas Untrained participants achieved greater Productivity at the expense of Quality and vice versa, these dimensions were unrelated across Trained participants.
+Thus, even though Untrained participants achieved the highest values on each metric, only trained individuals managed to maximize both simultaneously.
+Indeed, a core function of teaching is to reduce the search space that learners must explore and increase the likelihood of discovering globally as opposed to locally optimum solutions [cf.
+@hinton1996; @stout2013].
+In our study, Untrained individuals explored a greater range of basic behavioral variations not seen in the Trained group, including knapping on the floor, concentrating on working just a few cores (2-4, Figure) over the practice period, and showing less constrained nodule size preferences (Figure).
+It is notable that this variation occurred even in the presence of an observable expert example, suggesting it may be interesting for future experiments to address the impact of social context, expectations, and relationships on observational learning strategies [@kendal2018].
+
+We also found a strong positive effect of Grip Strength on Total Productivity independent of learning condition (Figure).
+While it is tempting to interpret this with respect to the demands for hand strength specifically, it is important to remember that grip strength is strongly correlated with total muscle strength [@wind2010] and overall fitness [@sasaki2007].
+Thus, it is best taken to indicate some importance of fitness generally in increasing the rate and intensity of core reduction by naïve knappers, potentially affecting rate of work and the kinetic energy of the swing as well as the handling of core and hammerstone.
+It thus provides further support for hypotheses positing stone tool making as a selection pressure on the functional anatomy of hand, arm, and shoulder [e.g., @williams-hatala2018], but initially appears orthogonal to variations in learning condition and knapping behaviors in our study.
+However, we also found that the effect of Grip Strength on per-core knapping Quantity is dependent on teaching (Figure).
+The absence of this effect in the uninstructed group reflects the weaker association between Total Productivity and per-core Quality across these participants (Fig.Xa) and shows Grip Strength increased uninstructed Productivity specifically by allowing them to knap more cores rather than to reduce individual cores more heavily.
+In keeping with this, uninstructed Grip Strength is positively correlated with Total Cores knapped (r2 = 0.54, p = 0.01).
+Conversely, strength allowed instructed participants to increase their average Quantity per core without affecting the total number of cores knapped (r2 = 0.18, p = 0.165).
+Thus, strength appears to have achieved its effects on core reduction rate and intensity in different ways, depending on teaching.
+This difference is likely related to the homogenizing effect of teaching on knapping rate (all instructed participants knapped 6 or more cores) and methods.
+Subjectively, knapping behaviors of uninstructed participants often appeared more physically demanding (e.g., greater number of non-productive blows, rapid and unregulated battering) which would imply different demands on both strength and aerobic fitness [@williams-hatala2021; @mateos2019].
+However, this remains to be systematically investigated.
+
+In this respect, it is also important to note that we do not know how well the knapping objectives and strategies communicated by the expert in our experiment correspond to actual Oldowan goals and behaviors.
+The instructor has successfully replicated assemblage-level patterning at Gona [@stout2019] but Oldowan behavior is variable across space and time [e.g., @braun2019] and alternative knapping methods might maximize different values (productivity, quality, effort), especially in novices [@putt2015] (Putt 2015).
+Nevertheless, the effect of instruction to constrain behavioral exploration and homogenize outcomes is clear.
+We expect that this effect would generalize to the teaching of alternative knapping goals and behaviors, although this remains to be tested.
+
+## Learning Strategies
+
+One major goal of this experiment was to test the viability of a moderate, two-hour, learning period for studies of skill acquisition.
+Unfortunately, we found that this duration was insufficient to capture learning effects for Oldowan-like flake production.
+The lack of performance change over the period (Figure) cannot be attributed to a ceiling effect (i.e., rapid task mastery at the outset of the practice period) as participants remained well below expert levels and continued to display the high within-individual variability typical of naïve/novice knapping [@eren2011; @pargeter2019].
+This negative result was unexpected but is broadly consistent with evidence that Early Stone Age flaking, while conceptually simple, requires substantial practice for perceptual-motor skill development [@nonaka2010; @stout2015; @pargeter2020].
+Future investigations of learning variation across individuals and/or experimental conditions may thus need to incorporate longer practice periods to capture skill acquisition processes.
+In theory, much shorter knapping trials might be used to assess the variation in initial performance across individuals and under different conditions that is captured in our study.
+However, the presence of substantial core-to-core variation within individuals cautions against overly brief experiments that might not provide a representative sample.
+Greater durations also allow for the expression of different learning strategies over time, even in the absence of directional performance change.
+
+At a basic level, learners of any new task must balance investment in task exploration vs.
+exploitation of knowledge and skills already in hand [@sutton2018].
+Premature exploitation risks settling for a sub-optimal local solution whereas continued exploration sacrifices more immediate payoffs.
+Managing this trade-off is especially challenging for complex, real-world tasks like stone knapping, and is thought to depend on the interplay of uncertainty and reward expectation [@wilson2021].
+Teaching and social learning generally have the potential to provide low-cost information about task structure and payoffs [@rendell2010; @kendal2018], which if adopted, would be expected to affect exploration/exploitation decisions.
+Such adoption is itself known to be influenced by individual cognitive differences, for example if higher fluid intelligence allows observers to better understand observed tasks [@vostroknutov2018] or if individuals vary in their tendency to use and value social information [@toelch2014; @molleman2019].
+
+In our study, we did not observe any effect of fluid intelligence (RPM) on knapping outcomes but did find strong interactions of learning condition with participant visuospatial working memory and social information use tendency (Figure).
+As expected, uninstructed individuals with higher scores on these dimensions displayed higher Total Productivity and average per-core flaking Quantity (although the effect on n-Back on Quantity did not achieve significance).
+We attribute these effects to increased ability to hold relevant morphological/spatial information in mind and a tendency to benefit from observing successful strategies of others, including the expert model.
+In contrast, instructed individuals with higher scores tended to have lower Productivity and Quantity.
+We interpret this unexpected effect to an increased tendency to privilege exploratory learning behavior over exploitation.
+In particular, we suggest that trained participants might knap more slowly and less productively if higher working memory capacity inclined them to experiment more with morphological/spatial variables highlighted by the instructor or if a predisposition to use social information use caused them to invest greater time and effort attending to and trying out observed actions and/or instructions.
+These suggestions remain to be tested by further work.
+
+Unfortunately, the training period in our current experiment was insufficient to capture learning effects and so we have no evidence of the effects of these individual differences and putative exploration/exploitation tradeoffs to the ultimate achievement of expertise.
+A similar negative effect of instruction on knapping outcomes during early stage learning was reported by Putt et al.
+[-@putt2014], and has been interpreted to reflect learners experimenting with advanced techniques before they have the perceptual-motor skill to execute them [@whiten2015; @stout2015].
+Such effects might be further explored with more detailed behavioral data, as opposed to purely lithic data, and with longer learning periods.
+
+## Limitations and Prospects
+
+Although our exploratory study produced a number of robust results with respect to the effects of instruction and individual differences on lithic products, it is clearly limited by a small sample size, short training duration, and lack of detailed quantification of observed behaviors.
+These are limitations that can hopefully be addressed in future studies building on the methods and evidence presented here.
+For example, it is notable that our study failed to document any reliable effects on knapping Quality.
+Obviously, this might reflect an actual lack of such effects, but it may also indicate a need for more sensitive measures and/or increased sample size and training duration to identify subtle or delayed effects.
+One aspect of our attempt to balance pragmatic costs and benefits in our study was to test the efficacy of relatively limited lithic analysis.
+More detailed ongoing analyses of core morphology and debitage features (e.g., typology, cutting edge length, platform dimensions) may yet reveal a more reliable signal of knapping quality.
+Results of the Quality model in particular also seem to suffer from the uneven distribution and discrete rather than continuous nature of scores on our RPM and n-Back tests.
+Concerns about the sampling of variation on these dimensions could be addressed with larger samples or by pre-screening participants to ensure more even representation.
+Alternative psychometric tests (e.g., full rather than short version of the RPM) might also provide more sensitive and continuous measures.
+
+Another major limitation that our study shares with all other published experiments on knapping skill acquisition is that we do not address variation in social and cultural context or in teaching style.
+Currently, we have little basis other than personal experience/tradition [@callahan1979; @whittaker1994; @shea2015] and theoretical speculation [@stout2013; @whiten2015] from which to assess which pedagogical techniques are most effective even in WEIRD contexts.
+No study to date has considered how variation in teacher skill [@shea2015] or social relationship to participants might impact learning under different conditions.
+To properly address these questions would require a major research program, including both cross-cultural comparative studies [@barrett2020] and more naturalistic study designs.
+While costly, such research would produce results of broad relevance to anthropologists, biologists, psychologists, and sociologists interested in teaching and learning.
+
+# Conclusions
+
+# Acknowledgments
+
+# References

### 2021-08-14 code/PaST_r_code.R
add all required data and update the code.

@@ -1,4 +1,4 @@
-##Libraries####
+#### Libraries ####
 detachAllPackages <- function() {
       
       basic.packages <- c("package:stats","package:graphics","package:grDevices","package:utils","package:datasets","package:methods","package:base")
@@ -12,16 +12,15 @@ detachAllPackages <- function() {
 }
 
 detachAllPackages()
+Sys.setenv(JAVA_HOME="")
+Sys.setenv(JAVA_HOME='C:\\Program Files\\Java\\jre1.8.0_301')
+library(tidyverse)
 library(car)
-library(ggplot2)
 library(reshape2)
-library(tidyr)
 library(plyr)
-library(dplyr)
 library(DescTools)
-library(stringr)
-library("factoextra")
-library("psych") 
+library(factoextra)
+library(psych) 
 library(Hmisc)
 library(glmulti)
 library(report)
@@ -31,36 +30,37 @@ library(rcompanion)
 library(caret)
 library(permuco)
 
-#### Import datasets #### ####
 
-experiment_data<-read.csv("/Users/JustinPargeter/Google Drive/Projects/PaST/Pilot experiment/Lithics/Experiment database/past_stats/past_stats/past_cores_result.csv") %>%
+#### Import datasets ####
+
+experiment_data<-read.csv("data/past_cores_result.csv") %>%
       rename(subject_id = Subject,
              core_number = nodule_number) %>%
       mutate(subject_id=as.factor(subject_id),
              core_number=as.factor(core_number),
              Condition=as.factor(Condition))
 
-experiment_core_usedata<-read.csv("/Users/JustinPargeter/Google Drive/Projects/PaST/Pilot experiment/Lithics/Experiment database/past_stats/past_stats/core_use_records.csv") %>%
+experiment_core_usedata<-read.csv("data/core_use_records.csv") %>%
       mutate(subject=as.factor(subject),
              core_number=as.factor(core_number))
 
-lithic_data<-read.csv("past_pilot_lithics.csv") %>%
+lithic_data<-read.csv("data/past_pilot_lithics.csv") %>%
       rename(flake_mass = flake_mass,
              flake_max_thickness = max_thickness,
              core_final_mass=core_mass) %>%
       mutate(core_number_nyu=core_number) %>%
       rename(subject_id_nyu=subject_id) 
 
-extra_thickness_flakes<-read.csv("extra_flakes.csv") 
[…]

### 2021-08-15 manuscript/manuscript.Rmd
update the result section and reinstall the R to load rjava package.

@@
 ## Stone Tool Making
-
 After individual difference testing, participants engaged in a 2-hour stone tool making session, with a 15-minute break after 1 hour.
@@
 ### Raw Materials
-
 Each participant was provided with 9 cores for use over the 2-hour experiment.
@@
 Cores were sorted by shape and weight and then distributed evenly to each participant.
-As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length x Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p = 0.9) of cores provided.
+As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length $\times$ Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p = 0.9) of cores provided.
 This was also true comparing the two experimental conditions (Taught vs.
@@
 ### Experimental Conditions
-
 In both conditions, three researchers were present to record activities and collect materials.
@@
 ## Lithic Analysis
-
 All finished cores were weighed and measured (L, W, T).
@@
 ## Statistical Analyses
-
 To evaluate the association between psychometric, motor-skill, and training measures and technological outcomes, we adopted an information-theoretic approach [@burnham2002].
@@
 Continuous predictors were centered such that zero represents the sample average, and units are standard deviations.
-The full model was fitted with the lm function in R 3.2.3, and the glmulti package [@Bartoń 2015] was used for multi-modal selection and model comparison.
+The full model was fitted with the lm function in R 3.2.3, and the glmulti package was used for multi-modal selection and model comparison.
 
@@
 # Results
-
 Following a recent protocol to enhance the reproducibility and data transparency of archaeological research [@marwick2017], detailed results of all analyses and assessments of the data structure are available in our paper's supplementary materials and through Github (<https://github.com/Raylc/PaST-pilot>).
 Here we limit discussion to the major ﬁndings regarding flaking performance and individual differences.
-The purpose of this section is to determine whether training impacted subject flaking performance and if any of the subject's individual psychometric and motor-skill aptitudes predict flaking performance.
+We were particularly interested in: 1) group level effects of experimental condition, 2) individual differences in aptitude and learning, and 3) potential interactions between learning conditions and individual differences.
+To address these questions, we employed data reduction (Principal Component Analysis) to derive two summary metrics of flaking performance, compared these factors across the two experimental conditions, and built multivariate models examining the relations between our various psychometric measures, subject’s motor skill scores, and our two lithic performance factors.
 
 ## Principal Component analyses
-
-The following two sections outline factor analyses designed to summarize our main study metrics tracking individual variation in flake sizes and shapes and lithic performance measures.
+The following two sections outline factor analyses designed to summarize our main study metrics tracking individual variation in DP sizes and shapes and lithic performance measures.
 
 ### Detached Piece size and shape
-
-To better understand the relationship between flake shape and training/individual variation, we entered our nine flake linear plan measurements along with maximum flake length and thickness into a principal component analysis (PCA) from which summary coordinates were extracted.
+To better understand the relationship between DP shape and training/individual variation, we entered our nine flake linear plan measurements along with maximum flake length and thickness into a principal component analysis (PCA) from which summary coordinates were extracted.
 Bartlett's Test of Sphericity was significant ($\chi ^2$ (10) =4480, p \< .01) indicating that the set of variables are adequately related for factor analysis.
@@
 Factor 3 therefore tracks flakes with a typical shape having a thin cross-section, wider base, and narrower tip.
-We used these three flake shape coordinates to approximate flake size and shape in the project's flake performance factor analysis.
+We used these three flake shape coordinates to approximate DP size and shape in the project's flake performance factor analysis.
 
 ### Lithic flaking performance measures
+To better understand the relationship between our various lithic performance measurements and to reduce data dimensionality, we conducted a second principal component analysis examining the study’s six lithic performance measures (count of large pieces [>40mm and 5g], mass of large pieces relative to total detached mass, core delta mass, and the three flake shape factors).
+All of these measures were summarized for each core and unique factor scores were calculated from these core-specific measures.
+Bartlett's Test of Sphericity was significant ($\chi ^2$ (6) =3185, p \< .01) indicating that the set of variables are at least adequately related for factor analysis.
 
-To better understand the relationship between our various lithic performance measurements and to reduce these data dimensionality, we conducted a second principal component analysis examining the study's seven lithic performance measures (mass of flakes relative to flaked core mass, count of large flakes [\>40mm and 5g], core delta mass, three flake shape factors, and the total number of cores used).
-All of these measures except the total cores flaked were summarized for each core.
-The Bartlett's Test of Sphericity was significant ($\chi ^2$ (6) =3950, p \< .01) indicating that the set of variables are at least adequately related for factor analysis.
+The analysis yielded two factors explaining a total of 56% of the variance for the entire set of variables (Table).
+Factor 1 (hereafter “Quantity”) explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and on core delta mass.
+Performance factor 2  (hereafter “Quality”) covers 27% of the sample variance and measures flaking quality as reflected in high positive loadings on Shape Factors 1 (size) and 3 (thin, “flake-like” shape) and a negative loading on Shape Factor 2 (“splinter-like” thickness and elongation).
+High scores on Quality thus reflect production of larger, relatively thinner, and more typically flake-shaped vs.
+splinter-shaped DPs.
 
-The analysis yielded two factors explaining a total of 52.3% of the variance for the entire set of variables.
-Factor 1 tracks flaking quantity due to high positive loadings on mass flakes/flaked mass, core delta mass, total cores used, and flake shape factor 2.
-Negative loadings on flake factor 1 (flake size) and 3 (basal/tip width shape) suggest that flaking quantity comes at the expense of producing lots of small and thick stone splinters/chunks.
-This first factor explained 28% of the variance.
-Higher factor 1 values reflect higher flaking quantities.
-The second factor covers 24% of the sample variance.
-Factor 2 measures one's ability to carefully flake cores due to high positive loadings on large flake count and lower or neutral loadings on mass flakes/flaked mass, core delta mass, and total cores used.
-The resulting flakes are larger, relatively thinner, and more typically shaped due to high positive loadings on flake factor 1 (size) and flake factor 2 (relative thickness) contrasted with a high negative loading on flake shape factor 3 (basal/tip width).
-Higher factor two values show increased quality flaking performance.
-
-**Figure** shows the covariance between our two flake performance principal components.
-While there is no significant relationship between the two factors, and the two study groups do not show significantly different slopes (p = 0.3), there are never-the-less differences between the two groups.
-Trained individuals are able to increase flake quality while increasing flake quality, whereas untrained individuals increase their flaking quantity at the expense of flake quality.
+These two factors address flaking performance at the level of individual cores, however we were also interested in the overall productivity/rate of work of each participant over the entire two hours.
+For example, looking at a knappers average Quality and Quantity factor scores would not differentiate between a participant who spent the entire time exhaustively reducing one core vs.
+another participant who did the same to all nine of their allotted cores in the same time.
+To capture this aspect of variation we calculated a simple Total Productivity metric as the sum of all mass a participant removed from cores during the experiment.
 
 ## Relationships between Performance Measures
-
 This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (Figures).
@@
 Xa), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj.
-R2 = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj.
-R2 = 0.4) group.
+$R^2$ = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj.
+$R^2$ = 0.4) group.
 Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj.
-R2 = 0.7), but no relation in the Trained group (Fig.
+$R^2$ = 0.7), but no relation in the Trained group (Fig.
 Xb).
@@
 Quantity (Fig.
-Xc) did not achieve significance (F[1, 17] = 0.6, p 0.2).
+Xc) did not achieve significance (F[1, 17] = 0.6, p = 0.2).
 
@@
 ## Do trained, untrained, and expert knappers perform differently?
+Here we compare our flaking outcomes (DP size/shape and flaking performance factors) between the trained and untrained groups.
+Our expert demonstrator/instructor is included as a performance benchmark.
 
-Here we compare our flaking outcomes (flake size/shape and flaking performance factors) between the trained and untrained groups to see if training in any way impacted their flaking performance.
-We added a performance comparison (individual flake and core knapping performance metrics) between these two novice groups and our expert knapper to test for differences at different points in the stone flaking skill spectrum.
+Table summarizes the results of ANOVA tests group level difference in central tendency on various performance measures.
+We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors.
+In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (Figure).
+All three of these results show either medium or large effect sizes.
+In all three comparisons, the trained group’s data distributions tended towards the expert sample although they were not significantly different from the untrained group (Figures-Core examples too).
+We also observed a significant difference in shape factor 2 (splinters) driven by the expert’s lower values, but with a very low (<0.01) effect size.
+These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.
 
-**Table** summarizes the group level performance tests.
-The results show no significant differences in flaking performance between the trained and untrained groups measured by our two flaking performance factors.
-Trained and untrained individuals overall performed equally as well in terms of flaking quality and quantity.
-Three-way flake size and shape comparisons between our expert knapper and the two novice groups show significant differences in flake shape factor 2 (relative thickness), but with a very low (\<0.01) effect size.
-This difference is driven by the expert's lower overall relative flake thickness.
-Regardless of training, our experimental subjects produced flakes that were on average the same size and shape as those of the expert trainer.
-We did, however, find several significant differences in the three-way comparisons of our individual flaking performance measures.
-The expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and on average had significantly smaller cores (effect size = 0.27) (**Figure**).
-All three of these results show either medium or large effect sizes.
-In all three comparisons, the trained group's data distributions tended towards the expert sample (although they were not significantly different from the untrained group) (**Figures-Core examples too**).
-These results show that core reduction intensity and large flake production track this experiment's greatest differences between expert and novice performance.
+While we did not find significant differences in central tendency between our two experimental groups, results (Figure and Figure) did indicate lower variance in the trained group.
+To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the trained and untrained individuals using the F-test on either core-averaged or flake specific variances.
+Table and Figure present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass).
+In most instances, variance in the untrained group exceeds that of trained individuals by 1.5 to 4.7 times.
+The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the untrained group (Figure), rather than to shift the mean.
 
 ## Does performance change over time?
-
-Here we use the relative order in which each subject flaked their cores to test for changes in flaking performance across the 2hr experiment.
-For these analyses, we calculated the relative percentage for each core relative to the total cores knapped by each subject.
+In addition to comparing overall performance during the two hour experiment, we also wanted to determine if groups or individuals differed in learning (i.e., performance change) over the period.
+For these analyses, we calculated the learning stage as the ordinal number of each core out of the total number knapped by each subject (i.e., core 2 of 4 or 4 of 8 both equal 50% complete).
 These relative core use-order percentages were then binned into 20 percent brackets for core-order and group-level comparisons.
-Flaking outcomes were tracked using the two flake performance factors (quality and quantity flaking).
+Flaking outcomes were tracked using the two performance factors (Quality and Quantity).
 We added the nodule starting mass to track whether training/practice times impacted raw material selection.
 
-**Table** shows no significant training effects across the two flaking performance measures either as grouped data or between individuals (**Figures)**.
+Table shows no significant training effects across the two performance measures either as grouped data or between individuals (Figures).
 This result demonstrated that flaking outcomes did not change dramatically across the study interval.
-The one significant main training effect related to core starting mass (with a strong main training effect size = 0.25).
-On average, core starting masses increase as subjects flake more cores, showing that subjects started with smaller nodules first that were smaller and easier to hold/flake.
-As the experiment wore on, they were left with larger and more challenging nodules.
-The small main effect of training condition is driven by higher starting nodule masses in the untrained group at the beginning, half way through, and at the end of the experiment.
+This lack of significant learning effects is confirmed by an inspection of individual learning curves (Figure).
+The one significant main training effect related to core starting mass (with a strong effect size = 0.25).
+On average, core starting masses start low and increase, showing that participants selected smaller nodules first.
+As the smaller nodules in their allotment were depleted, participants were left to knap larger, less preferred nodules.
+This preference for smaller cores is somewhat less pronounced in the untrained group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the untrained group (Figure).
 
 ## Do individual differences in motor skill and psychometric measures predict flaking performance?
-
-One of the experiment's primary goals was to test if measures of individual variation in motor skill and intelligence/motivation predict success in stone flaking.
-To address this goal, we built two multivariate models examining the relations between our various psychometric measures, subject's motor skill scores, and our two lithic performance factors (quantity flaking and quality flaking).
+One of the experiment’s primary goals was to test if measures of individual perceptual-motor and cognitive variation predict success in stone flaking across different training conditions.
+To address this goal, we built three multivariate models examining the relations between training conditions, individual difference measures, and our three lithic performance measures (overall productivity and average per-core Quantity and Quality).
 These models enabled us to determine which of the psychometric and motor skill factors are better predictors of a participant's flaking performance in the study.
 
-We have already demonstrated that subjects selected progressively larger nodules throughout the experiment.
-It is important now to understand whether nodule variability had any impact on our flaking results.
-Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p \< 0.01, R2 = 0.54) we included nodule mass as a covariate to control for any variance in flaking performance that may be driven by nodule differences.
-Our two motor skill and strength measures (grip strength and Fitt's performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, R2 = 0.41).
+We considered all possible interactions between five individual difference measures, core size, training condition, and the three performance measures, with each subject providing one data point.
+Each model’s continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitts score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations.
+Our two motor skill and strength measures (grip strength and Fitt's performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, $R^2$ = 0.41).
 However, these two measures track complementary components of athleticism (strength vs.
@@
 
-We considered all possible interactions between five individual difference measures, core size, training condition, and the two lithic flaking performance factors (quantity and quality flaking) (wherein each subject provides one data point).
-Each model's continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitt's score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations.
+The full models were fitted with the lm function in R 3.2.3, and we used the Glmulti package’s automated model selection algorithm to select the best performing model (lowest AICc score) (see methods for further details on the multimodal selection process).
+All three models follow the same complete model statement as follows:
+
+*Flaking performance variable $\sim$ Training condition + Highest n-back level + Raven’s Progressive Matrix score + BEAST score + Fitt’s score + Grip strength*
+
+For our two per-core performance factors (Quantity and Quality) it is also relevant to consider how individual core features may have affected performance.
+We found no evidence of individual or group level practice effects over the two hours, so we did not include core order in the models.
+We did, however, find that subjects selected progressively larger nodules throughout the experiment.
+It is thus important to understand whether nodule variability had any impact on our flaking results.
+Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p < 0.01, $R^2$ = 0.54) we included nodule mass as a covariate to control for any variance in flaking performance that may be driven by nodule differences.
 
 ### Model 1: Individual differences and quantity flaking
+Our first model examined variance in overall flaking productivity measured by each subject’s combined flaked mass (nodule starting mass-core final mass).
+This provides a basic measure of variation in individuals’ success detaching pieces and reducing cores from a standardized (see Methods) raw material supply.
+From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of -18 (Average AIC = -13).
+This model comprised the following statement with two main and three interaction effects:
 
-The first full model examined variance in the quantity of flaking tracked by our first performance factor explaining increases in the number of cores used, the degree of reduction on each core, and large/relatively thick flake production.
-The full model was fitted with the lm function in R 3.2.3, and we used the Glmulti package's automated model selection algorithm to select the best performing model (lowest AICc score) (see methods for further details on the multimodal selection process).
-The complete model statement is as follows:
+*Total flaked mass $\sim$ Training condition + Grip strength + RPM$\times$Highest n-back level + Fitts score$\times$BEAST score + Grip strength$\times$RPM*
 
-*Quantity flaking* $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength
+This model explains a statistically significant and substantial proportion of variance in flaking productivity ($R^2$ = 0.84, F (6, 14) = 12.7, p < 0.01, adj.
+$R^2$ = 0.77).
+A model residuals normality test shows no significant differences with the normal distribution (p = 0.72) indicating that this relationship is linear.
+A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 2, df = 6, p = 0.8).
 
-From a candidate pool of 55893 possible multivariate models, the best performing model returned an AICc value of 36 (Average AIC = 52).
+Table presents this model’s coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average.
+The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable.
+We found significant (p< 0.05) and substantial (Standardized Estimate >= 0.50, i.e.
+a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST.
+The main effect of Grip Strength (Figure), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.
+
+Effects of visuospatial working memory capacity and social information use are more complicated, as indicated by strong interactions with learning condition (Figures).
+In each case, higher scores were associated with better performance in the uninstructed group but worse performance in the instructed group.
+Positive effects in the uninstructed group were as expected, given the hypothesized importance of spatial cognition (Coolidge and Wynn 2005) and social learning (Morgan et al.
+2015) in the acquisition of knapping skills.
+Negative effects in the trained group are unexpected but presumably reflect differences in learning strategies adopted under instruction.
+
+### Model 2: Individual differences and quality flaking
+The second full model examined the variance in average flaking Quantity per core.
+It thus complements our first model assessing overall productivity by testing for differences in reduction intensity at the level of individual cores.
+From a candidate pool of 55893 possible multivariate models, the best performing model returned an AICc value of 32 (Average AIC = 44).
 This model comprised the following statement with three main and four interaction effects:
 
-*Quantity flaking* $\sim$ Training condition + Highest n-back level + BEAST score + Grip strength + Fitt's score$\cdot$Raven's Progressive Matrix score + Training condition$\cdot$Highest n-back level + Training condition$\cdot$BEAST score + Nodule mass (as control)
+*Quantity $\sim$ Highest n-back level  + BEAST score + Fitt’s score + Grip strength + Training condition$\times$Highest n-back level + Training condition$\times$BEAST score + Training condition$\times$Grip strength + Nodule mass (as control)*
 
-This model explains a statistically significant and substantial proportion of variance in quantity flaking (R2 = 0.84, F (8, 12) = 8.2, p \< 0.01, adj.
-R2 = 0.74).
-A model evaluation comparing training vs.
-test data subsets shows no evidence for overfitting (Difference in R2 between training and test models = 0.02).
-A model residuals normality test shows no significant differences with the normal distribution (p = 0.35) indicating that this relationship (as required) is linear.
-A Breusch-Pagan test showed no evidence for heteroskedasticity (whether variance for all observations in our data set are the same) (BP = 8.2, df = 7, p = 0.3).
+This model explains a statistically significant and substantial proportion of variance in quantity flaking ($R^2$ = 0.7, F (8, 12) = 3.6, p = 0.02, adj.
+$R^2$ = 0.5).
+A model residuals normality test shows no significant differences with the normal distribution (p = 0.38) indicating that this relationship (as required) is linear.
+A Breusch-Pagan test showed no evidence for heteroskedasticity (whether variance for all observations in our data set are the same) (BP = 4.4, df = 8, p = 0.8).
 
-\
-**Table** presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average.
-The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable.
-Significant increases in quantity flaking were found for subjects with training (Est.
-= 0.63 [0.15, 1.11]), higher n-back levels (Est.
-= 0.71 [0.29, 1.13]), grip strength (Est.
-= 0.66 [0.36, 0.95]), and BEAST scores (Est.
-= 0.41 [0.04, 0.78]) regardless of starting nodule sizes (**Figure**).
-This suggests that several independent factors tracking the effects of training, social information use, visuo-spatial working memory, and strength improve an individual's ability to flake in greater quantities.
+The Quantity model roughly paralleled results for Total Production, yielding substantial and significant interactions between training condition, n-back level, BEAST scores, and grip strength.
+As with Total Production, higher visuospatial n-back levels and BEAST scores were associated with lower Quantity scores in the trained group but higher or unchanged Quantity in the untrained group (Figure).
 
-The model produced two significant interactions between training condition and n-back level and BEAST scores.
-Subjects in the trained group with higher n-back levels show lower quantity flaking scores (Est.
-= -1.35 [-1.85, -0.84]) while subjects in the trained group with higher BEAST scores also show lower quantity flaking scores (Est.
-= -0.83 [-1.37, -0.28]) when all the other variables are held constant.
-At first glance, these results appear to contrast with expectations regarding the effects of visuo-spatial working memory and social information use on success in technological tasks.
-However, looking at the matter more holistically it is clear that untrained subjects produce flaking quantities in a different way that relies on visuo-spatial working memory rather than taught information.
-When teaching is provided, being aware of one's surroundings (i.e.
-copying from other novices) can have a negative effect on flaking quantity.
-The interaction between BEAST scores and training shows that social information use acts to level the effects of an individual's propensity to "muscle" through the flaking task (both groups converge on an average quantity flaking score with higher BEAST scores).
-However, the direction of this change is different depending on training with untrained and socially inclined individuals increasing flaking quantities in contrast to trained and socially inclined individuals decreasing flaking quantities.
+Unlike Total Productivity, the effect of Grip Strength on per-core Quantity was mediated by an interaction with learning condition (Figure).
+Thus, high Grip Strength enabled individuals in both groups to produce more total debitage, but only Instructed individuals translated Grip Strength into more intense reduction of individual cores, including not only delta mass, but also number and proportion of larger pieces.
 
-### Model 2: Individual differences and quality flaking
-
-Our second model examining variance in quality flaking follows the same complete model statement we used for the quantity flaking with six covariates.
-From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of 32 (Average AIC = 41).
+### Model 3: Individual differences and quality flaking
+Our third model examining variance in Quality follows the same complete model statement we used for Quantity.
+From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of 32 (Average AIC = 39).
 This model comprised the following statement with three main and four interaction effects:
 
-*Quality flaking* $\sim$ Highest n-back level + Fitt's score + Grip strength + Fitt's score$\cdot$BEAST score + Grip strength$\cdot$BEAST score + Grip strength$\cdot$Fitt's score + Training condition$\cdot$Grip strength + Nodule mass (as control)
+*Quality flaking $\sim$ Highest n-back level  + Fitt’s score + Grip strength + Fitt’s score$\times$BEAST score + Grip strength$\times$BEAST score + Grip strength$\times$Fitt’s score + Training condition$\times$Grip strength + Nodule mass (as control)*
 
-This model explains a statistically significant and substantial proportion of variance in careful flaking (R2 = 0.78, F (8, 12) = 5.5, p \< 0.01, adj.
-R2 = 0.64) in the absence of any main training effects.
-A model evaluation comparing training vs.
-test data subsets shows some evidence for overfitting (Difference in R2 between training and test models = 0.17), which is unfortunately unavoidable with our small sample sizes.
-A model residuals normality test shows no significant differences with the normal distribution (p = 0.13) indicating that this relationship is linear.
-A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 3.4, df = 8, p = 0.9).
+This model explains a statistically significant and substantial proportion of variance in Quality ($R^2$ = 0.75, F (8, 12) = 4.6, p < 0.01, adj.
+$R^2$ = 0.6) in the absence of any main training effects.
+A model residuals normality test shows no significant differences with the normal distribution (p = 0.41) indicating that this relationship is linear.
+A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 7, df = 8, p = 0.5).
 
-\
-**Table** presents this model's coefficients and summary outputs following the same data format as **Table**.
-The results show three significant main effects between our measures of motor-skill, strength, visuo-spatial working memory, and quality flaking.
-A one unit increase in Fitt's score (Est.
-= -0.7 [-1.08, -0.31]), grip strength (Est.
-= -0.9 [-1.32, -0.53]), visuo-spatial working memory (n-back level) (Est.
-= -0.3 [-0.58, -0.07]), and nodule starting mass (Est.
-= -0.3 [-0.58, -0.04]) results in decreased flaking quality regardless of training.
-Collectively, these results show that quality flaking comes at a cost in terms of a subject's motor abilities and that quality flaking decreases as subjects are faced with flaking larger and more challenging nodules.
+The Quality model did produce two statistically (p< 0.05) significant interaction effects (RPM * BEAST & Fitts * n-back).
+However, these interactions had relatively small effects on Quality (<0.5) and we believe that interpreting these results from our small, exploratory study would be inappropriate.
+Figure shows the uneven distribution of data points for these interactions, which suggests vulnerability to leveraging effects of a small number of extreme value combinations.
 
-The quality flaking model shows two significant interaction effects involving grip strength, BEAST scores, and training.
-The interaction between grip strength and BEAST scores has a significant positive effect on careful flaking (Est.
-= 0.5 [0.22, 0.86]).
-**Figure** illustrates this interaction whereby a subject's propensity to use social information mitigates against individual differences in grip strength.
-Strong subjects with a low propensity to use social information tend to perform worst in terms of quality flaking.
-This result makes sense if one considers that our quantity flaking model showed how social information equalizes performance levels in trained and untrained subjects.
-
-The second interaction involves grip strength and training.
-A one unit increase in this interaction results in a significant positive effect on careful flaking (Est.
-= 1 [0.51, 1.47]), but only in the untrained group.
-**Figure** illustrates this interaction showing how training helps even out flaking performance differences between stronger and weaker individuals.
-Stronger individuals perform worse in terms of quality flaking when they are untrained.
-Trained subjects, regardless of grip strength, perform equally well in terms of quality flaking.
-
-### Model 3: Individual differences and quality flaking
+Without a larger sample, it is not possible to determine if these are anomalous outliers or simply represent a poorly sampled part of the broader population.
 
 ## Behavioral observations
-
 We designed this exploratory study primarily to trial experimental design elements such as training time, conditions, and raw materials and to collect preliminary data on the effect of individual differences and training on knapping outcomes.
@@
 # Discussion
-
 The most salient finding of our exploratory study is that the presence/absence of teaching clearly impacted knapping performance but did so in nuanced and individually variable ways that have not been explored in previous studies.

### 2021-08-15 code/PaST_r_code.R

@@ -13,7 +13,7 @@ detachAllPackages <- function() {
 
 detachAllPackages()
 Sys.setenv(JAVA_HOME="")
-Sys.setenv(JAVA_HOME='C:\\Program Files\\Java\\jre1.8.0_301')
+Sys.setenv(JAVA_HOME='C:\\Program Files\\Java\\jre1.8.0_301\\bin\\server')
 library(tidyverse)
 library(car)
 library(reshape2)

### 2021-08-16 manuscript/manuscript.Rmd
add some toy figures and experiment with figure crossreferencing.

@@
    - \linenumbers
+   - \usepackage{float}
+   - \floatplacement{figure}{H}
+   - \usepackage{caption}
+   - \captionsetup[figure]{font=scriptsize}
 linestretch: 1.5
@@
 
-“*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.*”[@montagu1976: 267]
+"*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.*"[@montagu1976: 267]
 
@@
 Studies of simple flake production similarly document large initial variation [@stout2015] and rapid early progress [@stout2015; @putt2019; @stoutKnappingSkillEarliest2006].
-We designed the current study to test the utility of studying learning and variation during the first two hours of simple flaking instruction/practice, in hopes of finding a viable compromise between experimental realism and cost
+We designed the current study to test the utility of studying learning and variation during the first two hours of simple flaking instruction/practice, in hopes of finding a viable compromise between experimental realism and cost.
 
 ## Raw materials and knapping skill
+
+*Such undertakings -- based on raw material which is never standard, and with gestures of percussion that are never perfectly delivered -- cannot be reduced to an elementary repetition of gestures...the realization of elaborate knapping activities necessitates a critical monitoring of the situation and of the decisions adopted all through the process.
+[@pelegrin1990: 117]*
 
@@
 In considering these issues, we again chose to explore a middle path between pragmatism and realism by employing commercially purchased basalt similar to that known from East African Oldowan sites, allowing clast size and shape to vary within set limits, and selecting the particular clasts provided to each participant to approximate the same distribution.
-
-\
 
@@
 ## Stone Tool Making
+
 After individual difference testing, participants engaged in a 2-hour stone tool making session, with a 15-minute break after 1 hour.
@@
 Each group of participants was randomly assigned to one of two experimental conditions: no teaching or teaching.
-In both conditions, participants were first given an opportunity to inspect and handle examples (**Figure**) of the kind of stone tools (flakes) they are being asked to produce.
+In both conditions, participants were first given an opportunity to inspect and handle examples (**Fig.
+\@ref(fig:fig1)**) of the kind of stone tools (flakes) they are being asked to produce.
 They were told that their objective was to produce as many flakes as possible from the materials provided.
@@
 Such a design would also be expected to increase behavioral variability, demanding correspondingly larger samples of participants to identify patterns and making direct comparisons with the taught condition.
+```{r fig1, echo=FALSE, out.width = '75%', fig.align='center', fig.cap="Subjects examining demonstration flakes prior to the experiment.
+The demonstration flakes were made from the same basalt as used in the experiment with the same knapping technique."}
+knitr::include_graphics(here::here("figure/Fig1.PNG"))
+```
 
 ### Raw Materials
+
 Each participant was provided with 9 cores for use over the 2-hour experiment.
@@
 Broken hammerstones were replaced from the reserve to maintain a consistent number and range of choices.
-Each hammerstone was numbered and participants' choices were recorded along with the number of the core(s) being worked on with a particular hammerstone.\
+Each hammerstone was numbered and participants' choices were recorded along with the number of the core(s) being worked on with a particular hammerstone.
 
 ### Experimental Conditions
+
 In both conditions, three researchers were present to record activities and collect materials.
@@
 ## Lithic Analysis
+
 All finished cores were weighed and measured (L, W, T).
@@
 ## Statistical Analyses
+
 To evaluate the association between psychometric, motor-skill, and training measures and technological outcomes, we adopted an information-theoretic approach [@burnham2002].
@@
 
-\
+# Results
 
-# Results
 Following a recent protocol to enhance the reproducibility and data transparency of archaeological research [@marwick2017], detailed results of all analyses and assessments of the data structure are available in our paper's supplementary materials and through Github (<https://github.com/Raylc/PaST-pilot>).
@@
 We were particularly interested in: 1) group level effects of experimental condition, 2) individual differences in aptitude and learning, and 3) potential interactions between learning conditions and individual differences.
-To address these questions, we employed data reduction (Principal Component Analysis) to derive two summary metrics of flaking performance, compared these factors across the two experimental conditions, and built multivariate models examining the relations between our various psychometric measures, subject’s motor skill scores, and our two lithic performance factors.
+To address these questions, we employed data reduction (Principal Component Analysis) to derive two summary metrics of flaking performance, compared these factors across the two experimental conditions, and built multivariate models examining the relations between our various psychometric measures, subject's motor skill scores, and our two lithic performance factors.
 
 ## Principal Component analyses
+
 The following two sections outline factor analyses designed to summarize our main study metrics tracking individual variation in DP sizes and shapes and lithic performance measures.
@@
 ### Detached Piece size and shape
+
 To better understand the relationship between DP shape and training/individual variation, we entered our nine flake linear plan measurements along with maximum flake length and thickness into a principal component analysis (PCA) from which summary coordinates were extracted.
@@
 ### Lithic flaking performance measures
-To better understand the relationship between our various lithic performance measurements and to reduce data dimensionality, we conducted a second principal component analysis examining the study’s six lithic performance measures (count of large pieces [>40mm and 5g], mass of large pieces relative to total detached mass, core delta mass, and the three flake shape factors).
+
+To better understand the relationship between our various lithic performance measurements and to reduce data dimensionality, we conducted a second principal component analysis examining the study's six lithic performance measures (count of large pieces [\>40mm and 5g], mass of large pieces relative to total detached mass, core delta mass, and the three flake shape factors).
 All of these measures were summarized for each core and unique factor scores were calculated from these core-specific measures.
@@
 The analysis yielded two factors explaining a total of 56% of the variance for the entire set of variables (Table).
-Factor 1 (hereafter “Quantity”) explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and on core delta mass.
-Performance factor 2  (hereafter “Quality”) covers 27% of the sample variance and measures flaking quality as reflected in high positive loadings on Shape Factors 1 (size) and 3 (thin, “flake-like” shape) and a negative loading on Shape Factor 2 (“splinter-like” thickness and elongation).
+Factor 1 (hereafter "Quantity") explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and on core delta mass.
+Performance factor 2 (hereafter "Quality") covers 27% of the sample variance and measures flaking quality as reflected in high positive loadings on Shape Factors 1 (size) and 3 (thin, "flake-like" shape) and a negative loading on Shape Factor 2 ("splinter-like" thickness and elongation).
 High scores on Quality thus reflect production of larger, relatively thinner, and more typically flake-shaped vs.
@@
 ## Relationships between Performance Measures
+
 This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (Figures).
@@
 ## Do trained, untrained, and expert knappers perform differently?
+
 Here we compare our flaking outcomes (DP size/shape and flaking performance factors) between the trained and untrained groups.
@@
 All three of these results show either medium or large effect sizes.
-In all three comparisons, the trained group’s data distributions tended towards the expert sample although they were not significantly different from the untrained group (Figures-Core examples too).
-We also observed a significant difference in shape factor 2 (splinters) driven by the expert’s lower values, but with a very low (<0.01) effect size.
+In all three comparisons, the trained group's data distributions tended towards the expert sample although they were not significantly different from the untrained group (Figures-Core examples too).
+We also observed a significant difference in shape factor 2 (splinters) driven by the expert's lower values, but with a very low (\<0.01) effect size.
 These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.
@@
 ## Does performance change over time?
+
 In addition to comparing overall performance during the two hour experiment, we also wanted to determine if groups or individuals differed in learning (i.e., performance change) over the period.
@@
 ## Do individual differences in motor skill and psychometric measures predict flaking performance?
-One of the experiment’s primary goals was to test if measures of individual perceptual-motor and cognitive variation predict success in stone flaking across different training conditions.
+
+One of the experiment's primary goals was to test if measures of individual perceptual-motor and cognitive variation predict success in stone flaking across different training conditions.
 To address this goal, we built three multivariate models examining the relations between training conditions, individual difference measures, and our three lithic performance measures (overall productivity and average per-core Quantity and Quality).
@@
 We considered all possible interactions between five individual difference measures, core size, training condition, and the three performance measures, with each subject providing one data point.
-Each model’s continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitts score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations.
+Each model's continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitts score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations.
 Our two motor skill and strength measures (grip strength and Fitt's performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, $R^2$ = 0.41).
@@
 
-The full models were fitted with the lm function in R 3.2.3, and we used the Glmulti package’s automated model selection algorithm to select the best performing model (lowest AICc score) (see methods for further details on the multimodal selection process).
+The full models were fitted with the lm function in R 3.2.3, and we used the Glmulti package's automated model selection algorithm to select the best performing model (lowest AICc score) (see methods for further details on the multimodal selection process).
 All three models follow the same complete model statement as follows:
 
-*Flaking performance variable $\sim$ Training condition + Highest n-back level + Raven’s Progressive Matrix score + BEAST score + Fitt’s score + Grip strength*
+*Flaking performance variable* $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength
 
@@
 It is thus important to understand whether nodule variability had any impact on our flaking results.
-Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p < 0.01, $R^2$ = 0.54) we included nodule mass as a covariate to control for any variance in flaking performance that may be driven by nodule differences.
+Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p \< 0.01, $R^2$ = 0.54) we included nodule mass as a covariate to control for any variance in flaking performance that may be driven by nodule differences.
 
 ### Model 1: Individual differences and quantity flaking
-Our first model examined variance in overall flaking productivity measured by each subject’s combined flaked mass (nodule starting mass-core final mass).
-This provides a basic measure of variation in individuals’ success detaching pieces and reducing cores from a standardized (see Methods) raw material supply.
+
+Our first model examined variance in overall flaking productivity measured by each subject's combined flaked mass (nodule starting mass-core final mass).
+This provides a basic measure of variation in individuals' success detaching pieces and reducing cores from a standardized (see Methods) raw material supply.
 From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of -18 (Average AIC = -13).
@@
 
-*Total flaked mass $\sim$ Training condition + Grip strength + RPM$\times$Highest n-back level + Fitts score$\times$BEAST score + Grip strength$\times$RPM*
+*Total flaked mass* $\sim$ Training condition + Grip strength + RPM$\times$Highest n-back level + Fitts score$\times$BEAST score + Grip strength$\times$RPM
 
-This model explains a statistically significant and substantial proportion of variance in flaking productivity ($R^2$ = 0.84, F (6, 14) = 12.7, p < 0.01, adj.
+This model explains a statistically significant and substantial proportion of variance in flaking productivity ($R^2$ = 0.84, F (6, 14) = 12.7, p \< 0.01, adj.
 $R^2$ = 0.77).
@@
 
-Table presents this model’s coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average.
+Table presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average.
 The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable.
-We found significant (p< 0.05) and substantial (Standardized Estimate >= 0.50, i.e.
+We found significant (p\< 0.05) and substantial (Standardized Estimate \>= 0.50, i.e.
 a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST.
@@
 ### Model 2: Individual differences and quality flaking
+
 The second full model examined the variance in average flaking Quantity per core.
@@
 
-*Quantity $\sim$ Highest n-back level  + BEAST score + Fitt’s score + Grip strength + Training condition$\times$Highest n-back level + Training condition$\times$BEAST score + Training condition$\times$Grip strength + Nodule mass (as control)*
+*Quantity* $\sim$ Highest n-back level + BEAST score + Fitt's score + Grip strength + Training condition$\times$Highest n-back level + Training condition$\times$BEAST score + Training condition$\times$Grip strength + Nodule mass (as control)
 
@@
 ### Model 3: Individual differences and quality flaking
+
 Our third model examining variance in Quality follows the same complete model statement we used for Quantity.
@@
 
-*Quality flaking $\sim$ Highest n-back level  + Fitt’s score + Grip strength + Fitt’s score$\times$BEAST score + Grip strength$\times$BEAST score + Grip strength$\times$Fitt’s score + Training condition$\times$Grip strength + Nodule mass (as control)*
+*Quality flaking* $\sim$ Highest n-back level + Fitt's score + Grip strength + Fitt's score$\times$BEAST score + Grip strength$\times$BEAST score + Grip strength$\times$Fitt's score + Training condition$\times$Grip strength + Nodule mass (as control)
 
-This model explains a statistically significant and substantial proportion of variance in Quality ($R^2$ = 0.75, F (8, 12) = 4.6, p < 0.01, adj.
+This model explains a statistically significant and substantial proportion of variance in Quality ($R^2$ = 0.75, F (8, 12) = 4.6, p \< 0.01, adj.
 $R^2$ = 0.6) in the absence of any main training effects.
@@
 
-The Quality model did produce two statistically (p< 0.05) significant interaction effects (RPM * BEAST & Fitts * n-back).
-However, these interactions had relatively small effects on Quality (<0.5) and we believe that interpreting these results from our small, exploratory study would be inappropriate.
+The Quality model did produce two statistically (p\< 0.05) significant interaction effects (RPM \* BEAST & Fitts \* n-back).
+However, these interactions had relatively small effects on Quality (\<0.5) and we believe that interpreting these results from our small, exploratory study would be inappropriate.
 Figure shows the uneven distribution of data points for these interactions, which suggests vulnerability to leveraging effects of a small number of extreme value combinations.
@@
 ## Behavioral observations
+
 We designed this exploratory study primarily to trial experimental design elements such as training time, conditions, and raw materials and to collect preliminary data on the effect of individual differences and training on knapping outcomes.
@@
 # Discussion
+
 The most salient finding of our exploratory study is that the presence/absence of teaching clearly impacted knapping performance but did so in nuanced and individually variable ways that have not been explored in previous studies.
@@
 
+# Figures
+
 # References

### 2021-08-16 manuscript/manuscript.Rmd
all figures are updated and inserted properly in the main text.

@@
 It is plausible that these discordant results reflect actual diversity in how humans acquire and master stone tool making skills.
-However, this failure of results to generalize across artificial experimental manipulations [[cf](cf.%20Yarkoni%202020).
+However, this failure of results to generalize across artificial experimental manipulations [cf.
 @yarkoni2020] also raises doubts regarding the external validity [@eren2016] of conclusions with respect to real-world Paleolithic learning contexts.
@@
 Each group of participants was randomly assigned to one of two experimental conditions: no teaching or teaching.
-In both conditions, participants were first given an opportunity to inspect and handle examples (**Fig.
-\@ref(fig:fig1)**) of the kind of stone tools (flakes) they are being asked to produce.
+In both conditions, participants were first given an opportunity to inspect and handle examples (**Figure** \@ref(fig:fig1)) of the kind of stone tools (flakes) they are being asked to produce.
 They were told that their objective was to produce as many flakes as possible from the materials provided.
@@
 Such a design would also be expected to increase behavioral variability, demanding correspondingly larger samples of participants to identify patterns and making direct comparisons with the taught condition.
-```{r fig1, echo=FALSE, out.width = '75%', fig.align='center', fig.cap="Subjects examining demonstration flakes prior to the experiment.
+
+```{r fig1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects examining demonstration flakes prior to the experiment.
 The demonstration flakes were made from the same basalt as used in the experiment with the same knapping technique."}
-knitr::include_graphics(here::here("figure/Fig1.PNG"))
+knitr::include_graphics(here::here("figure/Figure1.jpg"))
 ```
@@
 These cores were produced from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer.
-This produced irregular, angular chunks for use in the experiment, weighing between 459g - 1876g (mean = 975g).
+This basalt has not been mechanically compared [e.g., rebound hardness, @braun2009] to East African basalts, but appears qualitatively similar to finer grained examples from sites around Lake Turkana (D.
+Braun, pers.
+comm.) and at Gona.
+Spalling produced irregular, angular chunks (**Figure** \@ref(fig:fig2)) for use in the experiment, weighing between 459g - 1876g (mean = 975g).
 All cores were weighed, measured (Length, Width, Thickness), and painted white so that new fracture surfaces could be discriminated from those created during production.
@@
 As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length $\times$ Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p = 0.9) of cores provided.
-This was also true comparing the two experimental conditions (Taught vs.
+This was also true for cores provided to participants across the two experimental conditions (Taught vs.
 Untaught mean weight = 1001g vs.
 956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4; mean shape = 221.43 vs.
-221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p = 0.05).
+221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p= 0.05).
 Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores actually used across participants and conditions could still emerge as a result of selection bias.
+
+```{r fig2, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="3D scan of a basalt core prior to knapping.
+Cores were spray painted white to facilitate subsequent technological analyses of the flake scars and flaking intensity.
+Scale in the background is in 10mm blocks."}
+knitr::include_graphics(here::here("figure/Figure2.jpg"))
+```
 
@@
 In both conditions, three researchers were present to record activities and collect materials.
-Participants were seated in a circle [@shea2015] and experiments were video recorded using two cameras (Figure 1).
-Participants were free to select hammerstones from the common pile and to work on any or all of their nine assigned cores in any order they preferred.
+Participants were seated in a circle [@shea2015] and experiments were video recorded using two cameras.
+Participants were free to select hammerstones from the common pile and to work on any or all of their nine assigned cores in any order they preferred (**Figure** \@ref(fig:fig3)).
 However, each core and all associated debitage were collected before participants were allowed to start working on a new core, so it was not possible to partially work and then return to a particular core later.
 The order of cores used and associated hammerstones were recorded for each participant during the experiment.
+
+```{r fig3, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects selecting hammerstones."}
+knitr::include_graphics(here::here("figure/Figure3.jpg"))
+```
 
@@
 Participants were asked to avoid any form of communication about the tool making task specifically, and they complied with this request.
-Participants in this condition thus had the opportunity to observe tool making by an expert and/or by other learners, should they choose to do so, but received no intentional instruction.
+Participants in this condition thus had the opportunity to observe tool making (**Figure** \@ref(fig:fig4)) by an expert and/or by other learners, should they choose to do so, but received no intentional instruction.
+
+```{r fig4, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects observing expert knapper in the untaught condition."}
+knitr::include_graphics(here::here("figure/Figure4.jpg"))
+```
 
@@
 
-This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (Figures).
-As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (Fig.
-Xa), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj.
+This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (**Figure** \@ref(fig:fig5) and **Figure** \@ref(fig:fig6)).
+As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (**Figure** \@ref(fig:fig5)a), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj.
 $R^2$ = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj.
@@
 Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj.
-$R^2$ = 0.7), but no relation in the Trained group (Fig.
-Xb).
+$R^2$ = 0.7), but no relation in the Trained group (**Figure** \@ref(fig:fig5)b).
 A qualitatively similar trend with respect to Quality vs.
-Quantity (Fig.
-Xc) did not achieve significance (F[1, 17] = 0.6, p = 0.2).
+Quantity (**Figure** \@ref(fig:fig5)c) did not achieve significance (F[1, 17] = 0.6, p = 0.2).
+
+```{r fig5, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Relationships between flaking quantity, quality, productivity, and the two training conditions.
+Each dot represents a participant, colors represent training conditions."}
+knitr::include_graphics(here::here("figure/Figure5.jpg"))
+```
 
@@
 Experimental artifacts illustrating these trade-offs are presented in Figure.
+
+```{r fig6, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Comparison of untrained and trained detached pieces illustrating the ability of Trained participants to maximize Quality and Productivity at the same time."}
+knitr::include_graphics(here::here("figure/Figure6.jpg"))
+```
 
@@
 We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors.
-In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (Figure).
+In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (**Figure** \@ref(fig:fig7)).
 All three of these results show either medium or large effect sizes.
-In all three comparisons, the trained group's data distributions tended towards the expert sample although they were not significantly different from the untrained group (Figures-Core examples too).
+In all three comparisons, the trained group's data distributions tended towards the expert sample although they were not significantly different from the untrained group (**Figure** \@ref(fig:fig8)).
 We also observed a significant difference in shape factor 2 (splinters) driven by the expert's lower values, but with a very low (\<0.01) effect size.
@@
 
-While we did not find significant differences in central tendency between our two experimental groups, results (Figure and Figure) did indicate lower variance in the trained group.
+```{r fig7, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Results showing significant differences between the instructor (expert) and novice flaking performance."}
+knitr::include_graphics(here::here("figure/Figure7.png"))
+```
+
+```{r fig8, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Comparison of untaught, taught, and expert cores."}
+knitr::include_graphics(here::here("figure/Figure8.jpg"))
+```
+
+While we did not find significant differences in central tendency between our two experimental groups, results (**Figure** \@ref(fig:fig7)) did indicate lower variance in the trained group.
 To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the trained and untrained individuals using the F-test on either core-averaged or flake specific variances.
-Table and Figure present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass).
+Table and **Figure** \@ref(fig:fig9) present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass).
 In most instances, variance in the untrained group exceeds that of trained individuals by 1.5 to 4.7 times.
-The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the untrained group (Figure), rather than to shift the mean.
+The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the untrained group (**Figure** \@ref(fig:fig9)), rather than to shift the mean.
+
+```{r fig9, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Variance comparisons between trained and untrained individuals across various flaking performance metrics.
+Note the lower variance of the trained group in all the plots."}
+knitr::include_graphics(here::here("figure/Figure9.jpg"))
+```
 
@@
 
-Table shows no significant training effects across the two performance measures either as grouped data or between individuals (Figures).
+Table shows no significant training effects across the two performance measures either as grouped data or between individuals (**Figure** \@ref(fig:fig10) and **Figure** \@ref(fig:fig11)).
 This result demonstrated that flaking outcomes did not change dramatically across the study interval.
@@
 As the smaller nodules in their allotment were depleted, participants were left to knap larger, less preferred nodules.
-This preference for smaller cores is somewhat less pronounced in the untrained group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the untrained group (Figure).
+This preference for smaller cores is somewhat less pronounced in the untrained group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the untrained group (Figure???).
+
+```{r fig10, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Comparisons of nodule starting mass across the study period by training condition.
+Results show a significant relationship between nodule starting mass and length of time in the study."}
+knitr::include_graphics(here::here("figure/Figure10.jpg"))
+```
+
+```{r fig11, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Individual learning curves derived from each subject’s relative core use order and the two flake performance factors."}
+knitr::include_graphics(here::here("figure/Figure11.jpg"))
+```
 
@@
 
-### Model 1: Individual differences and quantity flaking
+### Model 1: Individual differences and overall productivity
 
@@
 a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST.
-The main effect of Grip Strength (Figure), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.
+The main effect of Grip Strength (**Figure** \@ref(fig:fig12)), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.
 
-Effects of visuospatial working memory capacity and social information use are more complicated, as indicated by strong interactions with learning condition (Figures).
+```{r fig12, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Significant main effect of grip strength on overall flaking productivity."}
+knitr::include_graphics(here::here("figure/Figure12.jpeg"))
+```
+
+Effects of visuospatial working memory capacity and social information use are more complicated, as indicated by strong interactions with learning condition (**Figure** \@ref(fig:fig13)).
 In each case, higher scores were associated with better performance in the uninstructed group but worse performance in the instructed group.
@@
 Negative effects in the trained group are unexpected but presumably reflect differences in learning strategies adopted under instruction.
+
+```{r fig13, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Two significant interaction effects of training condition, social information, and visuo-spatial working memory on overall flaking productivity."}
+knitr::include_graphics(here::here("figure/Figure13.png"))
+```
 
@@
 The Quantity model roughly paralleled results for Total Production, yielding substantial and significant interactions between training condition, n-back level, BEAST scores, and grip strength.
-As with Total Production, higher visuospatial n-back levels and BEAST scores were associated with lower Quantity scores in the trained group but higher or unchanged Quantity in the untrained group (Figure).
+As with Total Production, higher visuospatial n-back levels and BEAST scores were associated with lower Quantity scores in the trained group but higher or unchanged Quantity in the untrained group (**Figure** \@ref(fig:fig14)).
 
-Unlike Total Productivity, the effect of Grip Strength on per-core Quantity was mediated by an interaction with learning condition (Figure).
+```{r fig14, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Two significant interaction effects of training condition, visuo-spatial working memory, and social information, on overall flaking quality."}
+knitr::include_graphics(here::here("figure/Figure14.jpg"))
+```
+
+Unlike Total Productivity, the effect of Grip Strength on per-core Quantity was mediated by an interaction with learning condition (**Figure** \@ref(fig:fig15)).
 Thus, high Grip Strength enabled individuals in both groups to produce more total debitage, but only Instructed individuals translated Grip Strength into more intense reduction of individual cores, including not only delta mass, but also number and proportion of larger pieces.
+
+```{r fig15, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Three significant interaction effects of training condition, visuo-spatial working memory and social information, and training and grip strength use on quantity flaking."}
+knitr::include_graphics(here::here("figure/Figure15.jpg"))
+```
 
@@
 However, these interactions had relatively small effects on Quality (\<0.5) and we believe that interpreting these results from our small, exploratory study would be inappropriate.
-Figure shows the uneven distribution of data points for these interactions, which suggests vulnerability to leveraging effects of a small number of extreme value combinations.
+**Figure** \@ref(fig:fig16) shows the uneven distribution of data points for these interactions, which suggests vulnerability to leveraging effects of a small number of extreme value combinations.
+
+```{r fig16, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Significant interactions with low estimated effect sizes from the quality flaking model."}
+knitr::include_graphics(here::here("figure/Figure16.jpg"))
+```
 
@@
 However, even casual behavior observation was sufficient to reveal an unexpected effect.
-Whereas all trained participants copied the general posture and technique of the expert (free hand knapping seated in a chair) fully half (6) of the uninstructed participants experimented with or even knapped all of their cores using the floor as a support (Figure).
+Whereas all trained participants copied the general posture and technique of the expert (free hand knapping seated in a chair) fully half (6) of the uninstructed participants experimented with or even knapped all of their cores using the floor as a support (**Figure** \@ref(fig:fig17)).
 Three of these participants were in the same session, which is also the only group composed of just three individuals.
 In this group, knapping on the ground appears to have been transmitted from one participant to the other two based on appearance order and the point of gaze of participants.
+
+```{r fig17, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Novices in the untaught condition developing and transmitting a form of bipolar technology involving core reduction on the floor."}
+knitr::include_graphics(here::here("figure/Figure17.jpg"))
+```
 
@@
 The absence of this effect in the uninstructed group reflects the weaker association between Total Productivity and per-core Quality across these participants (Fig.Xa) and shows Grip Strength increased uninstructed Productivity specifically by allowing them to knap more cores rather than to reduce individual cores more heavily.
-In keeping with this, uninstructed Grip Strength is positively correlated with Total Cores knapped (r2 = 0.54, p = 0.01).
-Conversely, strength allowed instructed participants to increase their average Quantity per core without affecting the total number of cores knapped (r2 = 0.18, p = 0.165).
+In keeping with this, uninstructed Grip Strength is positively correlated with Total Cores knapped ($R^2$ = 0.54, p = 0.01).
+Conversely, strength allowed instructed participants to increase their average Quantity per core without affecting the total number of cores knapped ($R^2$ = 0.18, p = 0.165).
 Thus, strength appears to have achieved its effects on core reduction rate and intensity in different ways, depending on teaching.

### 2021-08-16 manuscript/manuscript.Rmd
all figures and table added.

@@
    - \captionsetup[figure]{font=scriptsize}
+   - \captionsetup[table]{font=scriptsize}
 linestretch: 1.5
@@
 Of these, 90 were selected as suitable for use.
-These weighed between 213g-1360g (mean = 425) and varied in elongation (L/W = 1.01 to 2.65) and relative thickness (LxW/T = 90.48 to 283.67).
-Forty-five stones were placed in the middle of the knapping area (Figure 2) for participants to freely choose from during the experiment.
+These weighed between 213g-1360g (mean = 425) and varied in elongation (L/W = 1.01 to 2.65) and relative thickness (L$\times$W/T = 90.48 to 283.67).
+Forty-five stones were placed in the middle of the knapping area (**Figure** \@ref(fig:fig3)) for participants to freely choose from during the experiment.
 Broken hammerstones were replaced from the reserve to maintain a consistent number and range of choices.
@@
 
+**Table** \@ref(tab:tab1) lists details for the study’s various lithic attributes.
 All finished cores were weighed and measured (L, W, T).
@@
 By including weight in our cut-off criteria we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.
+```{r tab1, echo=FALSE, message=FALSE}
+library(tidyverse)
+df1 <- data.frame("Variable" = c("Delta mass", "Mass of large detached pieces / total debitage mass", "Length", "Width", "Thickness"), 
+                 "Variable class" = c("Ratio", "Ratio", "Linear measurement", "Linear measurement", "Linear measurement"), 
+                 "Definition" = c("Final core mass as a percentage of original nodule mass.
+Higher values show more completely flaked nodules.", "Combined mass of all detached pieces >40mm in maximum dimension and 5g in mass divided by the mass of all material detached from a specific core.
+Higher values show cores with more detached pieces per knapped material.", "Measurement along each core and flake's longest axis in mm.", "Measurement taken at 10% increments along a detached piece's longest axis starting at the detached piece's tip (in mm).
+Same measurement taken orthogonal to a core's length.", "Measurement taken at the maximum point in a detached piece's profile and orthogonal to width on the cores."),
+                "Recording method" = c("Digital scale", "Calipers, counts, digital scale", "2D photogrammetry", "2D photogrammetry", "Calipers") )
+
+kableExtra::kbl(df1, booktabs = T, caption = "Overview of lithic variables and recording methods used in the study.") %>%
+  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7) %>%
+  kableExtra::column_spec(1, width = "10em") %>%
+  kableExtra::column_spec(3, width = "20em")%>%
+  kableExtra::column_spec(4, width = "10em")
+
+```
 
@@
 
-The analysis yielded three factors explaining a total of 90% of the variance for the entire 11 measurement variable set **(Table**).
+The analysis yielded three factors explaining a total of 90% of the variance for the entire 11 measurement variable set (**Table** \@ref(tab:tab2)).
 Factor 1 tracks flake size with higher scores indicating larger flakes since all 11 measures load positively on this factor.
@@
 We used these three flake shape coordinates to approximate DP size and shape in the project's flake performance factor analysis.
+```{r tab2, echo=FALSE, message=FALSE}
+library(tidyverse)
+df2<-read.csv(here::here("data/tables/table2.csv"), fileEncoding = 'UTF-8-BOM')
+kableExtra::kbl(df2, booktabs = T, caption = "Lithic size/shape PCA factor loadings.") %>%
+  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
+kableExtra::row_spec(0, bold = T)
+```
 
@@
 
-The analysis yielded two factors explaining a total of 56% of the variance for the entire set of variables (Table).
+The analysis yielded two factors explaining a total of 56% of the variance for the entire set of variables (**Table** \@ref(tab:tab3)).
 Factor 1 (hereafter "Quantity") explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and on core delta mass.
@@
 splinter-shaped DPs.
+```{r tab3, echo=FALSE, message=FALSE}
+library(tidyverse)
+df3<-read.csv(here::here("data/tables/table3.csv"), fileEncoding = 'UTF-8-BOM')
+kableExtra::kbl(df3, booktabs = T, caption = "Overview of principal components results on the six flake performance measures. DP = detached piece.") %>%
+  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
+kableExtra::row_spec(0, bold = T)
+```
 
@@
 Thus, it appears that Trained participants achieved higher Total Productivity by increasing average flaking Quantity across cores and without sacrificing Quality, whereas Untrained participants found other ways to vary Total Productivity (e.g., number of cores knapped rather than Quantity per core, see variance Table and Figure) and generally increased productivity at the expense of Quality.
-Experimental artifacts illustrating these trade-offs are presented in Figure.
+Experimental artifacts illustrating these trade-offs are presented in **Figure???**.
 
@@
 
-Table summarizes the results of ANOVA tests group level difference in central tendency on various performance measures.
+**Table** \@ref(tab:tab4) summarizes the results of ANOVA tests group level difference in central tendency on various performance measures.
 We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors.
@@
 These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.
+```{r tab4, echo=FALSE, message=FALSE}
+library(tidyverse)
+df4<-read.csv(here::here("data/tables/table4.csv"), fileEncoding = 'UTF-8-BOM')
+kableExtra::kbl(df4, booktabs = T, caption = "Summary group-wise comparison statistics contrasting flake performance metrics by experiment condition and with the expert instructor.
+All comparisons are three-way except for the two skill PC factors and the total mass removed, which are two-way comparisons between the subject population as these factors were not applied to the expert data sample.
+Eta2 = ANOVA effect size (>0.1 = medium effect size, >0.2 = large effect size).
+* = non-parametric permuted p-values to account for unequal sample variances.
+SS = sum of squared differences, df = degrees of freedom, MS = mean squared difference.") %>%
+  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
+kableExtra::row_spec(0, bold = T)
+```
 
@@
 To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the trained and untrained individuals using the F-test on either core-averaged or flake specific variances.
-Table and **Figure** \@ref(fig:fig9) present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass).
+**Table** \@ref(tab:tab5) and **Figure** \@ref(fig:fig9) present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass).
 In most instances, variance in the untrained group exceeds that of trained individuals by 1.5 to 4.7 times.
 The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the untrained group (**Figure** \@ref(fig:fig9)), rather than to shift the mean.
+```{r tab5, echo=FALSE, message=FALSE}
+library(tidyverse)
+df5<-read.csv(here::here("data/tables/table5.csv"), fileEncoding = 'UTF-8-BOM')
+kableExtra::kbl(df5, booktabs = T, caption = "F-test results comparing variances between trained and untrained subjects across 
+our various flaking performance metrics.") %>%
+  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
+kableExtra::row_spec(0, bold = T)
+```
 
@@
 
-Table shows no significant training effects across the two performance measures either as grouped data or between individuals (**Figure** \@ref(fig:fig10) and **Figure** \@ref(fig:fig11)).
+**Table** \@ref(tab:tab6) shows no significant training effects across the two performance measures either as grouped data or between individuals (**Figure** \@ref(fig:fig10) and **Figure** \@ref(fig:fig11)).
 This result demonstrated that flaking outcomes did not change dramatically across the study interval.
@@
 This preference for smaller cores is somewhat less pronounced in the untrained group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the untrained group (Figure???).
+```{r tab6, echo=FALSE, message=FALSE}
+library(tidyverse)
+df6<-read.csv(here::here("data/tables/table6.csv"), fileEncoding = 'UTF-8-BOM')
+kableExtra::kbl(df6, booktabs = T, caption = "Summary group-wise comparison statistics contrasting flake performance metrics by experiment condition and relative core order.
+All comparisons are three-way.
+Eta2 = ANOVA effect size (>0.1 = medium effect size, >0.2 = large effect size).
+SS = sum of squared differences, df = degrees of freedom, MS = mean squared difference.") %>%
+  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
+kableExtra::row_spec(0, bold = T)
+```
 
@@
 
-*Flaking performance variable* $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength
+*Flaking performance variable $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength*
 
@@
 A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 2, df = 6, p = 0.8).
+```{r tab7, echo=FALSE, message=FALSE}
+library(tidyverse)
+df7<-read.csv(here::here("data/tables/table7.csv"), fileEncoding = 'UTF-8-BOM')
+colnames(df7)[colnames(df7) %in% c("X95..CI", "t..12.")] <- c("95% CI", "t(12)")
+kableExtra::kbl(df7, booktabs = T, caption = "Summary and estimates for the overall flaking productivity model.") %>%
+  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
+kableExtra::row_spec(0, bold = T)
+```
 
-Table presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average.
+**Table** \@ref(tab:tab7) presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average.
 The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable.
@@
 The main effect of Grip Strength (**Figure** \@ref(fig:fig12)), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.
-
 ```{r fig12, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Significant main effect of grip strength on overall flaking productivity."}
@@
 
-*Quantity* $\sim$ Highest n-back level + BEAST score + Fitt's score + Grip strength + Training condition$\times$Highest n-back level + Training condition$\times$BEAST score + Training condition$\times$Grip strength + Nodule mass (as control)
+*Quantity $\sim$ Highest n-back level + BEAST score + Fitt's score + Grip strength + Training condition$\times$Highest n-back level + Training condition$\times$BEAST score + Training condition$\times$Grip strength + Nodule mass (as control)*
 
@@
 A Breusch-Pagan test showed no evidence for heteroskedasticity (whether variance for all observations in our data set are the same) (BP = 4.4, df = 8, p = 0.8).
+
+**Table** \@ref(tab:tab8) presents this model’s coefficients and summary outputs, following the same format as Table. 
+```{r tab8, echo=FALSE, message=FALSE}
+library(tidyverse)
+df8<-read.csv(here::here("data/tables/table8.csv"), fileEncoding = 'UTF-8-BOM')
+colnames(df8)[colnames(df8) %in% c("X95..CI", "t..12.")] <- c("95% CI", "t(12)")
+kableExtra::kbl(df8, booktabs = T, caption = "Summary and estimates for Quantity flaking model.") %>%
+  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
+kableExtra::row_spec(0, bold = T)
+```
 
@@
 
-*Quality flaking* $\sim$ Highest n-back level + Fitt's score + Grip strength + Fitt's score$\times$BEAST score + Grip strength$\times$BEAST score + Grip strength$\times$Fitt's score + Training condition$\times$Grip strength + Nodule mass (as control)
+*Quality flaking $\sim$ Highest n-back level + Fitt's score + Grip strength + Fitt's score$\times$BEAST score + Grip strength$\times$BEAST score + Grip strength$\times$Fitt's score + Training condition$\times$Grip strength + Nodule mass (as control)*
 
@@
 A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 7, df = 8, p = 0.5).
+```{r tab9, echo=FALSE, message=FALSE}
+library(tidyverse)
+df9<-read.csv(here::here("data/tables/table9.csv"), fileEncoding = 'UTF-8-BOM')
+colnames(df9)[colnames(df9) %in% c("X95..CI", "t..12.")] <- c("95% CI", "t(12)")
+kableExtra::kbl(df9, booktabs = T, caption = "Summary and estimates for the Quality flaking model.") %>%
+  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
+kableExtra::row_spec(0, bold = T)
+```
+
+**Table** \@ref(tab:tab9) presents this model’s coefficients and summary outputs following the same data format as **Table** \@ref(tab:tab7) & **Table** \@ref(tab:tab8).
+The results show no effects that were both significant and substantial (Estimate >= 0.5).
 
@@
 
-In our experiment, the strongest effects of teaching were to reduce variance (Figure, Table) rather than shift mean values.
+In our experiment, the strongest effects of teaching were to reduce variance (**Figure** \@ref(fig:fig9), **Table** \@ref(tab:tab5)) rather than shift mean values.
 In particular, teaching acted as a "safety net" that homogenized performance by reducing the frequency of extremely poor outcomes (i.e., learning failures).
@@
 
-# Figures
-
 # References

### 2021-08-16 code/PaST_r_code.R

@@ -10,15 +10,15 @@ detachAllPackages <- function() {
       if (length(package.list)>0)  for (package in package.list) detach(package, character.only=TRUE)
       
 }
-
 detachAllPackages()
-Sys.setenv(JAVA_HOME="")
-Sys.setenv(JAVA_HOME='C:\\Program Files\\Java\\jre1.8.0_301\\bin\\server')
-library(tidyverse)
 library(car)
+library(ggplot2)
 library(reshape2)
+library(tidyr)
 library(plyr)
+library(dplyr)
 library(DescTools)
+library(stringr)
 library(factoextra)
 library(psych) 
 library(Hmisc)
@@ -29,7 +29,6 @@ library(sjPlot)
 library(rcompanion)
 library(caret)
 library(permuco)
-
 
 #### Import datasets ####
 
@@ -229,13 +228,6 @@ subject_core_flakes <- pca_flakes[, 1:4]
 
 pca_flakes_complete<-prcomp(pca_flakes[,c(15,24,26:34)], scale. = T)
 
-# Biplot of variables and factors
-
-fviz_pca_var(pca_flakes_complete,
-             col.ind = "cos2", # Color by the quality of representation
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
-             repel = TRUE)
-
 # Eigenvalues
 
 eig.val_flakes <- get_eigenvalue(pca_flakes_complete)
@@ -263,30 +255,6 @@ complete_past_data_shape_pca<-join(complete_past_data_plusflakes,
                                    joined_to_identifiers,
                                 by=c("subject","Condition","core_number","flake_number")) %>%
   mutate(flake_pca_dim1=flake_pca_dim1*-1) #change sign so larger values = larger flakes
-
-# plot subjects' data against absolute size measure to track allometry in variance. 
-
-ggplot(complete_past_data_shape_pca, aes(y = log(flake_mass), x = flake_pca_dim2)) +
-      geom_point(alpha = 0.5) +
-      #geom_smooth(method='loess') +
-      theme_classic() +
-      ylab("Flake mass (log)") +
-      xlab("Flake shape Factor 2 (thickness relative to width)")+
-  theme(text = element_text(size=20))
-
-ggplot(complete_past_data_shape_pca, aes(y = log(flake_mass), x = flake_pca_dim3)) +
-      geom_point(alpha = 0.5) +
[…]

### 2021-08-17 manuscript/manuscript.Rmd
figures remade using R.

@@
 4)  Fluid intelligence [@cattell1963] refers to the capacity to engage in abstract reasoning and problem solving in a way that is minimally dependent on prior experience.
-It complements "crystallized intelligence" (the ability to apply learned procedures and knowledge) as one of the two factors (gf, gc) comprising so-called "general intelligence" (g).
-Fluid intelligence is closely related to the executive control of attention and manipulation of information held in working memory [@engle2018](Engle 2018).
+It complements "crystallized intelligence" (the ability to apply learned procedures and knowledge) as one of the two factors ($g_f$, $g_c$) comprising so-called "general intelligence" (g).
+Fluid intelligence is closely related to the executive control of attention and manipulation of information held in working memory [@engle2018].
 It is hypothesized to support technological innovation [@coolidge2005] and/or the intentional learning of new skills [@unsworth2005; @stout2015].
@@
 df5<-read.csv(here::here("data/tables/table5.csv"), fileEncoding = 'UTF-8-BOM')
+colnames(df5)[colnames(df5) %in% c("X95..CI")] <- c("95% CI")
 kableExtra::kbl(df5, booktabs = T, caption = "F-test results comparing variances between trained and untrained subjects across 
@@
 ```{r fig12, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Significant main effect of grip strength on overall flaking productivity."}
-knitr::include_graphics(here::here("figure/Figure12.jpeg"))
+knitr::include_graphics(here::here("figure/Figure12.jpg"))
 ```

### 2021-08-17 code/PaST_r_code.R

@@ -29,6 +29,7 @@ library(sjPlot)
 library(rcompanion)
 library(caret)
 library(permuco)
+library(patchwork)
 
 #### Import datasets ####
 
@@ -403,15 +404,6 @@ subject_skillPC_data<-complete_past_data_shape_skill_pca_combined %>%
   select(c(subject, Condition, core_number, flake_number, skill_pca_dim1_quantity_flaking, skill_pca_dim2_quality_flaking)) %>%
   na.omit(skill_pca_dim1_quantity_flaking)
 
-ggplot(subset(subject_skillPC_data), 
-       aes(x=skill_pca_dim1_quantity_flaking, y=skill_pca_dim2_quality_flaking,color=Condition)) +
-      geom_point() +
-      geom_smooth(method='lm', se=T)+ 
-      scale_color_manual(name = "Training condition", labels = c("Untrained", "Trained"), values = c("blue", "red"))+
-      theme_classic() +
-      ylab("Factor 1 'Quality flaking' (higher = higher quality flaking)") +
-      xlab("Factor 2 'Quantity flaking' (higher = higher quantity flaking)")+
-      theme(text = element_text(size=16))
 
 summary(lm(skill_pca_dim1_quantity_flaking~skill_pca_dim2_quality_flaking+Condition, data=subject_skillPC_data))
 summary(lm(skill_pca_dim1_quantity_flaking~skill_pca_dim2_quality_flaking, data=subset(subject_skillPC_data,Condition==1 &skill_pca_dim2_quality_flaking>-1)))
@@ -440,7 +432,9 @@ var.test(skill_pca_dim1_quantity_flaking ~ Condition, complete_past_data_shape_s
 
 var.test(skill_pca_dim2_quality_flaking ~ Condition, complete_past_data_shape_skill_pca_combined_bycore, 
          alternative = "two.sided")
-ggplot(filter(complete_past_data_shape_skill_pca_combined_bycore,!subject=="demo"), 
+
+##Figure 9b
+p9b<-ggplot(filter(complete_past_data_shape_skill_pca_combined_bycore,!subject=="demo"), 
        aes(x=Condition, y=skill_pca_dim2_quality_flaking, color=Condition)) +
   geom_boxplot(outlier.size=0)+
   geom_point(size=3,position = position_jitterdodge()) +
@@ -448,7 +442,7 @@ ggplot(filter(complete_past_data_shape_skill_pca_combined_bycore,!subject=="demo
[…]

### 2021-08-18 manuscript/manuscript.Rmd
replace the figure 14.

@@
 ```{r fig7, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Results showing significant differences between the instructor (expert) and novice flaking performance."}
-knitr::include_graphics(here::here("figure/Figure7.png"))
+knitr::include_graphics(here::here("figure/Figure7.jpg"))
 ```
@@
 ```{r fig13, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Two significant interaction effects of training condition, social information, and visuo-spatial working memory on overall flaking productivity."}
-knitr::include_graphics(here::here("figure/Figure13.png"))
+knitr::include_graphics(here::here("figure/Figure13.jpg"))
 ```

### 2021-08-18 code/PaST_r_code.R

@@ -809,12 +809,12 @@ pred.mm_cond_nback<- ggpredict(PC1_best_model, terms = c("Highest_nBack","Condit
 levels(pred.mm_cond_nback$group) <- c("Untrained", "Trained")
 ##Fig 14a
 p14a<- plot(pred.mm_cond_nback,facet=F, add.data = F,show.title =F, alpha=0.07) + 
-  labs(title ="b",x = "Highest n-back level", 
+  labs(title ="a",x = "Highest n-back level", 
        y = "Quantity flaking (higher = more quantity flaking)")+
   theme_classic()+
   theme(text = element_text(size=20))
 
-.mm_beast_condition<- ggpredict(PC1_best_model, terms = c("BEAST","Condition"))  # this gives overall predictions for the model
+pred.mm_beast_condition<- ggpredict(PC1_best_model, terms = c("BEAST","Condition"))  # this gives overall predictions for the model
 levels(pred.mm_beast_condition$group) <- c("Untrained", "Trained")
 ##Fig 14b
 
@@ -1029,3 +1029,23 @@ summary(lm(skill_pca_dim2_quality_flaking~removed_mass_sum, data=model_data))
 summary(lm(skill_pca_dim2_quality_flaking~removed_mass_sum, data=subset(model_data, Condition=="0")))
 summary(lm(skill_pca_dim2_quality_flaking~removed_mass_sum, data=subset(model_data, Condition=="1")))
 
+## Fig 12
+pred.mm_grip<- ggpredict(quantity_best_model, terms = c("Grip_Strength"))  # this gives overall predictions for the model
+autoplot(pred.mm_grip) +
+  # Expected + CI
+  geom_CI_ribbon() +
+  geom_expected_line()+
+  layer_fit_data(size=3, color = "black")+
+  labs(x = "Grip strength",
+       y = "Total mass removed from cores")+
+  theme_classic()+
+  theme(text = element_text(size=20))
+
+
+## Supplementary tables
+## 1.lithic performance (total cores used, 
+## total flake mass/flaked core mass; 
+## total flaked mass; total flakes >40mm & 5 g)
+
+
+## 2. 

### 2021-08-18 manuscript/SI.Rmd
supplementary data created.

---
title: Supplementary Data for Testing the motor and cognitive foundations of Paleolithic social transmission
author: 
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
- Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Aditi Majoe^[Indepedent Researcher; am2752@cantab.ac.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---
```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1<-read.csv(here::here("data/tables/s1.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df1, booktabs = T, caption = "Lithic performance between two groups.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

```{r tab2, echo=FALSE, message=FALSE}
library(tidyverse)
df2<-read.csv(here::here("data/tables/s2.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df2, booktabs = T, caption = "Flake metrics between two groups.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

```{r tab3, echo=FALSE, message=FALSE}
library(tidyverse)
df3<-read.csv(here::here("data/tables/s3.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df3, booktabs = T, caption = "Starting nodule metrics between two groups.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

### 2021-08-18 manuscript/SI.tex
supplementary data created.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Supplementary Data for Testing the motor and cognitive foundations of Paleolithic social transmission},
  pdfauthor={Justin Pargeter; Megan Beney Kilgore; Cheng Liu; Aditi Majoe; Dietrich Stout},
  colorlinks=true,
  linkcolor=blue,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=Blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{textcomp}
\usepackage{utopia}
\usepackage{lineno}
\linenumbers
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{caption}
\captionsetup[figure]{font=scriptsize}
\captionsetup[table]{font=scriptsize}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Supplementary Data for Testing the motor and cognitive foundations of Paleolithic social transmission}
\author{Justin Pargeter\footnote{Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; \href{mailto:justin.pargeter@nyu.edu}{\nolinkurl{justin.pargeter@nyu.edu}}} \and Megan Beney Kilgore\footnote{Department of Anthropology, Emory University, Atlanta, GA, USA; \href{mailto:megan.elizabeth.beney@emory.edu}{\nolinkurl{megan.elizabeth.beney@emory.edu}}} \and Cheng Liu\footnote{Department of Anthropology, Emory University, Atlanta, GA, USA; \href{mailto:raylc1996@outlook.com}{\nolinkurl{raylc1996@outlook.com}}} \and Aditi Majoe\footnote{Indepedent Researcher; \href{mailto:am2752@cantab.ac.uk}{\nolinkurl{am2752@cantab.ac.uk}}} \and Dietrich Stout\footnote{Department of Anthropology, Emory University, Atlanta, GA, USA; \href{mailto:dwstout@emory.edu}{\nolinkurl{dwstout@emory.edu}}}}
\date{}

\begin{document}
\maketitle

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\setstretch{1.5}
\begin{verbatim}
## Warning in read.table(file = file, header = header, sep = sep, quote = quote, :
## invalid input found on input connection 'C:/FilesVC/PaST-pilot/data/tables/
## s1.csv'
\end{verbatim}

\begin{verbatim}
## Warning in read.table(file = file, header = header, sep = sep, quote = quote, :
## line 1 appears to contain embedded nulls
\end{verbatim}

\begin{verbatim}
## Warning in read.table(file = file, header = header, sep = sep, quote = quote, :
## incomplete final line found by readTableHeader on 'C:/FilesVC/PaST-pilot/data/
## tables/s1.csv'
\end{verbatim}

\begin{table}[H]

\caption{\label{tab:tab1}Lithic performance between two groups.}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{l}
\toprule
\cellcolor{gray!6}{PK...}\\

\bottomrule
\end{tabular}
\end{table}

\begin{verbatim}
## Warning in read.table(file = file, header = header, sep = sep, quote = quote, :
## invalid input found on input connection 'C:/FilesVC/PaST-pilot/data/tables/
## s2.csv'
\end{verbatim}

\begin{verbatim}
## Warning in read.table(file = file, header = header, sep = sep, quote = quote, :
## line 1 appears to contain embedded nulls
\end{verbatim}

\begin{verbatim}
## Warning in read.table(file = file, header = header, sep = sep, quote = quote, :
## incomplete final line found by readTableHeader on 'C:/FilesVC/PaST-pilot/data/
## tables/s2.csv'
\end{verbatim}

\begin{table}[H]

\caption{\label{tab:tab2}Flake metrics between two groups.}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{l}
\toprule
\cellcolor{gray!6}{PK...}\\

\bottomrule
\end{tabular}
\end{table}

\begin{verbatim}
## Warning in read.table(file = file, header = header, sep = sep, quote = quote, :
## invalid input found on input connection 'C:/FilesVC/PaST-pilot/data/tables/
## s3.csv'
\end{verbatim}

\begin{verbatim}
## Warning in read.table(file = file, header = header, sep = sep, quote = quote, :
## line 1 appears to contain embedded nulls
\end{verbatim}

\begin{verbatim}
## Warning in read.table(file = file, header = header, sep = sep, quote = quote, :
## incomplete final line found by readTableHeader on 'C:/FilesVC/PaST-pilot/data/
## tables/s3.csv'
\end{verbatim}

\begin{table}[H]

\caption{\label{tab:tab3}Starting nodule metrics between two groups.}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{l}
\toprule
\cellcolor{gray!6}{PK...}\\

\bottomrule
\end{tabular}
\end{table}

\end{document}

### 2021-08-18 manuscript/manuscript.Rmd
supplementary data created.

@@
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
+- Aditi Majoe^[Indepedent Researcher; am2752@cantab.ac.uk]
 - Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
@@
 
-"*The many slight differences...
+|   "*The many slight differences...
 being observed in the individuals of the same species inhabiting the same confined locality, may be called individual differences...
@@
 
-"*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.*"[@montagu1976: 267]
+|   "*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.*"[@montagu1976: 267]
 
@@
 
-*Such undertakings -- based on raw material which is never standard, and with gestures of percussion that are never perfectly delivered -- cannot be reduced to an elementary repetition of gestures...the realization of elaborate knapping activities necessitates a critical monitoring of the situation and of the decisions adopted all through the process.
-[@pelegrin1990: 117]*
+|   "*Such undertakings -- based on raw material which is never standard, and with gestures of percussion that are never perfectly delivered -- cannot be reduced to an elementary repetition of gestures...the realization of elaborate knapping activities necessitates a critical monitoring of the situation and of the decisions adopted all through the process." [@pelegrin1990: 117]*
 

### 2021-08-18 code/PaST_r_code.R

@@ -29,6 +29,7 @@ library(sjPlot)
 library(rcompanion)
 library(caret)
 library(permuco)
+library(ggggeffects)
 library(patchwork)
 
 #### Import datasets ####
@@ -512,7 +513,7 @@ p9d<-ggplot(filter(total_mass_data,!subject=="demo"),
   ylab("Total mass removed from cores")+
   theme(text = element_text(size=20))
 (p9a|p9b)/(p9c|p9d)
-ggsave("figure/Figure9.jpg",width = 40, height = 40, units = "cm", dpi=300)
+## ggsave("figure/Figure9.jpg",width = 40, height = 40, units = "cm", dpi=300)
 
 ## Skill factor comparisons between novice groups
 leveneTest(skill_pca_dim1_quantity_flaking ~ Condition, data = subject_skillPC_data)
@@ -576,7 +577,7 @@ p7b<-ggplot(complete_past_data_shape_skill_pca_combined_bycore, aes(x=condition_
   ylab("Core delta mass")+
   theme(text = element_text(size=20))
 p7a|p7b
-ggsave("figure/Figure7.jpg",width = 40, height = 20, units = "cm", dpi=300)
+## ggsave("figure/Figure7.jpg",width = 40, height = 20, units = "cm", dpi=300)
 
 leveneTest(delta_mass ~ condition_threeway, data = complete_past_data_shape_skill_pca_combined_bycore)
 aovperm(delta_mass ~ condition_threeway,
@@ -709,7 +710,7 @@ p11b<- ggplot(learning_rate_data,aes(x=adjusted_cores_used,y=skill_pca_dim2_qual
   theme(text = element_text(size=14))+
   theme(legend.position = "NULL")
 p11a|p11b
-ggsave("figure/Figure11.jpg",width = 50, height = 25, units = "cm", dpi=300)
+## ggsave("figure/Figure11.jpg",width = 50, height = 25, units = "cm", dpi=300)
 
 aov(log(core_start_mass)~Condition*core_order_stage, data=learning_rate_data) %>% report() %>% as.report_table(summary=TRUE)
 
@@ -723,7 +724,7 @@ ggplot(learning_rate_data,aes(x=Order,y=log(core_start_mass), color=Condition))+
   scale_x_continuous(breaks=seq(0,9,1))+
   theme_classic()+
   theme(text = element_text(size=20))
[…]

### 2021-08-20 code/PaST_r_code.R
Update PaST_r_code.R
I've made two small changes that correct the DF in our group-level comparisons bringing them in line with our subject numbers reported in the paper. I've updated the group-level comparison tables we'll submit with these details (https://www.dropbox.com/sh/ybr7u1612z5dov6/AABy4OR-Pi-SJMJb2j5B2_Ada?dl=0).

@@ -516,9 +516,15 @@ p9d<-ggplot(filter(total_mass_data,!subject=="demo"),
 ## ggsave("figure/Figure9.jpg",width = 40, height = 40, units = "cm", dpi=300)
 
 ## Skill factor comparisons between novice groups
+
+subject_skillPC_data<-complete_past_data_shape_skill_pca_combined_bycore %>%
+  subset(!subject=="demo") %>%
+  distinct(subject, .keep_all=T) %>%
+  select(c(subject,Condition,skill_pca_dim1_quantity_flaking,skill_pca_dim2_quality_flaking))
+
 leveneTest(skill_pca_dim1_quantity_flaking ~ Condition, data = subject_skillPC_data)
-aovperm(skill_pca_dim1_quantity_flaking ~ Condition,
-        data = subject_skillPC_data)
+summary(aov(skill_pca_dim1_quantity_flaking ~ Condition,
+        data = subject_skillPC_data))
 
 leveneTest(skill_pca_dim2_quality_flaking ~ Condition, data = filter(complete_past_data_shape_skill_pca_combined_bycore,!subject=="demo" ))
 aovperm(skill_pca_dim2_quality_flaking ~ Condition,

### 2021-08-21 manuscript/manuscript.Rmd
updated the text partially.

@@
 ---
-title: Testing the motor and cognitive foundations of Paleolithic social transmission
+title: Testing the social, cognitive, and motor foundations of Paleolithic skill reproduction
 author: 
 - Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
+- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
 - Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
-- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
 - Aditi Majoe^[Indepedent Researcher; am2752@cantab.ac.uk]
 - Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
-abstract: "Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret.
-Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability.
-However, we actually know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic.
-This paper presents results from a pilot experimental study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition.
+abstract: "Stone tools provide key evidence of human cognitive evolution but remain challenging to interpret.
+Tool-making skill-learning has been understudied even though: 1) the most salient cognitive demands of tool-making should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool-making ability.
+However, we know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic.
+This paper presents results from an exploratory study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool-making skill acquisition.
 We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs.
-direct active teaching) employing appropriate raw materials, practice time, and real human interaction.
-Participant performance was evaluated through analysis of the stone artifacts produced.
-Performance was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study.
-Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group.
-These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.\\par \\textbf{Keywords:} Oldowan; Stone toolmaking; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
+direct active teaching).
+We employed appropriate raw materials, a moderate practice time duration, and in-person, fully interactive instruction.
+Participant performance was evaluated through an analysis of the stone artifacts produced.
+We compared performance across experimental groups with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study.
+Our results show that 2 hours are insufficient to document learning-related performance change.
+However, teaching reduces variability in knapping rate, methods, and outcomes during early-stage learning, thus increasing the reliability of skill reproduction.
+Instruction also altered knapping quality vs.
+quantity trade-offs in the two groups and dramatically changed the effects of individual differences in strength, visuospatial working memory, and social learning tendencies on knapping outcomes.
+Our results provide further support for the hypothetical co-evolution of teaching, language, and tool-making, suggest that the presence/absence of instruction can fundamentally alter learning-related selection pressures on individuals, and offer lessons for the design of future experiments.\\par \\textbf{Keywords:} Oldowan; Stone tool-making; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
 fontsize: 11pt
@@
 Pathbreaking attempts to infer specific cognitive capacities from this evidence largely focused on the basic requirements of tool production [@isaac1976; @wynn1979; @gowlett1984; @wynn2004].
-More recently, increasing attention has been directed to the processes and demands of stone tool making skill acquisition [@roux1995; @stout2002; @stout2005; @geribàs2010; @nonaka2010; @stout2011; @putt2014; @hecht2015a; @duke2015; @morgan2015; @stout2015; @lombao2017; @putt2017; @cataldo2018; @putt2019; @pargeter2019; @pargeter2020].
-This is motivated by the expectation that the most salient cognitive demands of tool making should occur during learning rather than routine expert performance [@stout2015] and by interest in the relevance of different social learning mechanisms such as imitation [@rein2014; @stout2019], emulation [@tehrani2008; @wilkins2018], and language [@ohnuma1997; @putt2014; @morgan2015; @lombao2017; @putt2017; @cataldo2018] to the reproduction of Paleolithic technologies.
+More recently, researchers have directed increasing attention to the processes and demands of stone tool-making skill acquisition [@roux1995; @stout2002; @stout2005; @geribàs2010; @nonaka2010; @stout2011; @putt2014; @hecht2015a; @duke2015; @morgan2015; @stout2015; @lombao2017; @putt2017; @cataldo2018; @putt2019; @pargeter2019; @pargeter2020; @shipton2018].
+This increased attention is motivated by the expectation that tool-making's most salient cognitive demands should occur during learning rather than routine expert performance [@stout2015].
+Researchers are also increasingly focused on the relevance of different social learning mechanisms such as imitation [@rein2014; @stout2019], emulation [@tehrani2008; @wilkins2018], and language [@ohnuma1997; @putt2014; @morgan2015; @lombao2017; @putt2017; @cataldo2018] to the reproduction of Paleolithic technologies.
 
 Studies investigating these questions have used a range of different experimental designs (e.g., varying technological goals/instructions, training times, raw materials, live vs.
-recorded instruction, lithic/skill assessment metrics, pseudo-knapping tasks etc.) and reached disparate conclusions regarding the neurocognitive and social foundations of skill acquisition.
-It is plausible that these discordant results reflect actual diversity in how humans acquire and master stone tool making skills.
+recorded instruction, lithic/skill assessment metrics, pseudo-knapping tasks etc.) and reached disparate conclusions regarding skill acquisition’s neurocognitive and social foundations.
+It is plausible that these discordant results reflect actual diversity in how humans acquire and master stone tool-making skills.
 However, this failure of results to generalize across artificial experimental manipulations [cf.
-@yarkoni2020] also raises doubts regarding the external validity [@eren2016] of conclusions with respect to real-world Paleolithic learning contexts.
-To address this, we conducted an exploratory study that draws on lessons from previous research in an attempt to balance the pragmatic and theoretical tradeoffs inherent in experimental studies of stone knapping skill acquisition [@pargeter2019; @stout2015].
+@yarkoni2020] also raises doubts regarding the external validity [@eren2016] of conclusions concerning real-world Paleolithic learning contexts.
+To address this, we conducted an exploratory study that draws on lessons from previous research to balance the pragmatic and theoretical trade-offs inherent in experimental studies of stone knapping skill acquisition [@pargeter2019; @stout2015].
 
-Learning real-world skills like stone knapping is highly demanding of time and materials and difficult to control experimentally without sacrificing generalizability to real world conditions.
+Compared to artificial experimental tasks, learning real-world skills like stone knapping is highly demanding of time and materials and difficult to control experimentally without sacrificing generalizability to real world conditions.
 Prior efforts have attempted to navigate these challenges by using various combinations of 1) inauthentic raw materials that are less expensive, easier to standardize, and/or easier to knap, 2) video-recorded instruction that is uniform across participants and less demanding of experimenter time, 3) short learning periods, 4) small sample sizes, and 5) single learning conditions.
@@
 instruction, motivation) and assess individual variation (e.g.
-performance, psychometrics, neuroanatomy) that *also* have realistic learning periods." The current study attempts to strike a viable balance between these demands by investigating early-stage learning of a relatively simple technology (least effort, "Oldowan," flake production [@reti2016; @shea2016] under two instructional conditions while collecting data on individual differences in strength, coordination, cognition, social learning, self-control, and task engagement.
-Unlike any previous study, this allows us to address the likelihood that group effects of training conditions might be impacted by interactions with individual participant differences in aptitude, motivation, or learning style.
+performance, psychometrics, neuroanatomy) that *also* have realistic learning periods." The current study attempts to strike a viable balance between these demands by investigating early-stage learning of a relatively simple technology (least effort, "Oldowan," flake production [@reti2016; @shea2016]) under two instructional conditions while collecting data on individual differences in strength, coordination, cognition, and social learning.
+Unlike any previous study, this allows us to address the likelihood that interactions with individual participant differences in aptitude or learning style might impact group effects of training conditions.
 
-We focus on early stage learning because it has been found to be relatively rapid, variable across individuals, and predictive of later outcomes [@pargeter2019; @stout2015; @putt2019], and thus provides a reasonable expectation of generating meaningful data on skill and learning variation while minimizing training costs.
+We focus on early-stage learning because it is relatively rapid, variable across individuals, and predictive of later outcomes [@pargeter2019; @stout2015; @putt2019].
+The study thus provides a reasonable expectation of generating meaningful data on skill and learning variation while minimizing training costs.
 Moreover, understanding the minimum training times necessary to detect changes in tool making skill will help archaeologists design more realistic and cost-effective experiments.
-To further manage costs, we limited our study to only two learning conditions (observation only vs.
-active teaching).
-This targets a key controversy in human evolution, namely the origins of teaching and language [@gärdenfors2017; @morgan2015], while avoiding highly artificial manipulations of dubious relevance to real-world Paleolithic learning.
-These choices allowed us to invest more in other aspects of research design that we identified as theoretically important, including measurement of individual differences in cognition and behavior, inclusion of an in-person, fully interactive teaching condition, and use of naturalistic raw materials.
-Sample size remained small in this internally funded exploratory study but could easily be scaled up at funding levels typical of pre- and post-doctoral research grants in archaeology.
+We limited our study to only two learning conditions (observation only vs.
+active teaching) to further manage costs.
+This study design targets a key controversy in human evolution, namely the origins of teaching and language [@gärdenfors2017; @morgan2015] while avoiding artificial manipulations of dubious relevance to real-world Paleolithic learning [e.g., @cataldo2018].
+These choices allowed us to invest more in other aspects of research design that we identified as theoretically necessary, including measuring individual differences in cognition and behavior, including an in-person, fully interactive teaching condition, and using naturalistic raw materials.
+The sample size remained small in this internally funded exploratory study, but we could easily scale it up at funding levels typical of pre- and post-doctoral research grants in archaeology.
 
@@
 
-Individuals vary in aptitude and learning style for particular skills [@jonassen1993] but this has largely been ignored in studies of knapping skill acquisition, which have instead focused on group effects of different experimental conditions.
-There are good pragmatic reasons for this, as individual difference studies typically require larger sample sizes and additional data collection.
-However, overlooking these distinctions is not ideal since individual differences can provide valuable insight into the mechanisms, development, and evolution of cognition and behavior [@boogert2018].
-In particular, patterns of association between cognitive traits and behavioral performance can be used to test hypotheses about the cognitive demands of learning particular skills and the likely targets of natural selection acting on aptitude.
-More prosaically, individual differences can introduce an unexamined and uncontrolled source of variation in group level results.
-This is especially true in the relatively small "samples of convenience" typical of experimental archaeology.
+Individuals vary in aptitude and learning style for skills [@jonassen1993], but research on knapping skill acquisition has largely ignored this.
+These studies have instead focused on group effects of different experimental conditions.
+There are good pragmatic reasons for this, as individual difference studies typically require larger sample sizes and other data collection.
+However, overlooking these distinctions is not ideal since individual differences can provide valuable insight into cognition and behavior's mechanisms, development, and evolution [@boogert2018].
+Researchers can use patterns of association between cognitive traits and behavioral performance to test hypotheses about the cognitive demands of learning particular skills and the likely targets of natural selection acting on aptitude.
+More prosaically, individual differences can introduce an unexamined and uncontrolled source of variation in group-level results.
+This effect is especially true in the relatively small samples of convenience typical of experimental archaeology.
 
-While testing hypotheses in evolutionary cognitive archaeology remains a considerable challenge [@wynn2017], investigation of individual variation in modern research participants represents one promising direction.
-For any particular behavior of archaeological interest, it is expected that standing variation in modern populations should remain relevant to normal variation in learning aptitude.
+While testing hypotheses in evolutionary cognitive archaeology remains a considerable challenge [@wynn2017], investigating individual variation in modern research participants represents one promising direction.
+For any behavior of archaeological interest, we could expect that standing variation in modern populations should remain relevant to normal variation in learning aptitude.
 The presence of trait variation without impact on learning aptitude would provide strong evidence against the plausibility of the proposed evolutionary relationship.
-An absence of variation (i.e., past fixation and rigorous developmental canalization) is not expected given the known variability of human brains and cognition [@sherwood2017; @barrett2020].
-Any confirmatory findings of trait-aptitude correspondence would then have the testable implication that humans should be evolutionarily derived along the same dimension [e.g.
-@hecht2015b].
+We would not expect an an absence of variation (i.e., past fixation and rigorous developmental canalization) given the known variability of human brains and cognition [@sherwood2017; @barrett2020].
+Any confirmatory findings of trait-aptitude correspondence would have the testable implication that humans should be evolutionarily derived along the same dimension [@hecht2015b].
 
@@
 Hecht et al.
-[-@hecht2015a] reported training-related changes in white matter integrity (fractional anisotropy [FA]) that correlated with individual differences in practice time and striking accuracy change.
-The regional patterning of FA changes also varied across individuals, with only those individuals who displayed early increases in FA under the right ventral precentral gyrus (premotor cortex involved in movement planning and guidance) showing striking accuracy improvement over the training period.
+[-@hecht2015a] reported training-related changes in white matter structure correlated with individual differences in practice time and striking accuracy change.
+The regional patterning of white matter changes also varied across individuals.
+Only those who displayed early increases under the right ventral precentral gyrus (premotor cortex involved in movement planning and guidance) showed striking accuracy improvement over the training period.
 Putt et al.
-[-@putt2019] similarly found that the proportion of flakes to shatter produced by individuals during handaxe making correlated with dorsal precentral gyrus (motor cortex) activation.
+[-@putt2019] similarly found that the proportion of flakes-to-shatter produced by individuals during handaxe-making correlated with dorsal precentral gyrus (motor cortex) activation.
 Pargeter et al.
 [-@pargeter2020] used a flake prediction paradigm [modeled after @nonaka2010] to confirm that striking force and accuracy are important determinants of handaxe-making success.
-These findings all point to the central role of perceptual-motor systems [@stout2007] and coordination [@roux1995] in knapping skill acquisition.
+These findings point to the central role of perceptual-motor systems [@stout2007] and coordination [@roux1995] in knapping skill acquisition.
 In addition, Putt et al.
-[-@putt2019] also found successful flake production to be associated with prefrontal (working memory/cognitive control) activation and Stout et al.
+[-@putt2019] found successful flake production to be associated with prefrontal (working memory/cognitive control) activation, and Stout et al.
 [-@stoutCognitiveDemandsLower2015] found that prefrontal activation correlated with success at a strategic judgement (platform selection) task which in turn was predictive of success at out-of-scanner handaxe production.
-Such investigations are thus starting to chart out the more specific contributions of different neural systems to particular aspects of knapping skill acquisition.
-To date, however, the cognitive/functional interpretation of systems identified in this manner has largely relied on informal reverse inference (reasoning backward from observed activations to inferred mental processes) from published studies of other tasks that activated the same regions, an approach which is widely regarded as problematic [@poldrack2011].
 
+Such investigations are thus starting to chart the contributions of different neural systems to aspects of knapping skill acquisition, including motor and cognitive control aspects.
+To date, however, the cognitive/functional interpretation of systems identified in this manner has relied mainly on informal reverse inference (reasoning backward from observed activations to inferred mental processes) from published studies of other tasks that activated the same regions.
+Researchers widely regard this approach as problematic [@poldrack2011].
 Here we take a more direct, psychometric approach to measuring individual differences in perceptual-motor coordination and cognition.
-Psychometric instruments (e.g., tasks, questionnaires) are designed to assess variation in cognitive traits and states, such as fluid intelligence, working memory, attention, motivation, and personality, that have been of theoretical interest to cognitive archaeologists [e.g., @wynn2016].
-It is thus surprising that they have been almost entirely neglected in experimental studies of knapping skill.
+
+Studies implement psychometric instruments (e.g., tasks, questionnaires) to assess variation in cognitive traits and states, such as fluid intelligence, working memory, attention, motivation, and personality.
+These measures have been of theoretical interest to the broader field of cognitive archaeology [e.g., @wynn2016].
+It is thus surprising that experimental studies have almost entirely neglected them.
 In the only published example we are aware of, Pargeter et al.
@@
 However, we are unaware of any such studies of biometric influences on variation in knapping success.
-Finally, the time and effort demands of knapping skill acquisition suggest that differences in personality (e.g., self-control and "grit" [@pargeter2019], motivation [@stout2002], and social vs.
+Finally, individual, and contextual differences in adopting social vs.
 individual learning strategies [@miu2020] might also affect learning outcomes.
 We are again unaware of any previous studies that have assessed such effects.
-In this study, we assessed all participants with a battery of tests including grip strength, movement speed/accuracy, spatial working memory, fluid intelligence, self-control, tendency to use social information, and motivation/engagement with the tool making task.
-We were particularly interested in the possibility that these variables might not only impact learning generally, but might also have different effects under different learning conditions.
+This study evaluated all participants with a battery of tests, including grip strength, movement speed/accuracy, spatial working memory, fluid intelligence, and tendency to use social information.
+We were particularly interested in the possibility that these variables might not only impact learning generally but might also have different effects under different learning conditions.
 
@@
 [-@ohnuma1997] were the first to implement this suggestion in a study of Levallois flake production, followed by more recent studies of handaxe making [@putt2014; @putt2017] and simple flake production [@morgan2015; @cataldo2018; @lombao2017].
-This reflects recent interest in the hypothesis that language might be an adaptation for teaching [e.g., @laland2017; @stout2012].
-Teaching and learning demands of Paleolithic tool making would thus provide evidence of selective contexts favoring language evolution [@stout2010; @morgan2015; @montagu1976].
+This study design trajectory reflects a recent interest in the hypothesis that language might be an adaptation for teaching [e.g., @laland2017; @stout2012].
+Teaching and learning demands of Paleolithic tool-making would thus provide evidence of selective contexts favoring language evolution [@stout2010; @morgan2015; @montagu1976].
 
-Toth and Schick [-@toth1993] were, however, careful to point out that extinct hominid learning strategies and capacities might differ from modern experimental participants.
+However, Toth and Schick [-@toth1993] pointed out that extinct hominid learning strategies and capacities might differ from modern experimental participants.
 Even leaving aside potential species differences in social learning [cf.
@@
 Simply instructing such participants not to speak during an experiment (or to demonstrate but not gesture, etc.
-[@morgan2015]) is likely to underestimate the efficacy of non-verbal teaching and learning in cultural contexts where it is more common, let alone in a hypothetical pre-linguistic hominid species.
+[@morgan2015]) is likely to underestimate the efficacy of non-verbal teaching and learning in cultural contexts where it is more common, let alone in a hypothetical pre-linguistic hominid species [cf.
+@fuentes2015].
 
-Such concerns are exacerbated in experiments using pre-recorded instructional videos or extremely short training periods.
-Video does not allow the interactive teaching that is favored even in formal academic knapping classes [e.g., @shea2015] and is almost certainly typical of traditional learning contexts [e.g., @stout2002].
-It is not known how video presentation affects the efficacy of teaching generally, or the relative effectiveness of different forms of instruction.
-Going further, some experiments have manipulated the presence/absence of verbal instruction by presenting the same video with and without sound [@putt2017] or the sound track without the video [@cataldo2018].
-While this provides experimental control, it does not allow the instructor to adjust their multi-modal [@levinson2014] communication strategies as they would naturally do, for example through pointing and pantomime.
-To simply remove a communication channel without allowing any such adaptation is highly artificial and risks generating results that cannot be generalized beyond the specific context of the experiment [@yarkoni2020].
+Such concerns are exacerbated in experiments using pre-recorded instructional videos or brief training periods.
+Video does not allow the interactive teaching favored even in formal academic knapping classes [e.g., @shea2015] and is almost undoubtedly typical of traditional learning contexts [e.g., @stout2002].
+Researchers do not know how video presentation generally affects the efficacy of teaching stone knapping or the relative effectiveness of different forms of instruction.
+Moreover, some experiments have manipulated the presence/absence of verbal instruction by presenting the same video with and without sound [@putt2017] or the sound track without the video [@cataldo2018].
+While this provides experimental control, it does not allow the instructor to adjust their multimodal [@levinson2014] communication strategies as they would naturally do, for example, through pointing and pantomime.
+Simply removing a communication channel without allowing such adaptation is highly artificial and risks generating results that cannot be generalized beyond the specific context of the experiment [@yarkoni2020].
 Similarly, unnaturally short training periods (e.g., 5-15 minutes [@morgan2015; @lombao2017]) might misrepresent the relative efficacy of different teaching strategies under more realistic conditions [@whiten2015; @stout2015].
-Even the longest training times to date [@stout2015; @pargeter2019] have not produced knapping skills comparable to relevant archaeological examples, and were achieved by limiting sample size and using only one teaching condition.
+Even the most extended training times [@stout2015; @pargeter2019] have not produced knapping skills comparable to relevant archaeological examples and were achieved by limiting sample size and using only one teaching condition.
 
 For these reasons, we sought to explore a middle path between experimental expedience and realism by limiting our experiment to two relatively naturalistic learning conditions and a moderate learning period of two hours.
-As in previous experiments [@stout2011; @hecht2015a; @pargeter2019] the first condition was unrestricted, interactive instruction in small groups, essentially reproducing the "natural" teaching/learning context familiar [cf.
+As in previous experiments [@stout2011; @hecht2015a; @pargeter2019], the first condition was unrestricted, interactive instruction in small groups, essentially reproducing the "natural" teaching/learning context familiar [cf.
 @shea2015] to our WEIRD instructor and student participants.
-The second condition allowed observation only, with the experimenter visible making flakes but not interacting in any way with learners.
+The second condition allowed observation only, with the experimenter visible making flakes but not providing instruction or interacting with learners in any way during this activity.
 This absence of teaching is again a familiar social context for our participants and did not require any novel behaviors from the instructor.
-It matches the "imitation/emulation" condition of Morgan et al.
-[-@morgan2015] although we make no assumptions regarding learning mechanisms.
-We did not include a "reverse engineering" or "end-state emulation" condition in which only finished products were visible.
-This has been advocated as an important baseline or control condition [@whiten2015] to distinguish observational from individual learning, but is not likely to model any typical Paleolithic learning context nor to stand as an adequate proxy for the cognition of hominid species with different social learning capacities.
+Although we make no assumptions regarding learning mechanisms, it matches the "imitation/emulation" condition of Morgan et al.
+[-@morgan2015].
+We did not include a "reverse engineering" or "end-state emulation" condition where only finished products were visible.
+Researchers have advocated these states as an important baseline or control condition [@whiten2015] to distinguish observational from individual learning.
+However, they are not likely to model any typical Paleolithic learning context nor to stand as an adequate proxy for the cognition of hominid species with different social learning capacities.
 There is no reason to assume neurocognitive and behavioral processes of reverse-engineering problem solving in modern humans [e.g., @allen2020] approximate the social learning processes of hominids with more ape-like action observation/imitation capacities [@hecht2013a; @hecht2013b; @stout2019].
@@
 However, they observed the greatest, fastest, and most individually variable skill increases during the first 20 hours of practice.
-In addition, initial performance was moderately correlated with later achievement.
-This suggests that studying early-stage learning may be a pragmatic alternative, especially for research investigating individual differences in aptitude.
-Studies of simple flake production similarly document large initial variation [@stout2015] and rapid early progress [@stout2015; @putt2019; @stoutKnappingSkillEarliest2006].
-We designed the current study to test the utility of studying learning and variation during the first two hours of simple flaking instruction/practice, in hopes of finding a viable compromise between experimental realism and cost.
+In addition, initial performance moderately correlated with later achievement.
+This result suggests that studying early-stage learning may be a practical alternative, especially for research investigating individual differences in aptitude.
+Studies of simple flake production similarly document considerable initial variation [@stout2015] and rapid early progress [@stout2015; @putt2019; @stoutKnappingSkillEarliest2006].
+We designed the current study to test the utility of studying learning and variation during the first two hours of simple flaking instruction/practice in hopes of finding a viable compromise between experimental realism and cost.
 
@@
 Unfortunately, it can be difficult and/or expensive to procure authentic raw materials.
-Experimental studies of knapping skill have often used proxy materials such as flint [@morgan2015; @nonaka2010; @cataldo2018], limestone [@stoutKnappingSkillEarliest2006], porcelain [@khreisheh2013], or heat-treated chert [@putt2014; @putt2017; @putt2019]to model Oldowan and early Acheulean technologies executed in other materials.
+Experimental studies of knapping skill have often used proxy materials such as flint [@morgan2015; @nonaka2010; @cataldo2018], limestone [@stoutKnappingSkillEarliest2006], porcelain [@khreisheh2013], or heat-treated chert [@putt2014; @putt2017; @putt2019] to model Oldowan and early Acheulean technologies executed in other materials.
 As well as being more readily available, these proxies are generally easier to knap.
-This has the benefit of reducing required practice time, but it is unclear how it might affect learning demands more generally or the efficacy of different learning conditions/strategies specifically.
+This study design has the benefit of reducing required practice time.
+Still, it is unclear how it might affect learning demands more generally or the efficacy of different learning conditions/strategies precisely.
 
-To address this, some studies have attempted to more closely match experimental and archaeological raw material types [@stout2011; @duke2015; @pargeter2019].
+Some designs have attempted to closely match experimental and archaeological raw material types [@stout2011; @duke2015; @pargeter2019].
 However, raw materials vary across individual clasts within as well as between types.
-This has led to interest in standardizing experimental core morphology [@nonaka2010] and composition, even if this means using artificial materials such as porcelain [@khreisheh2013], brick [@geribàs2010; @lombao2017], or foam blocks [@schillinger2014].
+This choice has led to an interest in standardizing experimental core morphology [@nonaka2010] and composition, even if this means using artificial materials such as porcelain [@khreisheh2013], brick [@geribàs2010; @lombao2017], or foam blocks [@schillinger2014].
 Such manipulations enhance experimental control and internal validity [@eren2016] at the expense of external generalizability to actual archaeological conditions.
@@
 For example, Pargeter et al.
-[-@pargeter2020] found that predicting specific flaking outcomes on actual handaxe preforms was both more difficult and less technologically important than expected from previous work with standardized, frustum-shaped cores [@nonaka2010].
+[-@pargeter2020] found that predicting specific flaking outcomes on actual handaxe preforms was both more complex and less technologically important than expected from previous work with standardized, frustum-shaped cores [@nonaka2010].
 The alternative to control is to incorporate raw material size, shape, and composition as experimental variables [e.g., @stout2019].
-This allows consideration of raw material selection and response to variation as aspects of skill but correspondingly increases the sample sizes required to identify patterning.
-In considering these issues, we again chose to explore a middle path between pragmatism and realism by employing commercially purchased basalt similar to that known from East African Oldowan sites, allowing clast size and shape to vary within set limits, and selecting the particular clasts provided to each participant to approximate the same distribution.
+This choice allows consideration of raw material selection and response to variation as aspects of skill but correspondingly increases the sample sizes required to identify patterning.
+In considering these issues, we again chose to explore a middle path between pragmatism and realism by employing commercially purchased basalt like that known from East African Oldowan sites.
+We allowed clast size and shape to vary within set limits and selected the particular clasts provided to each participant to approximate the same distribution.
 
@@
 
-This research was approved by the Emory Institutional Review Board (IRB00113024).
+The Emory Institutional Review Board approved this research (IRB00113024).
 All participants provided written informed consent and completed a video release form (<https://databrary.org/support/irb/release-template.html>).
@@
 Twenty-four adult participants with no prior stone knapping experience were recruited from the Emory community using paper fliers and e-mail listserv advertisements.
-We were unable to replace one participant who failed to attend their scheduled session, resulting in a total sample of 23.
-Eleven participants (6 female, 5 male) completed the Untaught condition and 12 (8 female, 4 male) completed the Taught condition.
+Eleven participants (6 female, five male) completed the uninstructed condition, and 12 (8 female, four male) completed the instructed condition.
+We could not replace one participant who failed to attend their scheduled session, resulting in a total sample of 23.
 
@@
 
-Participants were asked to visit the Paleolithic Technology Lab at Emory University to complete one three-hour session.
-Participants were scheduled to attend in six groups of four, however one of these groups had only three participants due to a no-show on the day of the experiment.
+We asked participants to visit the Paleolithic Technology Lab at Emory University to complete one three-hour session.
+Participants attended in six groups of four, however, one of these groups had only three due to a no-show on the day of the experiment.
 Each visit began with the collection of individual differences measures, which took approximately one hour.
-After that, participants undertook 105 minutes (two hours minus a 15-minute break after 1 hour) of stone tool making practice.
-This session was video-recorded, and all lithic products were collected.
-After the tool making task, participants completed an "exit questionnaire" comprising the Intrinsic Motivation Inventory (see below).
+After that, participants undertook 105 minutes (two hours minus a 15-minute break after 1 hour) of stone tool-making practice.
+This session was video-recorded, and we collected all lithic products.
+After the tool-making task, participants completed an "exit questionnaire" comprising the Intrinsic Motivation Inventory (see below).
 
-Participants were compensated for their time with a \$30 gift card.
+We compensated participants for their time with a \$30 gift card.
 They also had the opportunity to earn a performance bonus of \$5, \$10, \$15 or \$20 on the gift card.
 They were told that this bonus would depend on "how well they did" on the last core of their practice session.
-The actual performance measure was not specified, but in order to allow on the spot payment a simple measure of the percentage of starting weight removed from the final core was used such that: \> 30% earned \$5, \> 40% earned \$10, \> 50% earned \$15, \> 75% earned \$20.
+The actual performance measure was not specified, but to allow on the spot payment we used a simple measure of the percentage of starting weight removed from the final core such that: \> 30% earned \$5, \> 40% earned \$10, \> 50% earned \$15, \> 75% earned \$20.
 
@@
 
-1)  Grip strength was measured in kilograms using an electronic hand dynamometer (Camry EH101).
-Strength was measured twice and the higher value recorded.
-Grip strength is a simple measure that is well correlated with overall muscular strength [@wind2010] and a range of other health and fitness measures [@sasaki2007].
-It is hypothesized to be relevant to generating kinetic energy for fracture initiation [@nonaka2010] as well as control and support of the hammerstone [@williams-hatala2018] and core [@faisal2010; @key2015].
+1)  *Grip strength* was measured in kilograms using an electronic hand dynamometer (Camry EH101).
+We measured strength twice and kept the higher value.
+Grip strength is a simple measure well correlated with overall muscular strength [@wind2010] and a range of other health and fitness measures [@sasaki2007].
+Researchers have hypothesized that grip strength is relevant for generating kinetic energy for fracture initiation [@nonaka2010] as well as control and support of the hammerstone [@williams-hatala2018] and core [@faisal2010; @key2015].
 
-2)  Motor accuracy was assessed using a "Fitts Law" reciprocal tapping task.
+2)  *Motor accuracy* was assessed using a "Fitts Law" reciprocal tapping task.
 Fitts Law describes the trade-off between speed and accuracy in human movement, classically measured by tapping back and forth between two targets of varying size and spacing [@fitts1954].
-Archaeologists have proposed [@stout2002; @pargeter2020] that management of this trade-off is critical to the accurate application of appropriate force seen in skilled knapping [@nonaka2010; @roux1995].
+Archaeologists have proposed [@stout2002; @pargeter2020] that managing this trade-off is critical to the accurate application of appropriate force seen in skilled knapping [@nonaka2010; @roux1995].
 We implemented this test on a Surface Pro tablet running free software (FittsStudy Version 4.2.8, default settings) developed by the Accessible Computing Experiences lab (Jacob O.
@@
 
-3)  Visuospatial working memory is the capacity to "hold in mind", which researchers have hypothesized to be important in stone toolmaking performance [@coolidge2005].
-It also might support a learning process known as 'chunking,' in which multiple items or operations are combined into summary chunks stored in long term memory, that is thought to be important in the acquisition of knapping and other skills [@pargeter2019].
+3)  *Visuospatial working memory* is the capacity to "hold in mind", which researchers have hypothesized to be important in stone tool-making performance [@coolidge2005].
+It also might support a learning process known as 'chunking', in which subjects combine multiple items or operations into summary chunks stored in long-term memory, which researchers think is essential in acquiring knapping and other skills [@pargeter2019].
 We measured visuospatial working memory using a free n-back task ([wmp.education.uci.edu/software/](https://wmp.education.uci.edu/software/)) developed by the Working Memory and Plasticity Laboratory at the University of California, Irvine (Susanne Jaeggi, PI) and implemented in E-Prime software on a desktop computer.
-In this task, participants are asked to remember the position of blue squares presented sequentially on the screen and touch a key when the current position matches that 1, 2, 3...n iterations back.
+In this task, instructors tell participants to remember the position of blue squares presented sequentially on the screen and touch a key when the current position matches that 1, 2, 3…n iterations back.
 Progression to blocks with increasing values of n is contingent on exceeding a threshold success rate.
-Performance was measured as the highest n achieved.
+We measured n-back task performance as the highest n achieved.
 
-4)  Fluid intelligence [@cattell1963] refers to the capacity to engage in abstract reasoning and problem solving in a way that is minimally dependent on prior experience.
-It complements "crystallized intelligence" (the ability to apply learned procedures and knowledge) as one of the two factors ($g_f$, $g_c$) comprising so-called "general intelligence" (g).
-Fluid intelligence is closely related to the executive control of attention and manipulation of information held in working memory [@engle2018].
-It is hypothesized to support technological innovation [@coolidge2005] and/or the intentional learning of new skills [@unsworth2005; @stout2015].
+4)  *Fluid intelligence* [@cattell1963] refers to the capacity to engage in abstract reasoning and problem solving in a way that is minimally dependent on prior experience.
+It complements "crystallized intelligence" (the ability to apply learned procedures and knowledge) as one of the two factors ($g_f$, $g_c$) comprising so-called "general intelligence" ($g$).
+Fluid intelligence is closely related to the executive control of attention and manipulation of the information held in working memory [@engle2018].
+Researchers hypothesize that fluid intelligence supports technological innovation [@coolidge2005] and/or the intentional learning of new skills [@unsworth2005; @stout2015].
 We measured fluid intelligence using the short version [@bilker2012] of the classic Raven Progressive Matrices task, which requires participants to complete increasingly difficult pattern matching questions.
 
-5)  The use of social information for learning and decision making varies across individuals and societies [@molleman2019].
+5)  The *use of social information* for learning and decision-making varies across individuals and societies [@molleman2019].
 Such variation is a key topic for understanding social learning and cultural evolutionary processes [@kendal2018; @heyes2018; @miu2020] and represents a potential confound for assessing experimental effects of different social learning conditions.
 We measured participants' tendency to rely on social information vs.
-their own insights using the Berlin Estimate AdjuStment Task (BEAST) developed by Molleman et al.
+their insights using the Berlin Estimate AdjuStment Task (BEAST) developed by Molleman et al.
 [-@molleman2019].
 In this task, participants are present with large arrays of items on a screen and asked to estimate the number present.
-They are then provided with another person's estimate and allowed to provide a second estimate.
-The participants' average adjustment between first and second estimates provides a measure of their propensity to rely on social information.
+They are then provided with another person's estimate and allowed to give a second estimate.
+The participants' average adjustment between the first and second estimates measures their propensity to rely on social information.
 
@@
 After individual difference testing, participants engaged in a 2-hour stone tool making session, with a 15-minute break after 1 hour.
-Participants were instructed not to seek out additional training or information on stone tool making (i.e., via the internet) during these breaks.
-Each group of participants was randomly assigned to one of two experimental conditions: no teaching or teaching.
-In both conditions, participants were first given an opportunity to inspect and handle examples (**Figure** \@ref(fig:fig1)) of the kind of stone tools (flakes) they are being asked to produce.
-They were told that their objective was to produce as many flakes as possible from the materials provided.
-This meant that even the untaught condition included some minimal instruction (being told the objective) , however this was considered to be unavoidable without creating a much more elaborate and naturalistic context in which participants would develop their own technological goals.
-Such a design would also be expected to increase behavioral variability, demanding correspondingly larger samples of participants to identify patterns and making direct comparisons with the taught condition.
+During these breaks, participants were instructed not to seek additional training or information on stone tool-making (i.e., via the internet).
+For each group of participants, we randomly assigned subjects to one of two experimental conditions: no teaching or teaching.
+In both conditions, participants were first allowed to inspect and handle examples made by the expert model/instructor (**Figure** \@ref(fig:fig1)) of the kind of stone tools (flakes) they were asked to produce.
+They were told that their objective was to detach as many flakes as possible from the materials provided.
+These instructions meant that even the uninstructed condition included some minimal instruction (being told the objective).
+However, this was unavoidable without creating a more elaborate and naturalistic context where participants would develop their own technological goals.
+We expected that such a design would also increase behavioral variability, demanding correspondingly larger samples of participants to identify patterns and making direct comparisons with the instructed condition problematic.
 
@@
 
-Each participant was provided with 9 cores for use over the 2-hour experiment.
-These cores were produced from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer.
-This basalt has not been mechanically compared [e.g., rebound hardness, @braun2009] to East African basalts, but appears qualitatively similar to finer grained examples from sites around Lake Turkana (D.
-Braun, pers.
-comm.) and at Gona.
-Spalling produced irregular, angular chunks (**Figure** \@ref(fig:fig2)) for use in the experiment, weighing between 459g - 1876g (mean = 975g).
-All cores were weighed, measured (Length, Width, Thickness), and painted white so that new fracture surfaces could be discriminated from those created during production.
+Each participant was provided with nine nodules for use over the 2-hour experiment.
+We made these nodules from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer.
+This basalt has not been mechanically compared [e.g., rebound hardness, @braun2009] to East African basalts.
+Still, the basalt appears qualitatively like finer-grained examples at Gona (D.
+Stout, pers.
+obs.).
+Spalling produced irregular, angular chunks (**Figure** \@ref(fig:fig2)) for the experiment, weighing between 459g - 1876g (mean = 975g).
+All cores were weighed, measured (Length, Width, Thickness), and painted white so that we could discriminate new fracture surfaces from those created during production (see supplementary Table 3).
 Cores were sorted by shape and weight and then distributed evenly to each participant.
-As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length $\times$ Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p = 0.9) of cores provided.
-This was also true for cores provided to participants across the two experimental conditions (Taught vs.
-Untaught mean weight = 1001g vs.
+As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length $\times$ Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = 0.6, df1=22, df2 = 184, p = 0.9) of cores provided.
+This was also true for cores provided to participants across the two experimental conditions (Instructed vs.
+Uninstructed mean weight = 1001g vs.
 956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4; mean shape = 221.43 vs.
 221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p= 0.05).
-Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores actually used across participants and conditions could still emerge as a result of selection bias.
+Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores used across participants and conditions could still emerge because of selection bias.
 
@@
 
-Sixty pounds of 3-to-5 inch basalt "Mexican Beach Pebbles" were purchased from a landscaping supply company for use as hammerstones in the experiment.
-Of these, 90 were selected as suitable for use.
+Sixty pounds of 3-to-5 inch basalt "Mexican Beach Pebbles" were purchased from a landscaping supply company as hammerstones in the experiment.
+Of these, we selected 90 as suitable for use.
 These weighed between 213g-1360g (mean = 425) and varied in elongation (L/W = 1.01 to 2.65) and relative thickness (L$\times$W/T = 90.48 to 283.67).
-Forty-five stones were placed in the middle of the knapping area (**Figure** \@ref(fig:fig3)) for participants to freely choose from during the experiment.
+We placed roughly two-thirds of the stones in the middle of the knapping area (**Figure** \@ref(fig:fig3)) for participants to freely choose from during the experiment.
 Broken hammerstones were replaced from the reserve to maintain a consistent number and range of choices.
-Each hammerstone was numbered and participants' choices were recorded along with the number of the core(s) being worked on with a particular hammerstone.
+We numbered each hammerstone and recorded each participants' choices along with the number of the core(s) they worked on with a particular hammerstone.
 
@@
 In both conditions, three researchers were present to record activities and collect materials.
-Participants were seated in a circle [@shea2015] and experiments were video recorded using two cameras.
-Participants were free to select hammerstones from the common pile and to work on any or all of their nine assigned cores in any order they preferred (**Figure** \@ref(fig:fig3)).
+Participants were seated in a circle [@shea2015], and experiments were video recorded using two cameras.
+Participants were free to select hammerstones from the common pile and work on any of their nine assigned cores in any order they preferred (**Figure** \@ref(fig:fig3)).
 However, each core and all associated debitage were collected before participants were allowed to start working on a new core, so it was not possible to partially work and then return to a particular core later.
@@
 
-In the untaught condition, a researcher (DS) sat with the participants and made stone tools but remained silent and made no effort to facilitate learning (e.g., through gesture, modified performance, facial expression, attention direction, or verbal instruction).
-Over the 2-hour period, the researcher completely reduced four cores (one every \~30 minutes).
-Participants were not restricted from talking to each other, as this would create an unnatural and potentially stressful social context that might affect learning.
-Participants were asked to avoid any form of communication about the tool making task specifically, and they complied with this request.
-Participants in this condition thus had the opportunity to observe tool making (**Figure** \@ref(fig:fig4)) by an expert and/or by other learners, should they choose to do so, but received no intentional instruction.
+In the uninstructed condition, a researcher (DS) sat with the participants and made stone tools but remained silent and made no effort to facilitate learning (e.g., through gesture, modified performance, facial expression, attention direction, or verbal instruction).
+Over the 2 hours, the researcher completely reduced four cores (one every \~30 minutes).
+We did not restrict participants from talking to each other, creating an unnatural and potentially stressful social context that might affect learning.
+Participants were asked to avoid any form of communication about the tool-making task specifically, and they complied with this request.
+Participants in this condition thus had the opportunity to observe tool-making (**Figure** \@ref(fig:fig4)) by an expert and/or by other learners, should they choose to do so, but received no intentional instruction.
 
@@
 
-In the Taught condition, there were no restrictions on participant interaction and the researcher engaged in direct active teaching [@kline2015] of tool-making techniques through verbal instruction, demonstration, gesture, and shaping of behavior.
-The instructor has a moderate level of experience teaching basic knapping skills to students in undergraduate archaeology classes and to participants in previous knapping research [e.g., @stout2011].
-The pedagogical strategy employed was based on the instructor's own learning experiences and theoretical interpretations [e.g., @pargeter2020], and focused on coaching participants in effective body postures, movement patterns, and grips as well as the assessment of viable core morphology.
+In the instructed condition, there were no restrictions on participant interaction, and the researcher engaged in direct active teaching [@kline2015] of tool-making techniques through verbal instruction, demonstration, gesture, and shaping of behavior.
+The instructor has a moderate level of experience teaching basic knapping skills to students in undergraduate archaeology classes and participants in previous knapping research [e.g., @stout2011].
+The pedagogical strategy employed was based on the instructor's own learning experiences and theoretical interpretations [e.g., @pargeter2020] and focused on coaching participants in effective body postures, movement patterns, and grips and the assessment of viable core morphology.
 
@@
 **Table** \@ref(tab:tab1) lists details for the study’s various lithic attributes.
-All finished cores were weighed and measured (L, W, T).
-Delta weight was calculated as (Start weight-End weight)/Start weight.
+We weighed and measured all the finished cores (L, W, T).
+Delta weight was calculated as (nodule start weight-core end weight)/nodule start weight.
 All detached pieces (DPs) were collected and weighed.
-We did not sort DPs into types (e.g., whole flakes, fragments) as this would have greatly increased processing time and it is not clear that such distinctions add relevant information regarding utility/desirability beyond that supplied by metrics [@stout2019].
+We did not sort DPs into types (e.g., whole flakes, fragments), which would have greatly increased processing time.
+It is not clear that such distinctions add relevant information regarding utility/desirability beyond that supplied by metrics [@stout2019].
 All DPs larger than 40mm in maximum dimension were photographed and measured.
@@
 We selected a higher threshold for both pragmatic (analysis time) and theoretical reasons.
-Flake use experiments have shown that flakes weighing less than 5--10 g or with a surface area below 7--10 cm2 [@prasciunas2007] or with a maximum dimension \<50-60 mm [@key2014] become markedly inefficient for basic cutting tasks.
-Similarly, data from Oldowan replication experiments [@stout2019] show that the utility index (flake cutting edge/flake mass1/3) \* (1 - exp[-0.31 \* (flake maximum dimension -- 1.81)]) developed by Morgan et al.
-(2015) falls off rapidly below 40mm maximum dimension ( Mean Utility \< 40mm = 0.508; \>=40mm = 0.946; t= 11.99, df = 707, p \< 0.000).
-By including weight in our cut-off criteria we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.
+Flake use experiments have shown that flakes weighing less than  5–10 g or with a surface area below 7–10 $cm^2$ [@prasciunas2007] or with a maximum dimension \<50-60 mm [@key2014] become markedly inefficient for basic cutting tasks.
+Similarly, data from Oldowan replication experiments [@stout2019] show that the utility index (flake cutting edge/flake mass$^{1/3}$) \* (1 - exp[-0.31 \* (flake maximum dimension - 1.81)]) developed by Morgan et al.
+(2015) falls off rapidly below 40mm maximum dimension (Mean Utility \< 40mm = 0.508; \>=40mm = 0.946; t= 11.99, df = 707, p \< 0.000).
+By including weight in our cut-off criteria, we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.
+We recorded the absolute number of above-threshold DPs produced per core and calculated a mass-based proportion of larger DPs by dividing the combined weight of DPs \>40mm in maximum dimension and 5g in weight by the weight of all DPs.
+Higher values show cores from which proportionally more of the detached mass comprises large pieces.
 ```{r tab1, echo=FALSE, message=FALSE}
@@
 
-For measurement, DP length was defined as the longest axis and width as the maximum dimension orthogonal to length.
-Thickness was defined as the maximum dimension orthogonal to the plane formed by L and W and was measured using calipers.
+We defined DP length as the longest axis and width as the maximum dimension orthogonal to length for measurement (see supplementary Table 2).
+Our thickness measurement follows the maximum dimension orthogonal to the plane formed by L and W, which we measured using calipers.
 L, W, and plan-view area measurements were taken from photographs captured using a Canon Rebel T3i fitted with a 60 mm macro lens and attached to a photographic stand with adjustable upper and lower light fittings.
 The camera was positioned directly above the flakes and kept at a constant height.
-DPs were positioned irrespective of any technological features so that the longest axis was vertical, and the wider end was placed toward the bottom of the photograph.
+We arranged DPs irrespective of any technological features so that the longest axis was vertical and placed the broader end toward the bottom of the photograph.
 
-Photographs were post-processed using Equalight software to adjust for lens and lighting falloff that result from bending light through a lens and its aperture which can affect measurements taken from photographs.
-Each image was shot with a scale that was then used to rectify the photograph's pixel scale to a real-world measurement scale in Adobe Photoshop.
-Images were converted to binary black and white format and silhouettes of the tools were extracted in Adobe Photoshop.
-We then used a custom ImageJ [@rueden2017] script [@pargeter2019] to measure DP length and take nine width measurements at 10% increments of length starting at the base of each DP.
+* Photographs were post-processed using Equalight software to adjust for lens and lighting falloff resulting from bending light through a lens and its aperture, affecting measurements taken from photos.
+We shot each image with a scale then used this scale to rectify the photograph's pixel scale to a real-world measurement scale in Adobe Photoshop.
+Images were converted to binary black and white format, and silhouettes of the tools were extracted in Adobe Photoshop.
+We then used a custom ImageJ [@rueden2017] script [@pargeter2019]to measure DP length and take nine width measurements at 10% increments of length starting at the tip of each DP (i.e., 10% width = tip, 90% width = base).
 We used the built-in ImageJ tool to measure DP area.
-A "Proportion Larger DPs" was calculated per core as the combined weight of all DPs \>40mm in maximum dimension and 5g in weight divided by the weight of all DPs.
-Higher values show cores with proportionally more large DPs.
 
@@
 
-To evaluate the association between psychometric, motor-skill, and training measures and technological outcomes, we adopted an information-theoretic approach [@burnham2002].
-Information-theoretic approaches provide methods for model selection using all possible combinations of variables while avoiding problems associated with significance-threshold stepwise selection.
-We used the corrected Akaike information criterion (AICc) to rate each possible combination of predictors on the balance between goodness of fit (likelihood of the data given the model) and parsimony (number of parameters).
-The AICc consists of the log likelihood (i.e., how well does the model fit the data?) and a penalty term for the number of parameters that must be estimated in the model (i.e., how parsimonious is the model?), with a correction for small sample sizes (AICc converges to the standard AIC at large samples).
-A lower AICc indicates a more generalizable model and we used it to compare and rank various possible models.
-Each analysis begins with a full model that includes all predictors of interest.
-All possible combinations of predictors are then fit, and the resulting models are ranked and weighted based on their AICc.
-The "best" model is chosen because it has the lowest AICc score.
+We adopted an information-theoretic approach to evaluate the association between psychometric, motor-skill, training measures, and technological outcomes [@burnham2002].
+Information-theoretic techniques provide methods for model selection using all possible combinations of variables while avoiding problems associated with significance-threshold stepwise selection.
+We used the small sample corrected Akaike information criterion (AICc) to rate each possible combination of predictors on the balance between the goodness of fit (likelihood of the data given the model) and parsimony (number of parameters).
+The AICc consists of the log-likelihood (i.e., how well does the model fit the data?) and a penalty term for the number of parameters that must be estimated in the model (i.e., how parsimonious is the model?), with a correction for small sample sizes (AICc converges to the standard AIC at large samples).
+A lower AICc indicates a more generalizable model, and we use it to compare and rank various possible models.
+Each analysis begins with a complete model that includes all predictors of interest.
+All possible combinations of predictors are then fitted, and the resulting models are ranked and weighted based on their AICc.
+We chose the "best" model because it had the lowest AICc score.
 
 Continuous predictors were centered such that zero represents the sample average, and units are standard deviations.
-The full model was fitted with the lm function in R 3.2.3, and the glmulti package was used for multi-modal selection and model comparison.
+The full model was fitted with the lm function in R 3.2.3, and the *glmulti* package (Bartoń 2015) was used for multi-modal selection and model comparison.
 
@@
 Here we limit discussion to the major ﬁndings regarding flaking performance and individual differences.
-We were particularly interested in: 1) group level effects of experimental condition, 2) individual differences in aptitude and learning, and 3) potential interactions between learning conditions and individual differences.
+We were particularly interested in 1) group-level effects of experimental conditions, 2) individual differences in aptitude and learning, and 3) potential interactions between learning conditions and individual differences.
 To address these questions, we employed data reduction (Principal Component Analysis) to derive two summary metrics of flaking performance, compared these factors across the two experimental conditions, and built multivariate models examining the relations between our various psychometric measures, subject's motor skill scores, and our two lithic performance factors.
@@
 
-The following two sections outline factor analyses designed to summarize our main study metrics tracking individual variation in DP sizes and shapes and lithic performance measures.
+The following two sections outline factor analyses that summarize our main study metrics tracking individual variation in DP sizes and shapes and lithic performance measures.
 
@@
 
-To better understand the relationship between DP shape and training/individual variation, we entered our nine flake linear plan measurements along with maximum flake length and thickness into a principal component analysis (PCA) from which summary coordinates were extracted.
-Bartlett's Test of Sphericity was significant ($\chi ^2$ (10) =4480, p \< .01) indicating that the set of variables are adequately related for factor analysis.
+To better understand the relationship between DP shape and training/individual variation, we entered our nine linear plan measurements and maximum length and thickness into a principal component analysis (PCA) from which we extracted summary coordinates.
+Bartlett's Test of Sphericity was significant ($\chi ^2(10)$  = 4480, p \< .01), indicating that the set of variables are adequately related for factor analysis.
 
@@
 
+Our exploratory study produced findings with both theoretical and methodological implications.
+Perhaps the clearest result is that the effects of different experimental learning conditions and their interaction with individual cognitive and motor differences are complicated and must be interpreted with care.
+For example, we did not find robust effects of learning condition on mean group performance but did find that teaching reduced variation knapping rate, methods, and outcomes, thus providing a “safety net” against failed transmission.
+This increase in reliability of skill reproduction provides further support for the hypothetical co-evolution of teaching, language, and tool making (Stout 2010; Morgan et al.
+2015).
+At the same time, we found that two hours practice was insufficient to document learning-related performance change.
+This implies that longer experiment durations may be needed to study learning, even for basic Oldowan-like flaking, and that care should be taken interpreting results from shorter studies.
+It is thus important to have a quantitative standard of proficiency, ideally archaeological (Pargeter et al.
+2019; Stout et al.
+2019) but at least experimental (this study), against which to compare performance.
+Here, it was clear that participants remained in an early learning stage far from expert performance, and that teaching had important effects on the way that learners explored the novel task space.
+Teaching allowed learners to maximize both quality and quantity rather than trading off between them, and dramatically changed the effects of individual differences in strength, visuospatial working memory, and social learning tendencies on knapping outcomes.
+These results highlight the value of including individual difference measures in knapping experiments and suggest that the presence/absence of teaching in the Paleolithic would have fundamentally altered associated selective pressures (Stout et al.
+2019) in addition to affecting the reliability of skill reproduction.
+To unravel these complicated and nuanced relationships, future experiments might benefit from longer learning periods, video-based quantitative analyses of knapping behaviors (Geribas et al.
+2010; Stout et al.
+2021), more refined psychometric and lithic measures, a greater diversity of social and cultural contexts (Barrett 2020), and explicit attention to teacher skill, style, and interpersonal relationships with participants (Bevilacqua et al.
+2019).
 # Acknowledgments

### 2021-08-21 code/PaST_r_code.R

@@ -1051,8 +1051,7 @@ autoplot(pred.mm_grip) +
 
 
 ## Supplementary tables
-## 1.lithic performance (total cores used, 
-## total flake mass/flaked core mass; 
+## 1. Lithic performance comparison between two groups
 
 complete<-complete_past_data_psycho
 aggregate(complete$mass_flakes_flaked_mass_bycore, by=list(complete$Condition), FUN=function(x) cbind(mean(x), median(x), sd(x), min(x),max(x)))
@@ -1072,7 +1071,7 @@ wilcox.test(Total_Cores_Used ~ Condition, data = individual,
 plyr::count(individual, 'Condition')
 
 
-## 2. 
+## 2. Flake metrics comparison between two groups
 complete1<-complete_past_data_plusflakes %>% 
   mutate(value = replace(max_flake_length, 
                          max_flake_length == "n/a", NA)) %>% 
@@ -1097,7 +1096,7 @@ wilcox.test(flake_max_thickness ~ Condition, data = complete1,
 plyr::count(complete1, 'Condition')
 
 
-## 3.
+## 3. Starting nodule metrics comparison between two groups
 aggregate(experiment_data_filled$start_mass, by=list(experiment_data_filled$Condition), FUN=function(x) cbind(mean(x), median(x), sd(x), min(x),max(x)))
 aggregate(experiment_data_filled$maxlength, by=list(experiment_data_filled$Condition), FUN=function(x) cbind(mean(x), median(x), sd(x), min(x),max(x)))
 aggregate(experiment_data_filled$maxwidth, by=list(experiment_data_filled$Condition), FUN=function(x) cbind(mean(x), median(x), sd(x), min(x),max(x)))

### 2021-08-21 manuscript/manuscript.Rmd
updated the main text and supplementary tables.

@@
 ```{r fig1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects examining demonstration flakes prior to the experiment.
-The demonstration flakes were made from the same basalt as used in the experiment with the same knapping technique."}
+The demonstration lakes were made from the same basalt as used in the experiment with the same knapping technique."}
 knitr::include_graphics(here::here("figure/Figure1.jpg"))
@@
 Spalling produced irregular, angular chunks (**Figure** \@ref(fig:fig2)) for the experiment, weighing between 459g - 1876g (mean = 975g).
-All cores were weighed, measured (Length, Width, Thickness), and painted white so that we could discriminate new fracture surfaces from those created during production (see supplementary Table 3).
+All cores were weighed, measured (Length, Width, Thickness), and painted white so that we could discriminate new fracture surfaces from those created during production (see supplementary Table 1).
 Cores were sorted by shape and weight and then distributed evenly to each participant.
@@
 
-```{r fig4, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects observing expert knapper in the untaught condition."}
+```{r fig4, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects observing expert knapper in the uninstructed condition."}
 knitr::include_graphics(here::here("figure/Figure4.jpg"))
@@
 Continuous predictors were centered such that zero represents the sample average, and units are standard deviations.
-The full model was fitted with the lm function in R 3.2.3, and the *glmulti* package (Bartoń 2015) was used for multi-modal selection and model comparison.
+The full model was fitted with the *lm* function in R 4.0.3, and the *glmulti* package (Bartoń 2015) was used for multi-modal selection and model comparison.
 
@@
 The analysis yielded three factors explaining a total of 90% of the variance for the entire 11 measurement variable set (**Table** \@ref(tab:tab2)).
-Factor 1 tracks flake size with higher scores indicating larger flakes since all 11 measures load positively on this factor.
-Factor 2's loadings track the increasing relationship between thickness, length, and flake width.
-As factor 2 scores increase, flakes get thicker, longer, and narrow, resembling irregular splinters.
-Factor 3 tracks the relationship between flake proximal and distal width relative to thickness.
-As factor 3 scores go up, flakes get thinner and narrower at the distal ends and wider at the base.
-Factor 3 therefore tracks flakes with a typical shape having a thin cross-section, wider base, and narrower tip.
-We used these three flake shape coordinates to approximate DP size and shape in the project's flake performance factor analysis.
+Factor 1 tracks size with higher scores indicating larger pieces since all 11 measures positively load this factor.
+Factor 2's loadings track increasing thickness and length but not width.
+As factor 2 scores increase, pieces thus get thicker, longer, and relatively narrower, resembling irregular splinters.
+Factor 3 tracks increasing proximal breadth and decreasing thickness.
+As factor 3 scores go up, pieces get thinner overall and broader at the base.
+Factor 3, therefore, tracks pieces with a more typical flat, flake-like morphology having a thin cross-section and tapering shape.
+We used these three shape coordinates to approximate DP size and shape in the project's flaking performance factor analysis.
 ```{r tab2, echo=FALSE, message=FALSE}
@@
   kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
-kableExtra::row_spec(0, bold = T)
+kableExtra::row_spec(0, bold = T)%>%
+kableExtra::row_spec(1, bold = T)
 ```
@@
 
-To better understand the relationship between our various lithic performance measurements and to reduce data dimensionality, we conducted a second principal component analysis examining the study's six lithic performance measures (count of large pieces [\>40mm and 5g], mass of large pieces relative to total detached mass, core delta mass, and the three flake shape factors).
-All of these measures were summarized for each core and unique factor scores were calculated from these core-specific measures.
-Bartlett's Test of Sphericity was significant ($\chi ^2$ (6) =3185, p \< .01) indicating that the set of variables are at least adequately related for factor analysis.
+We conducted a second principal component analysis better to understand the relationship between our various lithic performance measurements and reduce data dimensionality.
+This analysis examined the study's six lithic performance measures (count of large pieces [\>40mm and 5g], mass of large pieces relative to total detached mass, core delta mass, and the three flake shape factors) (see supplementary Table 3 for details).
+These measures were summarized for each core, and individual factor scores were calculated from these core-specific measures.
+Bartlett's Test of Sphericity was significant ($\chi ^2(6)$  =3185, p \< .01) indicating that the set of variables are at least adequately related for factor analysis.
 
-The analysis yielded two factors explaining a total of 56% of the variance for the entire set of variables (**Table** \@ref(tab:tab3)).
-Factor 1 (hereafter "Quantity") explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and on core delta mass.
+The analysis yielded two factors explaining 56% of the variance for the entire set of variables (**Table** \@ref(tab:tab3)).
+Factor 1 (hereafter "Quantity") explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and core delta mass.
 Performance factor 2 (hereafter "Quality") covers 27% of the sample variance and measures flaking quality as reflected in high positive loadings on Shape Factors 1 (size) and 3 (thin, "flake-like" shape) and a negative loading on Shape Factor 2 ("splinter-like" thickness and elongation).
-High scores on Quality thus reflect production of larger, relatively thinner, and more typically flake-shaped vs.
+High scores on Quality thus reflect the making of larger, relatively thinner, and more typically flake-shaped vs.
 splinter-shaped DPs.
@@
   kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
-kableExtra::row_spec(0, bold = T)
+kableExtra::row_spec(0, bold = T)%>%
+kableExtra::row_spec(1, bold = T)
 ```
 
-These two factors address flaking performance at the level of individual cores, however we were also interested in the overall productivity/rate of work of each participant over the entire two hours.
-For example, looking at a knappers average Quality and Quantity factor scores would not differentiate between a participant who spent the entire time exhaustively reducing one core vs.
+These two factors address flaking performance at the level of individual cores.
+However, we were also interested in each participant's overall productivity/work rate over the entire two hours.
+For example, looking at a knappers average Quality and Quantity factor scores would not differentiate between a participant who spent the whole time exhaustively reducing one core vs.
 another participant who did the same to all nine of their allotted cores in the same time.
-To capture this aspect of variation we calculated a simple Total Productivity metric as the sum of all mass a participant removed from cores during the experiment.
+We calculated a simple Total Productivity metric as the sum of all mass a participant removed from cores during the experiment to capture this aspect of variation.
 
@@
 
-This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (**Figure** \@ref(fig:fig5) and **Figure** \@ref(fig:fig6)).
-As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (**Figure** \@ref(fig:fig5)a), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj.
-$R^2$ = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj.
+This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (**Figures** \@ref(fig:fig5) and \@ref(fig:fig6)).
+As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (**Figure** \@ref(fig:fig5)a), although this relationship is twice as strong in the instructed  (F[1, 9] = 33, p \< 0.01, Adj.
+$R^2$ = 0.8) compared to uninstructed  (F[1, 8] = 8, p = 0.02, Adj.
 $R^2$ = 0.4) group.
-Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj.
-$R^2$ = 0.7), but no relation in the Trained group (**Figure** \@ref(fig:fig5)b).
-A qualitatively similar trend with respect to Quality vs.
+Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the uninstructed group (F[1, 8] = 28, p = \<0.01, Adj.
+$R^2$ = 0.7), but no relation in the instructed group (**Figure** \@ref(fig:fig5)b).
+A qualitatively similar trend concerning Quality vs.
 Quantity (**Figure** \@ref(fig:fig5)c) did not achieve significance (F[1, 17] = 0.6, p = 0.2).
@@
 
-Thus, it appears that Trained participants achieved higher Total Productivity by increasing average flaking Quantity across cores and without sacrificing Quality, whereas Untrained participants found other ways to vary Total Productivity (e.g., number of cores knapped rather than Quantity per core, see variance Table and Figure) and generally increased productivity at the expense of Quality.
-Experimental artifacts illustrating these trade-offs are presented in **Figure???**.
+Thus, it appears that instructed participants achieved higher Total Productivity by increasing average flaking Quantity across cores and without sacrificing Quality.
+In contrast, uninstructed participants found other ways to vary Total Productivity (e.g., number of cores knapped rather than Quantity per core, see **Table** \@ref(tab:tab5), **Figure** \@ref(fig:fig8) & \@ref(fig:fig9)) and generally increased productivity at the expense of Quality.
+Experimental artifacts illustrating these trade-offs are presented in **Figure** \@ref(fig:fig6).
 
@@
 
-Here we compare our flaking outcomes (DP size/shape and flaking performance factors) between the trained and untrained groups.
-Our expert demonstrator/instructor is included as a performance benchmark.
+Here we compare our flaking outcomes (DP size/shape and flaking performance factors) between the instructed and uninstructed groups.
+We include the expert demonstrator/instructor as a performance benchmark.
 
-**Table** \@ref(tab:tab4) summarizes the results of ANOVA tests group level difference in central tendency on various performance measures.
-We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors.
-In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (**Figure** \@ref(fig:fig7)).
+**Table** \@ref(tab:tab4) summarizes the results of ANOVA tests group-level difference in central tendency on various performance measures.
+We found no significant differences between the instructed and uninstructed groups on our flaking Quantity and Quality factors.
+In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14).
+The expert also had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26) and left significantly smaller finished cores (effect size = 0.27) (**Figure** \@ref(fig:fig7)).
 All three of these results show either medium or large effect sizes.
-In all three comparisons, the trained group's data distributions tended towards the expert sample although they were not significantly different from the untrained group (**Figure** \@ref(fig:fig8)).
-We also observed a significant difference in shape factor 2 (splinters) driven by the expert's lower values, but with a very low (\<0.01) effect size.
-These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.
+The instructed group's data distributions tended towards the expert sample in all three comparisons, although they did not differ significantly from the uninstructed group (**Figure** \@ref(fig:fig7)).
+We also observed a significant difference in shape factor 2 (splinters) driven by the expert's lower values but with a very small (\< 0.01) effect size.
+These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance.
+In contrast, novices in experimental groups produced similar mean detached piece sizes and shapes as the expert trainer.
 ```{r tab4, echo=FALSE, message=FALSE}
@@
 
-```{r fig8, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Comparison of untaught, taught, and expert cores."}
+```{r fig8, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Comparison of uninstructed, instructed, and expert cores."}
 knitr::include_graphics(here::here("figure/Figure8.jpg"))
@@
 
-While we did not find significant differences in central tendency between our two experimental groups, results (**Figure** \@ref(fig:fig7)) did indicate lower variance in the trained group.
-To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the trained and untrained individuals using the F-test on either core-averaged or flake specific variances.
-**Table** \@ref(tab:tab5) and **Figure** \@ref(fig:fig9) present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass).
-In most instances, variance in the untrained group exceeds that of trained individuals by 1.5 to 4.7 times.
-The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the untrained group (**Figure** \@ref(fig:fig9)), rather than to shift the mean.
+While we did not find significant differences in central tendency between our two experimental groups, results (**Figure** \@ref(fig:fig7)) did indicate lower variance in the instructed group.
+To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the instructed and uninstructed individuals using the F-test on either core-averaged or flake-specific variances.
+**Table** \@ref(tab:tab5) and **Figure** \@ref(fig:fig9) present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and the total amount of flaked mass).
+In most instances, variance in the uninstructed group exceeds that of instructed individuals by 1.5 to 4.7 times.
+The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the uninstructed group (**Figure** \@ref(fig:fig9)), rather than to shift the mean.
 ```{r tab5, echo=FALSE, message=FALSE}
@@
 
-```{r fig9, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Variance comparisons between trained and untrained individuals across various flaking performance metrics.
+```{r fig9, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Variance comparisons between instructed and uninstructed individuals across various flaking performance metrics.
 Note the lower variance of the trained group in all the plots."}
@@
 
-In addition to comparing overall performance during the two hour experiment, we also wanted to determine if groups or individuals differed in learning (i.e., performance change) over the period.
-For these analyses, we calculated the learning stage as the ordinal number of each core out of the total number knapped by each subject (i.e., core 2 of 4 or 4 of 8 both equal 50% complete).
+In addition to comparing overall performance during the two-hour experiment, we also wanted to determine if groups or individuals differed in learning (i.e., performance change) over the period.
+We calculated the learning stage as the ordinal number of each core out of the total number knapped by each subject (i.e., core 2 of 4 or 4 of 8, both equal 50% complete) for these analyses.
 These relative core use-order percentages were then binned into 20 percent brackets for core-order and group-level comparisons.
-Flaking outcomes were tracked using the two performance factors (Quality and Quantity).
+We tracked flaking outcomes using the two performance factors (Quality and Quantity).
 We added the nodule starting mass to track whether training/practice times impacted raw material selection.
@@
 This result demonstrated that flaking outcomes did not change dramatically across the study interval.
-This lack of significant learning effects is confirmed by an inspection of individual learning curves (Figure).
-The one significant main training effect related to core starting mass (with a strong effect size = 0.25).
+An inspection of individual learning curves confirms this lack of significant learning effects (**Figure** \@ref(fig:fig11)).
+The main training effect was core starting mass (with a substantial effect size = 0.25).
 On average, core starting masses start low and increase, showing that participants selected smaller nodules first.
-As the smaller nodules in their allotment were depleted, participants were left to knap larger, less preferred nodules.
-This preference for smaller cores is somewhat less pronounced in the untrained group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the untrained group (Figure???).
+As participants depleted the smaller nodules in their allotment, they knapped larger, less preferred nodules.
+This preference for smaller cores is somewhat less pronounced in the uninstructed group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the uninstructed group (**Figure** \@ref(fig:fig11)).
 ```{r tab6, echo=FALSE, message=FALSE}
@@
 
-One of the experiment's primary goals was to test if measures of individual perceptual-motor and cognitive variation predict success in stone flaking across different training conditions.
-To address this goal, we built three multivariate models examining the relations between training conditions, individual difference measures, and our three lithic performance measures (overall productivity and average per-core Quantity and Quality).
-These models enabled us to determine which of the psychometric and motor skill factors are better predictors of a participant's flaking performance in the study.
+One of the experiment's primary goals was to test if individual perceptual-motor and cognitive variation measures predict success in stone flaking across different training conditions.
+We built three multivariate models to address this goal examining the relations between training conditions, individual difference measures, and our three lithic performance measures (overall productivity and average per-core Quantity and Quality).
+These models enabled us to determine which psychometric and motor skill factors are better predictors of a participant's flaking performance in the study.
 
@@
 Each model's continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitts score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations.
-Our two motor skill and strength measures (grip strength and Fitt's performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, $R^2$ = 0.41).
+Our two motor skill and strength measures (grip strength and Fitt’s performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, $R^2$ = 0.41).
 However, these two measures track complementary components of athleticism (strength vs.
-speed/accuracy tradeoffs) and so we decided to include both in the model selection process.
+speed/accuracy trade-offs), so we decided to include both in the model selection process.
 
-The full models were fitted with the lm function in R 3.2.3, and we used the Glmulti package's automated model selection algorithm to select the best performing model (lowest AICc score) (see methods for further details on the multimodal selection process).
 All three models follow the same complete model statement as follows:
@@
 
-For our two per-core performance factors (Quantity and Quality) it is also relevant to consider how individual core features may have affected performance.
+For our two per-core performance factors (Quantity and Quality), it is also relevant to consider how individual core features may have affected performance.
 We found no evidence of individual or group level practice effects over the two hours, so we did not include core order in the models.
 We did, however, find that subjects selected progressively larger nodules throughout the experiment.
-It is thus important to understand whether nodule variability had any impact on our flaking results.
-Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p \< 0.01, $R^2$ = 0.54) we included nodule mass as a covariate to control for any variance in flaking performance that may be driven by nodule differences.
+It is thus essential to understand whether nodule variability had any impact on our flaking results.
+Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p \< 0.01, $R^2$ = 0.54), we included nodule mass as a covariate to control for any variance in flaking performance that nodule differences may drive.
 
@@
 Our first model examined variance in overall flaking productivity measured by each subject's combined flaked mass (nodule starting mass-core final mass).
-This provides a basic measure of variation in individuals' success detaching pieces and reducing cores from a standardized (see Methods) raw material supply.
+This productivity metric provides a basic measure of variation in individuals' success detaching pieces and reducing cores from a standardized (see Methods) raw material supply.
 From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of -18 (Average AIC = -13).
@@
 
-**Table** \@ref(tab:tab7) presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average.
-The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable.
+**Table** \@ref(tab:tab7) presents this model's coefficients and summary outputs, wherein baseline refers to the uninstructed condition with all continuous predictors at the sample average.
+The parameter estimates for the continuous predictors reflect the expected change in utility for one standard deviation change in the predictor variable.
 We found significant (p\< 0.05) and substantial (Standardized Estimate \>= 0.50, i.e.
 a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST.
-The main effect of Grip Strength (**Figure** \@ref(fig:fig12)), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.
+The main effect of Grip Strength (**Figure** \@ref(fig:fig12)), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers, at least when we do not consider efficiency and quality.
 ```{r fig12, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Significant main effect of grip strength on overall flaking productivity."}
@@
 Effects of visuospatial working memory capacity and social information use are more complicated, as indicated by strong interactions with learning condition (**Figure** \@ref(fig:fig13)).
-In each case, higher scores were associated with better performance in the uninstructed group but worse performance in the instructed group.
-Positive effects in the uninstructed group were as expected, given the hypothesized importance of spatial cognition (Coolidge and Wynn 2005) and social learning (Morgan et al.
-2015) in the acquisition of knapping skills.
-Negative effects in the trained group are unexpected but presumably reflect differences in learning strategies adopted under instruction.
+In each case, higher scores were associated with better performance in the uninstructed group but *worse* performance in the instructed group.
+Positive effects in the uninstructed group were as expected, given the hypothesized importance of spatial cognition [@coolidge2005] and social learning [@morgan2015] in acquiring knapping skills.
+Negative effects in the instructed group are unexpected but presumably reflect differences in learning strategies adopted under instruction.
 
@@
 
-**Table** \@ref(tab:tab8) presents this model’s coefficients and summary outputs, following the same format as Table. 
+**Table** \@ref(tab:tab8) presents this model’s coefficients and summary outputs, following the same format as **Table** \@ref(tab:tab6) and **Table** \@ref(tab:tab7).
 ```{r tab8, echo=FALSE, message=FALSE}
@@
 The Quantity model roughly paralleled results for Total Production, yielding substantial and significant interactions between training condition, n-back level, BEAST scores, and grip strength.
-As with Total Production, higher visuospatial n-back levels and BEAST scores were associated with lower Quantity scores in the trained group but higher or unchanged Quantity in the untrained group (**Figure** \@ref(fig:fig14)).
+As with Total Production, higher visuospatial n-back levels and BEAST scores were associated with lower Quantity scores in the instructed group but higher or unchanged DP Quantity in the uninstructed group (**Figure** \@ref(fig:fig14)).
 
@@
 
-Unlike Total Productivity, the effect of Grip Strength on per-core Quantity was mediated by an interaction with learning condition (**Figure** \@ref(fig:fig15)).
-Thus, high Grip Strength enabled individuals in both groups to produce more total debitage, but only Instructed individuals translated Grip Strength into more intense reduction of individual cores, including not only delta mass, but also number and proportion of larger pieces.
+Unlike Total Productivity, an interaction with learning condition mediated the effect of Grip Strength on per-core Quantity (**Figure** \@ref(fig:fig15)).
+Only instructed individuals translated Grip Strength into a more intense reduction of individual cores, including delta mass and the number and proportion of larger pieces.
+Thus, high Grip Strength enabled individuals in both groups to produce more debitage.
 
@@
 The Quality model did produce two statistically (p\< 0.05) significant interaction effects (RPM \* BEAST & Fitts \* n-back).
-However, these interactions had relatively small effects on Quality (\<0.5) and we believe that interpreting these results from our small, exploratory study would be inappropriate.
+However, these interactions had relatively small effects on Quality (\<0.5), and we believe that interpreting these results from our small, exploratory study would be inappropriate.
 **Figure** \@ref(fig:fig16) shows the uneven distribution of data points for these interactions, which suggests vulnerability to leveraging effects of a small number of extreme value combinations.
@@
 
-We designed this exploratory study primarily to trial experimental design elements such as training time, conditions, and raw materials and to collect preliminary data on the effect of individual differences and training on knapping outcomes.
+We designed this exploratory study primarily to trial experimental design elements such as training time, conditions, and raw materials and to collect preliminary data on individual differences and training effects on knapping outcomes.
 We thus focused on collecting quantitative psychometric and lithic data.
-However, we also considered that quantifying participant knapping behaviors as well as products could be important for future studies.
+However, we also considered that quantifying participant knapping behaviors and products could be important for future studies.
 To support methods development in this regard, we made ad hoc notes on observed behaviors during the experiments and video-recorded all experiments to enable later, more systematic analyses yet to be completed.
 However, even casual behavior observation was sufficient to reveal an unexpected effect.
-Whereas all trained participants copied the general posture and technique of the expert (free hand knapping seated in a chair) fully half (6) of the uninstructed participants experimented with or even knapped all of their cores using the floor as a support (**Figure** \@ref(fig:fig17)).
-Three of these participants were in the same session, which is also the only group composed of just three individuals.
+Whereas all instructed participants copied the general posture and technique of the expert (freehand knapping seated in a chair), fully half (6) of the uninstructed participants experimented with or even knapped all their cores using the floor as support (**Figure** \@ref(fig:fig17)).
+Three of these participants were in the same session, the only group composed of just three individuals.
 In this group, knapping on the ground appears to have been transmitted from one participant to the other two based on appearance order and the point of gaze of participants.
 
-```{r fig17, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Novices in the untaught condition developing and transmitting a form of bipolar technology involving core reduction on the floor."}
+```{r fig17, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Novices in the uninstructed condition developing and transmitting a form of bipolar technology involving core reduction on the floor."}
 knitr::include_graphics(here::here("figure/Figure17.jpg"))
@@
 
-The most salient finding of our exploratory study is that the presence/absence of teaching clearly impacted knapping performance but did so in nuanced and individually variable ways that have not been explored in previous studies.
-In fact, we did not observe any significant differences in mean performance between our experimental conditions.
-Some non-significant tendencies toward enhanced Instructed group performance suggest that a larger participant sample might detect significant effects, but also that the size of any such effects would likely remain small.
-This could reasonably lead to the conclusion that teaching does not substantially facilitate early stage knapping skill acquisition [cf.
-@putt2014; @ohnuma1997].
-Looking closer, however, we found a number of important teaching effects.
+The most salient finding of our exploratory study is that the presence/absence of teaching clearly impacted knapping performance but did so in nuanced and individually variable ways that previous research has not explored.
+We did not observe any significant differences in mean performance between our experimental conditions.
+Some non-significant tendencies toward enhanced instructed group performance suggest that a larger participant sample might detect significant effects, but also that the size of any such effects would likely remain small.
+This result could reasonably lead one to conclude that teaching does not substantially facilitate early-stage knapping skill acquisition [@putt2014; @ohnuma1997].
+Looking closer, however, it is evident that teaching has several significant effects.
 
@@
 In our experiment, the strongest effects of teaching were to reduce variance (**Figure** \@ref(fig:fig9), **Table** \@ref(tab:tab5)) rather than shift mean values.
-In particular, teaching acted as a "safety net" that homogenized performance by reducing the frequency of extremely poor outcomes (i.e., learning failures).
-This finding provides additional support for the hypothesis that teaching would have increased the reliability of Oldowan skill reproduction [@morgan2015] while simultaneously corroborating the view that basic flaking competence can be achieved in its absence [@tennie2017].
-Our results thus do not imply that teaching was required or even present during Oldowan times (but see @gärdenfors2017), but rather serve to reinforce the plausibility of co-evolutionary scenarios positing the cost/reliability of technological skill acquisition as a selection pressure favoring the evolution of teaching and language [@stout2010; @morgan2015; @stout2017].
+Instruction acted as a "safety net" that homogenized performance by reducing the frequency of inferior outcomes (i.e., learning failures).
+This finding provides additional support for the hypothesis that teaching would have increased the reliability of Oldowan skill reproduction [@morgan2015] while simultaneously corroborating the view that naïve knappers can achieve basic flaking competence in its absence [@tennie2017].
+Our results thus do not imply that novices required teaching or that teaching was even present during Oldowan times (@gärdenfors2017).
+Instead, the results serve to reinforce the plausibility of co-evolutionary scenarios positing the cost/reliability of technological skill acquisition as a selection pressure favoring the evolution of teaching and language [@stout2010; @morgan2015; @stout2017].
 
@@
 The current study complements the transmission chain design of Morgan et al.
-[-@morgan2015] by finding similar effects in more naturalistic learning contexts.
-Our design further allowed us to examine individual variation to better understand how teaching produces its effects.
-Whereas transmission chains are optimized to investigate iterative learning effects [but see @caldwell2020], they necessarily involve a different instructor/model for each participant.
-We sacrificed this iterative component in order to consider how the presence/absence of teaching affected the behavior of individuals under otherwise standardized learning conditions.
+[-@morgan2015] (2015) by finding similar effects in more naturalistic learning contexts.
+Our design further allowed us to examine individual variation to understand better how teaching produces its effects.
+Whereas transmission chains are optimized to investigate iterative learning effects [see @caldwell2020], they necessarily involve each participant's different instructor/model.
+We sacrificed this iterative component to consider how the presence/absence of teaching affected the behavior of individuals under otherwise standardized learning conditions.
 
-We found that a key impact of teaching was to alter basic flake production strategies, as reflected in the relationship between Total Productivity and detached piece Quality (Fig.Xb).
-Whereas Untrained participants achieved greater Productivity at the expense of Quality and vice versa, these dimensions were unrelated across Trained participants.
-Thus, even though Untrained participants achieved the highest values on each metric, only trained individuals managed to maximize both simultaneously.
-Indeed, a core function of teaching is to reduce the search space that learners must explore and increase the likelihood of discovering globally as opposed to locally optimum solutions [cf.
+We found that a critical impact of teaching was to alter basic flake production strategies, as reflected in the relationship between Total Productivity and detached piece Quality (**Figure** \@ref(fig:fig5)b).
+Whereas uninstructed participants achieved greater Productivity at the expense of Quality and vice versa, these dimensions were unrelated across instructed participants.
+Thus, even though uninstructed participants gained the highest values on each metric, only instructed individuals managed to maximize both simultaneously.
+Indeed, a core function of teaching is to reduce the search space that learners must explore and increase the likelihood of discovering globally instead of locally optimum solutions [cf.
 @hinton1996; @stout2013].
-In our study, Untrained individuals explored a greater range of basic behavioral variations not seen in the Trained group, including knapping on the floor, concentrating on working just a few cores (2-4, Figure) over the practice period, and showing less constrained nodule size preferences (Figure).
-It is notable that this variation occurred even in the presence of an observable expert example, suggesting it may be interesting for future experiments to address the impact of social context, expectations, and relationships on observational learning strategies [@kendal2018].
+In our study, uninstructed individuals explored a greater range of basic behavioral variations not seen in the instructed group, including knapping on the floor, concentrating on working just a few cores (2-4, **Figure** \@ref(fig:fig9)c) over the practice period.
+Uninstructed individuals also showed less constrained nodule size preferences (**Figure** \@ref(fig:fig8)).
+Notably, this variation occurred even in the presence of an observable expert example, suggesting it may be interesting for future experiments to address the impact of social context, expectations, and relationships on observational learning strategies [@kendal2018].
 
-We also found a strong positive effect of Grip Strength on Total Productivity independent of learning condition (Figure).
-While it is tempting to interpret this with respect to the demands for hand strength specifically, it is important to remember that grip strength is strongly correlated with total muscle strength [@wind2010] and overall fitness [@sasaki2007].
-Thus, it is best taken to indicate some importance of fitness generally in increasing the rate and intensity of core reduction by naïve knappers, potentially affecting rate of work and the kinetic energy of the swing as well as the handling of core and hammerstone.
-It thus provides further support for hypotheses positing stone tool making as a selection pressure on the functional anatomy of hand, arm, and shoulder [e.g., @williams-hatala2018], but initially appears orthogonal to variations in learning condition and knapping behaviors in our study.
-However, we also found that the effect of Grip Strength on per-core knapping Quantity is dependent on teaching (Figure).
-The absence of this effect in the uninstructed group reflects the weaker association between Total Productivity and per-core Quality across these participants (Fig.Xa) and shows Grip Strength increased uninstructed Productivity specifically by allowing them to knap more cores rather than to reduce individual cores more heavily.
-In keeping with this, uninstructed Grip Strength is positively correlated with Total Cores knapped ($R^2$ = 0.54, p = 0.01).
+We also found a strong positive effect of Grip Strength on Total Productivity independent of instruction (**Figure** \@ref(fig:fig12)).
+While it is tempting to interpret this with respect to the demands for hand strength specifically, it is important to remember that grip strength correlates with total muscle strength [@wind2010] and overall fitness [@sasaki2007].
+Thus, it is best taken to indicate some importance of fitness generally in increasing the rate and intensity of core reduction by naïve knappers, potentially affecting the rate of work and the kinetic energy of the swing and the handling of core and hammerstone.
+It thus provides further support for hypotheses positing stone tool-making as a selection pressure on the functional anatomy of hand, arm, and shoulder [e.g., @williams-hatala2018], but initially appears orthogonal to variations in learning condition and knapping behaviors in our study.
+However, we also found that the effect of Grip Strength on per-core knapping Quantity is dependent on teaching (**Figure** \@ref(fig:fig15)).
+The absence of this effect in the uninstructed group reflects the weaker association between Total Productivity and per-core Quality across these participants (**Figure** \@ref(fig:fig5)a).
+Grip Strength increased uninstructed Productivity specifically by allowing them to knap more cores rather than to reduce individual cores more heavily.
+In keeping with this, uninstructed Grip Strength positively correlates with Total Cores knapped  ($R^2$ = 0.54, p = 0.01).
 Conversely, strength allowed instructed participants to increase their average Quantity per core without affecting the total number of cores knapped ($R^2$ = 0.18, p = 0.165).
-Thus, strength appears to have achieved its effects on core reduction rate and intensity in different ways, depending on teaching.
-This difference is likely related to the homogenizing effect of teaching on knapping rate (all instructed participants knapped 6 or more cores) and methods.
-Subjectively, knapping behaviors of uninstructed participants often appeared more physically demanding (e.g., greater number of non-productive blows, rapid and unregulated battering) which would imply different demands on both strength and aerobic fitness [@williams-hatala2021; @mateos2019].
-However, this remains to be systematically investigated.
+Thus, strength appears to have affected core reduction rate and intensity in different ways, depending on teaching.
+This difference is likely related to the homogenizing effect of teaching on knapping rate (all instructed participants knapped six or more cores) and methods.
+Subjectively, uninstructed participants’ knapping behaviors often appeared more physically demanding (e.g., a greater number of non-productive blows, rapid and unregulated battering).
+This result would imply different demands on strength and aerobic fitness [@williams-hatala2021; @mateos2019].
+However, future research will need to test this observation systematically.
 
 In this respect, it is also important to note that we do not know how well the knapping objectives and strategies communicated by the expert in our experiment correspond to actual Oldowan goals and behaviors.
-The instructor has successfully replicated assemblage-level patterning at Gona [@stout2019] but Oldowan behavior is variable across space and time [e.g., @braun2019] and alternative knapping methods might maximize different values (productivity, quality, effort), especially in novices [@putt2015] (Putt 2015).
+The instructor has successfully replicated assemblage-level patterning at Gona [@stout2019].
+Still, Oldowan behavior is variable across space and time [e.g., @braun2019], and alternative knapping methods might maximize different values (productivity, quality, effort), especially in novices [@putt2015].
 Nevertheless, the effect of instruction to constrain behavioral exploration and homogenize outcomes is clear.
-We expect that this effect would generalize to the teaching of alternative knapping goals and behaviors, although this remains to be tested.
+We expect this effect to generalize to teaching alternative knapping goals and behaviors, although future work should test this prediction.
 
@@
 
-One major goal of this experiment was to test the viability of a moderate, two-hour, learning period for studies of skill acquisition.
+One primary goal of this experiment was to test the viability of a moderate, two-hour learning period for skill acquisition studies.
 Unfortunately, we found that this duration was insufficient to capture learning effects for Oldowan-like flake production.
-The lack of performance change over the period (Figure) cannot be attributed to a ceiling effect (i.e., rapid task mastery at the outset of the practice period) as participants remained well below expert levels and continued to display the high within-individual variability typical of naïve/novice knapping [@eren2011; @pargeter2019].
+We cannot attribute the lack of performance change over the period (**Figure** \@ref(fig:fig11)) to a ceiling effect (i.e., rapid task mastery at the outset of the practice period).
+Participants remained well below expert levels and continued to display the high within-individual variability typical of naïve/novice knapping [@eren2011; @pargeter2019].
 This negative result was unexpected but is broadly consistent with evidence that Early Stone Age flaking, while conceptually simple, requires substantial practice for perceptual-motor skill development [@nonaka2010; @stout2015; @pargeter2020].
-Future investigations of learning variation across individuals and/or experimental conditions may thus need to incorporate longer practice periods to capture skill acquisition processes.
-In theory, much shorter knapping trials might be used to assess the variation in initial performance across individuals and under different conditions that is captured in our study.
+Future investigations of learning variation across individuals and/or experimental conditions may thus need to incorporate more extended practice periods to capture skill acquisition processes.
+In theory, future experiments might use shorter knapping trials to assess the variation in initial performance across individuals and under different conditions captured in our study.
 However, the presence of substantial core-to-core variation within individuals cautions against overly brief experiments that might not provide a representative sample.
@@
 exploitation of knowledge and skills already in hand [@sutton2018].
-Premature exploitation risks settling for a sub-optimal local solution whereas continued exploration sacrifices more immediate payoffs.
-Managing this trade-off is especially challenging for complex, real-world tasks like stone knapping, and is thought to depend on the interplay of uncertainty and reward expectation [@wilson2021].
-Teaching and social learning generally have the potential to provide low-cost information about task structure and payoffs [@rendell2010; @kendal2018], which if adopted, would be expected to affect exploration/exploitation decisions.
-Such adoption is itself known to be influenced by individual cognitive differences, for example if higher fluid intelligence allows observers to better understand observed tasks [@vostroknutov2018] or if individuals vary in their tendency to use and value social information [@toelch2014; @molleman2019].
+Premature exploitation risks settling for a sub-optimal local solution, whereas continued exploration sacrifices more immediate payoffs.
+Managing this trade-off is especially challenging for complex, real-world tasks like stone knapping and is thought to depend on the interplay of uncertainty and reward expectation [@wilson2021].
+Teaching and other social learning resources have the potential to provide low-cost information about task structure and payoffs [@rendell2010; @kendal2018], which, if adopted, would be expected to affect exploration/exploitation decisions.
+Such adoption is itself known to be influenced by individual cognitive differences, for example, if higher fluid intelligence allows observers to better understand observed tasks [@vostroknutov2018] or if individuals vary in their tendency to use and value social information [@toelch2014; @molleman2019].
 
-In our study, we did not observe any effect of fluid intelligence (RPM) on knapping outcomes but did find strong interactions of learning condition with participant visuospatial working memory and social information use tendency (Figure).
+Our study did not observe any effect of fluid intelligence (RPM) on knapping outcomes.
+Still, we found strong learning condition interactions with participant visuospatial working memory and social information use tendency (**Figure** \@ref(fig:fig13)).
 As expected, uninstructed individuals with higher scores on these dimensions displayed higher Total Productivity and average per-core flaking Quantity (although the effect on n-Back on Quantity did not achieve significance).
-We attribute these effects to increased ability to hold relevant morphological/spatial information in mind and a tendency to benefit from observing successful strategies of others, including the expert model.
+We attribute these effects to an increased ability to hold relevant morphological/spatial information in mind and a tendency to benefit from observing successful strategies of others, including the expert model.
 In contrast, instructed individuals with higher scores tended to have lower Productivity and Quantity.
-We interpret this unexpected effect to an increased tendency to privilege exploratory learning behavior over exploitation.
-In particular, we suggest that trained participants might knap more slowly and less productively if higher working memory capacity inclined them to experiment more with morphological/spatial variables highlighted by the instructor or if a predisposition to use social information use caused them to invest greater time and effort attending to and trying out observed actions and/or instructions.
+We interpret this unexpected effect as an increased tendency to privilege exploratory learning behavior overexploitation.
+We suggest that instructed participants might knap more slowly and less productively if higher working memory capacity inclined them to experiment more with morphological/spatial variables highlighted by the instructor.
+Instructed subjects’ predisposition to use social information might have caused them to invest greater time and effort attending to and trying out observed actions and/or instructions.
 These suggestions remain to be tested by further work.
 
-Unfortunately, the training period in our current experiment was insufficient to capture learning effects and so we have no evidence of the effects of these individual differences and putative exploration/exploitation tradeoffs to the ultimate achievement of expertise.
-A similar negative effect of instruction on knapping outcomes during early stage learning was reported by Putt et al.
-[-@putt2014], and has been interpreted to reflect learners experimenting with advanced techniques before they have the perceptual-motor skill to execute them [@whiten2015; @stout2015].
-Such effects might be further explored with more detailed behavioral data, as opposed to purely lithic data, and with longer learning periods.
+Putt et al.
+[-@putt2014] report a similar negative effect of instruction on knapping outcomes during early-stage learning.
+They interpret this result to reflect learners experimenting with advanced techniques before they have the perceptual-motor skill to execute them [@whiten2015; @stout2015].
+Unfortunately, the training period in our current experiment was insufficient to capture learning effects, so we have no evidence of the impact of these individual differences and putative exploration/exploitation trade-offs to the ultimate achievement of expertise.
+Future research might explore such effects with more detailed behavioral data instead of purely lithic data and with more extended learning periods.
 
@@
 
-Although our exploratory study produced a number of robust results with respect to the effects of instruction and individual differences on lithic products, it is clearly limited by a small sample size, short training duration, and lack of detailed quantification of observed behaviors.
-These are limitations that can hopefully be addressed in future studies building on the methods and evidence presented here.
+Our exploratory study produced several robust results concerning the effects of instruction and individual differences on lithic products.
+However, we had a small sample size, short training duration, and a lack of detailed quantification of observed behaviors.
+We can hopefully address these limitations in future studies building on the methods and evidence presented here.
 For example, it is notable that our study failed to document any reliable effects on knapping Quality.
-Obviously, this might reflect an actual lack of such effects, but it may also indicate a need for more sensitive measures and/or increased sample size and training duration to identify subtle or delayed effects.
+This lack of a result might reflect an actual lack of such effects.
+Still, it may also indicate a need for more sensitive measures and/or increased sample size and training duration to identify subtle or delayed effects.
 One aspect of our attempt to balance pragmatic costs and benefits in our study was to test the efficacy of relatively limited lithic analysis.
-More detailed ongoing analyses of core morphology and debitage features (e.g., typology, cutting edge length, platform dimensions) may yet reveal a more reliable signal of knapping quality.
-Results of the Quality model in particular also seem to suffer from the uneven distribution and discrete rather than continuous nature of scores on our RPM and n-Back tests.
-Concerns about the sampling of variation on these dimensions could be addressed with larger samples or by pre-screening participants to ensure more even representation.
-Alternative psychometric tests (e.g., full rather than short version of the RPM) might also provide more sensitive and continuous measures.
+More detailed ongoing investigations of core morphology and debitage features (e.g., typology, cutting edge length, platform dimensions) may yet reveal a more reliable signal of knapping quality.
+Results of the Quality model also seem to suffer from the uneven distribution and discrete rather than continuous nature of scores on our RPM and n-Back tests.
+We could address concerns about sampling variation on these dimensions with larger samples or by pre-screening participants to ensure more even representation.
+Alternative psychometric tests (e.g., complete rather than the short version of the RPM) might also provide more sensitive and continuous measures.
 
-Another major limitation that our study shares with all other published experiments on knapping skill acquisition is that we do not address variation in social and cultural context or in teaching style.
+Another major limitation our study shares with all other published experiments on knapping skill acquisition is that we do not address variation in social and cultural context or teaching style [@kline2015] .
 Currently, we have little basis other than personal experience/tradition [@callahan1979; @whittaker1994; @shea2015] and theoretical speculation [@stout2013; @whiten2015] from which to assess which pedagogical techniques are most effective even in WEIRD contexts.
 No study to date has considered how variation in teacher skill [@shea2015] or social relationship to participants might impact learning under different conditions.
-To properly address these questions would require a major research program, including both cross-cultural comparative studies [@barrett2020] and more naturalistic study designs.
+Addressing these questions would require a significant research program, including cross-cultural comparative studies [@barrett2020; @fuentes2016] and more naturalistic study designs.
 While costly, such research would produce results of broad relevance to anthropologists, biologists, psychologists, and sociologists interested in teaching and learning.
@@
 Our exploratory study produced findings with both theoretical and methodological implications.
-Perhaps the clearest result is that the effects of different experimental learning conditions and their interaction with individual cognitive and motor differences are complicated and must be interpreted with care.
-For example, we did not find robust effects of learning condition on mean group performance but did find that teaching reduced variation knapping rate, methods, and outcomes, thus providing a “safety net” against failed transmission.
-This increase in reliability of skill reproduction provides further support for the hypothetical co-evolution of teaching, language, and tool making (Stout 2010; Morgan et al.
-2015).
-At the same time, we found that two hours practice was insufficient to document learning-related performance change.
-This implies that longer experiment durations may be needed to study learning, even for basic Oldowan-like flaking, and that care should be taken interpreting results from shorter studies.
-It is thus important to have a quantitative standard of proficiency, ideally archaeological (Pargeter et al.
-2019; Stout et al.
-2019) but at least experimental (this study), against which to compare performance.
-Here, it was clear that participants remained in an early learning stage far from expert performance, and that teaching had important effects on the way that learners explored the novel task space.
-Teaching allowed learners to maximize both quality and quantity rather than trading off between them, and dramatically changed the effects of individual differences in strength, visuospatial working memory, and social learning tendencies on knapping outcomes.
-These results highlight the value of including individual difference measures in knapping experiments and suggest that the presence/absence of teaching in the Paleolithic would have fundamentally altered associated selective pressures (Stout et al.
-2019) in addition to affecting the reliability of skill reproduction.
-To unravel these complicated and nuanced relationships, future experiments might benefit from longer learning periods, video-based quantitative analyses of knapping behaviors (Geribas et al.
-2010; Stout et al.
-2021), more refined psychometric and lithic measures, a greater diversity of social and cultural contexts (Barrett 2020), and explicit attention to teacher skill, style, and interpersonal relationships with participants (Bevilacqua et al.
-2019).
-# Acknowledgments
+Perhaps the clearest result is that the effects of different experimental learning conditions and their interaction with individual cognitive and motor differences are complicated, and researchers must interpret them with care.
+For example, we did not find robust effects of instruction on mean group performance.
+Still, we did find that teaching reduced variation knapping rate, methods, and outcomes, thus providing a "safety net" against failed transmission.
+This increase in reliability of skill reproduction offers further support for the hypothetical co-evolution of teaching, language, and tool-making [@morgan2015;@stout2010].
+At the same time, we found that two hours of practice was insufficient to document learning-related performance change.
+This result implies that we may need longer experiment durations to study learning, even for basic Oldowan-like flaking.
+Researchers should take care when interpreting results from shorter studies.
+It is thus important to have a quantitative standard of proficiency, ideally archaeological [@pargeter2019;@stout2019] but at least experimental (this study), against which to compare performance.
+Here, it was clear that participants remained in an early learning stage far from expert performance and that teaching had important effects on how learners explored the novel task space.
+Instruction allowed learners to maximize both quality and quantity rather than trading off between them and dramatically changed the effects of individual differences in strength, visuospatial working memory, and social learning tendencies on knapping outcomes.
+These results highlight the value of including individual difference measures in knapping experiments and suggest that the presence/absence of teaching in the Paleolithic would have fundamentally altered associated selective pressures [@fuentes2015; @stout2019] in addition to affecting the reliability of skill reproduction.
+To unravel these complicated and nuanced relationships, future experiments might benefit from more extended instruction periods, video-based quantitative analyses of knapping behaviors [@geribàs2010; @stout2021], more refined psychometric and lithic measures, a greater diversity of social and cultural contexts [@barrett2020], and explicit attention to teacher skill, style, and interpersonal relationships with participants [@bevilacqua2019].
 

### 2021-08-21 code/PaST_r_code.R

@@ -1051,9 +1051,9 @@ autoplot(pred.mm_grip) +
 
 
 ## Supplementary tables
-## 1. Lithic performance comparison between two groups
+## 3. Lithic performance comparison between two groups
 
-complete<-complete_past_data_psycho
+complete<-complete_past_data_shape_skill_pca_combined_bycore
 aggregate(complete$mass_flakes_flaked_mass_bycore, by=list(complete$Condition), FUN=function(x) cbind(mean(x), median(x), sd(x), min(x),max(x)))
 aggregate(complete$total_flaked_mass_bycore, by=list(complete$Condition), FUN=function(x) cbind(mean(x), median(x), sd(x), min(x),max(x)))
 aggregate(complete$total_flakes_above40mm_5g_bycore, by=list(complete$Condition), FUN=function(x) cbind(mean(x), median(x), sd(x), min(x),max(x)))
@@ -1063,6 +1063,7 @@ wilcox.test(total_flaked_mass_bycore ~ Condition, data = complete,
             paired = FALSE)
 wilcox.test(total_flakes_above40mm_5g_bycore ~ Condition, data = complete,
             paired = FALSE)
+
 plyr::count(complete, 'Condition')
 individual<-complete[!duplicated(complete$subject), ]
 aggregate(individual$Total_Cores_Used, by=list(individual$Condition), FUN=function(x) cbind(mean(x), median(x), sd(x), min(x),max(x)))
@@ -1072,7 +1073,7 @@ plyr::count(individual, 'Condition')
 
 
 ## 2. Flake metrics comparison between two groups
-complete1<-complete_past_data_plusflakes %>% 
+complete1<-complete_past_data_shape_skill_pca_combined_byflake %>% 
   mutate(value = replace(max_flake_length, 
                          max_flake_length == "n/a", NA)) %>% 
   drop_na()
@@ -1096,7 +1097,7 @@ wilcox.test(flake_max_thickness ~ Condition, data = complete1,
 plyr::count(complete1, 'Condition')
 
 
-## 3. Starting nodule metrics comparison between two groups
+## 1. Starting nodule metrics comparison between two groups
 aggregate(experiment_data_filled$start_mass, by=list(experiment_data_filled$Condition), FUN=function(x) cbind(mean(x), median(x), sd(x), min(x),max(x)))
[…]

### 2022-05-10 manuscript/manuscript.Rmd
made a few minor changes to figures.

@@
 (2015) falls off rapidly below 40mm maximum dimension (Mean Utility \< 40mm = 0.508; \>=40mm = 0.946; t= 11.99, df = 707, p \< 0.000).
-By including weight in our cut-off criteria, we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.
+By including weight in our cut-off criteria, we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40mm, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.
 We recorded the absolute number of above-threshold DPs produced per core and calculated a mass-based proportion of larger DPs by dividing the combined weight of DPs \>40mm in maximum dimension and 5g in weight by the weight of all DPs.

### 2022-05-10 code/PaST_r_code.R

@@ -438,9 +438,9 @@ var.test(skill_pca_dim2_quality_flaking ~ Condition, complete_past_data_shape_sk
 p9b<-ggplot(filter(complete_past_data_shape_skill_pca_combined_bycore,!subject=="demo"), 
        aes(x=Condition, y=skill_pca_dim2_quality_flaking, color=Condition)) +
   geom_boxplot(outlier.size=0)+
-  geom_point(size=3,position = position_jitterdodge()) +
-  scale_color_manual(labels = c("Untrained", "Trained"), values = c("blue", "red"))+
-  scale_x_discrete(labels = c('Untrained','Trained'))+
+  geom_point(size=3,position = position_jitterdodge(),aes(shape=Condition)) +
+  scale_color_manual(labels = c("observation only", "active teaching"), values = c("blue", "red"))+
+  scale_x_discrete(labels = c('observation only','active teaching'))+
   theme_classic() +
   theme(legend.position = "none")+
   ggtitle("b") +xlab("") +
@@ -463,9 +463,9 @@ var.test(total_flakes_above40mm_5g_bycore ~ Condition, complete_past_data_shape_
 p9a<-ggplot(filter(complete_past_data_shape_skill_pca_combined_bycore,!subject=="demo"), 
        aes(x=Condition, y=total_flakes_above40mm_5g_bycore, color=Condition)) +
   geom_boxplot(outlier.size=0)+
-  geom_point(size=3,position = position_jitterdodge()) +
-  scale_color_manual(labels = c("Untrained", "Trained"), values = c("blue", "red"))+
-  scale_x_discrete(labels = c('Untrained','Trained'))+
+  geom_point(size=3,position = position_jitterdodge(),aes(shape=Condition)) +
+  scale_color_manual(labels = c("observation only", "active teaching"), values = c("blue", "red"))+
+  scale_x_discrete(labels = c('observation only','active teaching'))+
   theme_classic() +
   theme(legend.position = "none")+
   ggtitle("a") +xlab("") +
@@ -491,9 +491,9 @@ var.test(removed_mass_sum ~ Condition, total_mass_data,
 p9c<-ggplot(filter(total_mass_data,!subject=="demo"), 
        aes(x=Condition, y=Total_Cores_Used, color=Condition)) +
   geom_boxplot(outlier.size=0)+
-  geom_point(size=3,position = position_jitterdodge()) +
-  scale_color_manual(labels = c("Untrained", "Trained"), values = c("blue", "red"))+
[…]
